### 2018-06-11 README.Rmd

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# taxald

## Package API

Design sketch/spec for package API:

- Given vector of taxonomic names, return taxonomic identifiers
- Given taxonomic identifiers, return heirarchical classification.

(Note that for some authorities, e.g. GBIF and TPL, taxonomic identifiers are only assigned to species names.  So given the name of a higher-level rank, no id could be returned)

- Given any higher-order rank name (e.g. `infraorder`) return identifiers of all member species.

- Given common names, resolve to accepted scientific name

- Resolve any known miss-spelling, recognized synonymous name, generic name or name-part to corresponding taxonomic name.

- Map between identifiers

- Normalize rank names

---

- Resolve a list of names across multiple authorities to attain higher coverage, 
- provide merged tables

See [schema.md](schema.md) for a sketch of the underlying database architecture imposed on the data source from each authority.

Note that all inputs and outputs should be vectorized.

### 2018-06-29 README.Rmd
some more :package: infrastructure

@@
 ---
+
+[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
+[![Travis build status](https://travis-ci.org/cboettig/taxald.svg?branch=master)](https://travis-ci.org/cboettig/taxald)
+[![Coverage status](https://codecov.io/gh/cboettig/taxald/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxald?branch=master)
+[![CRAN status](https://www.r-pkg.org/badges/version/taxald)](https://cran.r-project.org/package=taxald)
 
@@
 # taxald
+
+**This package is purley exploratory at this stage!**
+
+Central to `taxald` design is to create a *local* database of all avialable taxonomic authority data that we can query quickly, even when working with tens of thousands of species.
+The second key feature of `taxald` design is to return queries from each authority using a *consistent* and convenient table structure, facilitating queries that compare, combine, or work across names from multiple authorities.
+
+## Install and initial setup
+
+`taxald` should now be installable.  To get started, try:
+
+```{r eval=FALSE}
+devtools::install_github("cboettig/taxald")
+```
+
+Before we can use most `taxald` functions, we need to do a one-time installation of the database `taxald` uses for almost all commands.
+This can take a while to run, but needs only be done once.
+The database is installed on your local harddisk and will persist between R sessions.
+Download and install the database
+
+```{r}
+library(taxald)
+
+```
 
@@
 
----
-
 - Resolve a list of names across multiple authorities to attain higher coverage, 

### 2018-06-29 README.Rmd
some function documentation

@@
 The database is installed on your local harddisk and will persist between R sessions.
-Download and install the database
+Download and install the database like so:
 
@@
 
+create_taxadb()
 ```
 
+By default, this will install the database into a hidden `.taxald` folder in your home directory (`~/.taxald`).
+You can control this by passing the path to `create_taxadb()` or setting the environmental variable `TAXALD_HOME` to a different location.
+
+This function downloads each of the individual tables and loads them into a [MonetDBLite](https://www.monetdb.org) database.
+(MonetDBLite is much like SQLite, requiring no external server setup.
+However, for our purposes, MonetDB's columnar design provides significantly faster performance than SQLite, Postgres, or even some in-memory operations in `dplyr`.)
+
+Currently tables are pulled from the development cache on GitHub.
+Many of these tables are still being cleaned up and standardized, see [schema.md](schema.md).
+Multiple tablesare provided for each authority, though some of these contain redundant information, different orientations may prove more convenient or computationally efficient.
+
 ## Package API
+
+Once the database is installed, we can start to make some queries.
+
+First, let's get a nice big species list, say, all the birds (known to the Catalogue of Life):
+
+```{r}
+df <- descendents(name = "Aves", rank = "class", authority="col")
+```
 

### 2018-06-29 R/taxald.R

@@ -12,13 +12,42 @@
 
 ## Consider memoizing
 
+
+#' Return the full classification hierarchy for requested species or ids
+#' 
+#' @param species a character vector of species names, typically 
+#'  specified as (`Genus species` or `Genus species epithet`)
+#' @param id alternately users can provide a vector of species ids.
+#'  IDs must be prefixed matching the requested authority.  See `id`
+#'  column returned by most `taxald` functions for examples.
+#' @param authority from which authority should the hierachy be returned?
+#'  Default is 'itis'.  
+#' @param collect logical, default `TRUE`. Should we return an in-memory
+#' data.frame (default, usually the most convenient), or a reference to
+#' lazy-eval table on disk (useful for very large tables on which we may
+#' first perform subsequent filtering operations.) 
+#' @return a data.frame with one row for each requested species, 
+#'  giving the species id, and column for each of the unique-rank
+#'  levels of the species.  Note that the different authorities recognize
+#'  a range of different ranks.
+#'  
+#' @details Some authorities recognize multiple values of the same rank
+#' level, (i.e. a species may be assigned to two different "suborders").  
+#' In this case, only one has been included in the "hierarchy" schema.
+#' Likewise, some authorities (i.e. NCBI) use some unnamed rank levels 
+#' (i.e. a scientific name is associated with a clade, but the clade is 
+#' not associated with any traditionally recognized rank). In these cases, 
+#' see the "long" schema for more complete classification.  
+#' @export
 #' @importFrom dplyr right_join tibble collect
 #' 
 hierarchy <- function(species = NULL, 
                       id = NULL, 
                       authority = c("itis", "ncbi", "col", "tpl",
-                                    "gbif", "fb", "slb", "wd")){
+                                    "gbif", "fb", "slb", "wd"),
+                      collect = TRUE){
   
+  ## Simply put input into a table so we can do joins
   if(is.null(id) & !is.null(species)){
     df <- dplyr::tibble(species)
   } else if(!is.null(id) & is.null(species)){
@@ -29,20 +58,36 @@ hierarchy <- function(species = NULL,
     stop("id or species list must be provided")
   }
   
-  
+  ## A `join` is theoretically the fastest way to query for a large number
[…]

### 2018-06-30 README.Rmd
tidy up docs

@@
 library(taxald)
+```
 
+```{r eval=FALSE}
 create_taxadb()
@@
 
+```{r}
+library(tictoc) # let's display timing of queries for reference
+```
+
 Once the database is installed, we can start to make some queries.
@@
 ```{r}
+tic()
 df <- descendents(name = "Aves", rank = "class", authority="col")
+toc()
+```
+
+How many species did we get?
+
+```{r}
+length(df$name)
+```
+
+In general, this species list could have come from anywhere (i.e. some particular research project) and not from a single authority.
+
+We can get the full hierachical classification for this list of species:
+
+```{r}
+tic()
+hierarchy(df$name)
+toc()
 ```

### 2018-06-30 R/create_db.R

@@ -11,8 +11,8 @@
 #'
 #' @examples
 #' \dontrun{
-#'   tmp <- tempdir()
-#'   create_taxadb(tmp, authorities = "itis")
+#'   # tmp <- tempdir()
+#'   # create_taxadb(tmp, authorities = "itis")
 #'
 #' }
 create_taxadb <- function(path = fs::path(fs::path_home(), ".taxald"),

### 2018-06-30 R/taxald.R

@@ -86,7 +86,6 @@ hierarchy <- function(species = NULL,
 ids <- function(name = NULL,
                 authority = c("itis", "ncbi", "col", "tpl",
                               "gbif", "fb", "slb", "wd"),
-                schema = "taxonid",
                 collect = TRUE){
 
   
@@ -174,6 +173,7 @@ connect_db <- function(dbdir = Sys.getenv("TAXALD_HOME",
 #' 
 #' @param db a connection to the taxald database. Default will
 #' attempt to connect automatically.
+#' @param schema the table schema on which we want to run the query
 #' @importFrom dplyr tbl
 #' @inheritParams hierarchy
 #' @export 
@@ -189,3 +189,8 @@ taxa_tbl <- function(
   
   dplyr::tbl(db, tbl_name)
 }
+
+# FIXME: Provide a way to close a connection that is opened automatically
+# may it is it sufficient to do:
+#     db <- connect_db()
+#     DBI::dbDisconnect(db$con)

### 2018-06-30 tests/testthat/test-taxald.R

@@ -1,0 +1,3 @@
+
+
+library(taxald)

### 2018-07-31 README.Rmd
minimal examples

@@
 
-By default, this will install the database into a hidden `.taxald` folder in your home directory (`~/.taxald`).
+The default behavior installs only the ITIS database using the hierachy schema.
+While this should be sufficient for getting started on basic tasks like looking up classification or translating between IDs and scientific names, users will want to install additional authorities and schemas for advanced use.
+See the `authorities` vignette for details.
+By default, `taxald` will install the database into a hidden `.taxald` folder in your home directory (`~/.taxald`).
 You can control this by passing the path to `create_taxadb()` or setting the environmental variable `TAXALD_HOME` to a different location.
@@
 
-First, let's get a nice big species list, say, all the birds (known to the Catalogue of Life):
+First, let's get a nice big species list, say, all the birds (known to ITIS):
 
@@
 tic()
-df <- descendents(name = "Aves", rank = "class", authority="col")
+df <- descendants(name = "Aves", rank = "class")
 toc()
@@
 ```{r}
-length(df$name)
+length(df$species)
 ```
 
-In general, this species list could have come from anywhere (i.e. some particular research project) and not from a single authority.
+Generally, we have a species list could have come from some particular research project, and the names may not match those of any one particular authority.
+
+```{r}
+tic()
+ids(df$species)
+toc()
+```
 
@@
 tic()
-hierarchy(df$name)
+classification(df$species)
 toc()

### 2018-07-31 R/taxa_tbl.R

@@ -1,0 +1,58 @@
+
+#' Return a reference to a given table in the taxald database
+#' 
+#' @param db a connection to the taxald database. Default will
+#' attempt to connect automatically.
+#' @param schema the table schema on which we want to run the query
+#' @importFrom dplyr tbl
+#' @inheritParams classification
+#' @export 
+taxa_tbl <- function(
+  authority = c("itis", "ncbi", "col", "tpl",
+                "gbif", "fb", "slb", "wd"), 
+  schema = c("hierarchy", "taxonid", "synonyms", "common", "long"),
+  db = connect_db()){
+  
+  authority <- match.arg(authority)
+  schema <- match.arg(schema)
+  tbl_name <- paste(authority, schema, sep = "_")
+  
+  dplyr::tbl(db, tbl_name)
+}
+
+
+
+#' Connect to the taxald database
+#' 
+#' @param dbdir Path to the database. Defaults to `TAXALD_HOME` 
+#' environmental variable, which defaults to `~/.taxald`.
+#' @return Returns a `src_dbi` connection to the database
+#' @details Primarily useful when a lower-level interface to the
+#' database is required.  Most `taxald`` functions will connect
+#' automatically without the user needing to call this function.
+#' @importFrom DBI dbConnect
+#' @importFrom MonetDBLite MonetDBLite
+#' @importFrom dbplyr src_dbi
+#' @export
+#' @examples \dontrun{
+#' 
+#' db <- connect_db()
+#' 
+#' }
+connect_db <- function(dbdir = Sys.getenv("TAXALD_HOME", 
+                                          fs::path(fs::path_home(),
+                                                   ".taxald"))){
+  con <- DBI::dbConnect(MonetDBLite::MonetDBLite(), dbdir)
+  db <- dbplyr::src_dbi(con)
+}
+
+
+
+
+# tibble doesn't like null arguments
+#' @importFrom tibble lst tibble
+null_tibble <- function(...){
+  call <- Filter(Negate(is.null), tibble::lst(...))
+  do.call(tibble::tibble, call)
+}
+

### 2018-07-31 R/taxize.R

@@ -1,0 +1,11 @@
+## Drop in replacements for taxize
+
+
+get_ids <- function(names, db, ...){
+  ids(name = names, authority = db, ...)
+}
+
+
+get_uid <- function(sciname, ...){
+  ids(name = sciname, authority = "ncbi", ...)
+}

### 2018-08-02 README.Rmd
readme

@@
 
-Central to `taxald` design is to create a *local* database of all avialable taxonomic authority data that we can query quickly, even when working with tens of thousands of species.
+Central to `taxald` design is to create a *local* database of all available taxonomic authority data that we can query quickly, even when working with tens of thousands of species.
 The second key feature of `taxald` design is to return queries from each authority using a *consistent* and convenient table structure, facilitating queries that compare, combine, or work across names from multiple authorities.
@@
 
+```{r}
+library(taxald)
+library(tictoc) # let's display timing of queries for reference
+```
+
 Before we can use most `taxald` functions, we need to do a one-time installation of the database `taxald` uses for almost all commands.
@@
 
-```{r}
-library(taxald)
-```
-
-```{r eval=FALSE}
+```{r eval = TRUE, message=FALSE}
+tic()
 create_taxadb()
+toc()
 ```
@@
 ## Package API
-
-```{r}
-library(tictoc) # let's display timing of queries for reference
-```
 

### 2018-09-20 R/taxald.R
clean up
lighter dependencies, cleaner flow.

@@ -48,7 +48,7 @@
 #' extremely limited. Under most use cases, the defaults for `collect` and 
 #' `taxald_db` should be appropriate.  
 #' @export
-#' @importFrom dplyr right_join tibble collect
+#' @importFrom dplyr semi_join tibble collect
 #' 
 classification <- function(species = NULL, 
                            id = NULL, 
@@ -57,7 +57,7 @@ classification <- function(species = NULL,
                            collect = TRUE,
                            taxald_db = connect_db()){
   
-  out <- dplyr::right_join(taxa_tbl(authority = authority,
+  out <- dplyr::semi_join(taxa_tbl(authority = authority,
                                     schema = "hierarchy", 
                                     db = taxald_db), 
                            null_tibble(id, species), 
@@ -65,7 +65,7 @@ classification <- function(species = NULL,
   
   if(collect){ ## Return an in-memory object
     out <- dplyr::collect(out)
-    DBI::dbDisconnect(taxald_db$con)
+    DBI::dbDisconnect(taxald_db)
   }
   
   out
@@ -91,18 +91,18 @@ ids <- function(name = NULL,
                 taxald_db = connect_db()){
   
   out <- 
-    dplyr::right_join(
+    dplyr::semi_join(
       taxa_tbl(authority = authority, 
                schema = "hierarchy", 
                db = taxald_db), 
       dplyr::tibble(species = name),
-      copy = TRUE) %>% 
-    dplyr::select("id", "species")
+      copy = TRUE) 
+  out <- dplyr::select(out, "id", "species")
   
   
   if(collect){ ## Return an in-memory object
     out <- dplyr::collect(out)
-    DBI::dbDisconnect(taxald_db$con)
+    DBI::dbDisconnect(taxald_db)
   }
   
   out
@@ -118,8 +118,8 @@ ids <- function(name = NULL,
 #' @export
 #' @importFrom stats setNames
 # @importFrom rlang !! := UQ quo enquo
-#' @importFrom magrittr %>%
-#' @importFrom dplyr right_join select filter distinct
+# @importFrom magrittr %>%
+#' @importFrom dplyr semi_join select filter distinct
 descendants <- function(name = NULL, 
                         rank = NULL, 
                         id = NULL,
@@ -131,10 +131,11 @@ descendants <- function(name = NULL,
 
   ## technically could guess rank from name most but not all time
[…]

### 2018-09-20 tests/testthat/test-taxald.R

@@ -1,18 +1,17 @@
+context("taxald")
+
+library(testthat)
 library(taxald)
 library(dplyr)
-library(tictoc)
 
 
-testthat::test_that("we can set up a db and call basic functions", {
+test_that("we can set up a db and call basic functions", {
   
   tmp <- tempdir()
   Sys.setenv(TAXALD_HOME=tmp)
   
-  tic()
   create_taxadb()
-  toc()
-  
-  tic()
+
   df <- taxa_tbl(authority = "itis", schema = "hierarchy")
   chameleons <- df %>% filter(family == "Chamaeleonidae") %>% collect()
   
@@ -20,15 +19,13 @@ testthat::test_that("we can set up a db and call basic functions", {
   species <- ids(df$species)
   hier <- classification(df$species)
   
-  testthat::expect_is(df, "data.frame")
-  testthat::expect_is(species, "data.frame")
-  testthat::expect_is(hier, "data.frame")
-  testthat::expect_is(chameleons, "data.frame")
-  
-  testthat::expect_gt(dim(df)[1], 1)
-  testthat::expect_gt(dim(species)[1], 1)
-  testthat::expect_gt(dim(hier)[1], 1)
-  testthat::expect_gt(dim(chameleons)[1], 1)
-  toc()
+  expect_is(df, "data.frame")
+  expect_is(species, "data.frame")
+  expect_is(hier, "data.frame")
+  expect_is(chameleons, "data.frame")
+  expect_gt(dim(df)[1], 1)
+  expect_gt(dim(species)[1], 1)
+  expect_gt(dim(hier)[1], 1)
+  expect_gt(dim(chameleons)[1], 1)
 
 })

### 2018-09-21 R/classification.R
refactor
- no default disk location
- td_home() sets home location
- file names by function name
- more portable an consistent fn names

@@ -1,0 +1,62 @@
+
+
+#' Return the full classification hierarchy for requested species or ids
+#' 
+#' @param species a character vector of species names, typically 
+#'  specified as (`Genus species` or `Genus species epithet`)
+#' @param id alternately users can provide a vector of species ids.
+#'  IDs must be prefixed matching the requested authority.  See `id`
+#'  column returned by most `taxald` functions for examples.
+#' @param authority from which authority should the hierachy be returned?
+#'  Default is 'itis'.  
+#' @param collect logical, default `TRUE`. Should we return an in-memory
+#' data.frame (default, usually the most convenient), or a reference to
+#' lazy-eval table on disk (useful for very large tables on which we may
+#' first perform subsequent filtering operations.) 
+#' @param db a connection to the taxald database. See details.
+#' @return a data.frame with one row for each requested species, 
+#'  giving the species id, and column for each of the unique-rank
+#'  levels of the species.  Note that the different authorities recognize
+#'  a range of different ranks.
+#'  
+#' @details Some authorities recognize multiple values of the same rank
+#' level, (i.e. a species may be assigned to two different "suborders").  
+#' In this case, only one has been included in the "hierarchy" schema.
+#' Likewise, some authorities (i.e. NCBI) use some unnamed rank levels 
+#' (i.e. a scientific name is associated with a clade, but the clade is 
+#' not associated with any traditionally recognized rank). In these cases, 
+#' see the "long" schema for more complete classification.  
+#' 
+#' If setting `collect = FALSE`, consider calling `db = td_connect()`
+#' in a separate call and passing the resulting connection object, `db`
+#' explicitly to all subsequent `taxald` functions.  This will allow them
+#' to re-use the existing connection, which can also be used in conjunction
+#' with the returned results for further on-disk queries. In general, this will
+#' only be desirable when tables are extremely large or availability memory is 
+#' extremely limited. Under most use cases, the defaults for `collect` and 
+#' `db` should be appropriate.  
+#' @export
+#' @importFrom dplyr semi_join tibble collect
+#' 
+classification <- function(species = NULL, 
+                           id = NULL, 
+                           authority = c("itis", "ncbi", "col", "tpl",
+                                         "gbif", "fb", "slb", "wd"),
+                           collect = TRUE,
+                           db = td_connect()){
+  
+  out <- dplyr::semi_join(taxa_tbl(authority = authority,
[…]

### 2018-09-21 R/ids.R

@@ -1,0 +1,35 @@
+## FIXME Does not support lookup of non-species-level ids. Using taxonid schema would fix this.
+#' Return taxonomic identifiers from a given namespace
+#' 
+#' @param name a character vector of species names. 
+#' (Most authorities can also return ids for higher-level
+#'  taxonomic names).
+#' @inheritParams classification
+#' @return a data.frame with columns of `id`, scientific 
+#' `name`, and `rank` and a row for each species name queried.
+#' 
+#' @export
+ids <- function(name = NULL,
+                authority = c("itis", "ncbi", "col", "tpl",
+                              "gbif", "fb", "slb", "wd"),
+                collect = TRUE,
+                db = td_connect()){
+  
+  out <- 
+    dplyr::semi_join(
+      taxa_tbl(authority = authority, 
+               schema = "hierarchy", 
+               db = db), 
+      dplyr::tibble(species = name),
+      copy = TRUE) 
+  out <- dplyr::select(out, "id", "species")
+  
+  
+  if(collect && inherits(db, "DBIConnection")){
+    ## Return an in-memory object
+    out <- dplyr::collect(out)
+    DBI::dbDisconnect(db)
+  }
+  
+  out
+}

### 2018-09-21 R/taxald.R

@@ -12,146 +12,3 @@
 
 
 
-#' Return the full classification hierarchy for requested species or ids
-#' 
-#' @param species a character vector of species names, typically 
-#'  specified as (`Genus species` or `Genus species epithet`)
-#' @param id alternately users can provide a vector of species ids.
-#'  IDs must be prefixed matching the requested authority.  See `id`
-#'  column returned by most `taxald` functions for examples.
-#' @param authority from which authority should the hierachy be returned?
-#'  Default is 'itis'.  
-#' @param collect logical, default `TRUE`. Should we return an in-memory
-#' data.frame (default, usually the most convenient), or a reference to
-#' lazy-eval table on disk (useful for very large tables on which we may
-#' first perform subsequent filtering operations.) 
-#' @param taxald_db a connection to the taxald database. See details.
-#' @return a data.frame with one row for each requested species, 
-#'  giving the species id, and column for each of the unique-rank
-#'  levels of the species.  Note that the different authorities recognize
-#'  a range of different ranks.
-#'  
-#' @details Some authorities recognize multiple values of the same rank
-#' level, (i.e. a species may be assigned to two different "suborders").  
-#' In this case, only one has been included in the "hierarchy" schema.
-#' Likewise, some authorities (i.e. NCBI) use some unnamed rank levels 
-#' (i.e. a scientific name is associated with a clade, but the clade is 
-#' not associated with any traditionally recognized rank). In these cases, 
-#' see the "long" schema for more complete classification.  
-#' 
-#' If setting `collect = FALSE`, consider calling `taxald_db = connect_db()`
-#' in a separate call and passing the resulting connection object, `taxald_db`
-#' explicitly to all subsequent `taxald` functions.  This will allow them
-#' to re-use the existing connection, which can also be used in conjunction
-#' with the returned results for further on-disk queries. In general, this will
-#' only be desirable when tables are extremely large or availability memory is 
-#' extremely limited. Under most use cases, the defaults for `collect` and 
-#' `taxald_db` should be appropriate.  
-#' @export
-#' @importFrom dplyr semi_join tibble collect
-#' 
-classification <- function(species = NULL, 
-                           id = NULL, 
-                           authority = c("itis", "ncbi", "col", "tpl",
[…]

### 2018-09-21 R/td_create.R

@@ -1,12 +1,8 @@
-# FIXME consider a lightweight create_db that imports a single table,
-# from a single schema, possibly directly into memory?
 
-
- 
 #' create a local taxonomic database
 #'
-#' @param path a location on your computer where the database should be installed.
-#'  By default, will install to `.taxald` in your home directory.
+#' @param dbdir a location on your computer where the database should be installed.
+#'  By default, will look for a location given by `Sys.getenv("TAXALD_HOME")` if not specified.
 #' @param authorities a list (character vector) of authorities to be included in the
 #'  database. By default, will install `itis`.  See details for a list of recognized
 #'  authorities. Use `authorities="all"` to install all available authorities automatically.
@@ -28,7 +24,6 @@
 #'  - `wd`, wikidata
 #' @return path where database has been installed (invisibly)
 #' @export
-#' @aliases create_db create_taxadb
 #' @importFrom utils download.file
 #' @importFrom DBI dbConnect dbDisconnect
 #' @importFrom arkdb unark
@@ -39,13 +34,14 @@
 #'   # create_db(tmp, authorities = "itis")
 #'
 #' }
-create_db <- function(authorities = "itis", 
+td_create <- function(dbdir = Sys.getenv("TAXALD_HOME"), 
+                      authorities = "itis", 
                       schema = "hierarchy",
                       overwrite = FALSE,
-                      lines = 1e6,
-                      path = Sys.getenv("TAXALD_HOME", 
-                                        file.path(path.expand("~"),
-                                                 ".taxald"))){
+                      lines = 1e6
+                      ){
+  
+  dbdir <- td_home(dbdir)
   recognized_authorities = c("itis", 
                              "ncbi", 
                              "col", 
@@ -84,8 +80,8 @@ create_db <- function(authorities = "itis",
   dir.create(file.path(tmp, "taxald"), FALSE)
   utils::download.file(urls, file.path(tmp, "taxald", files))
   
-  dir.create(path, FALSE)
-  con <- DBI::dbConnect(MonetDBLite::MonetDBLite(), path)
+  dir.create(dbdir, FALSE)
+  con <- DBI::dbConnect(MonetDBLite::MonetDBLite(), dbdir)
   
   ## silence readr progress bar in arkdb
   progress <- getOption("readr.show_progress")
@@ -108,11 +104,9 @@ create_db <- function(authorities = "itis",
[…]

### 2018-09-21 R/td_home.R

@@ -1,0 +1,25 @@
+
+#' td_home
+#' @param dbdir a location on your computer where the database should be installed.
+#' By default, will look for a location given by `Sys.getenv("TAXALD_HOME")` if not specified.
+#' @param create should the location be created if no directory exits? 
+#' @return dbdir path that will be used.
+td_home <- function(dbdir = Sys.getenv("TAXALD_HOME"), create = TRUE){
+  if(dbdir == ""){
+    stop(paste0(
+      'Please set a location for the taxald database by running:\n\n\t',
+      'td_home("~/.taxald")', '\n\n',
+      'or setting the env variable TAXALD_HOME to the appropriate ',
+      'location'), call. = FALSE)
+  }
+  if(!dir.exists(dbdir) && create){
+    dir.create(dbdir, showWarnings = FALSE, recursive = TRUE)
+  }
+
+  ## Persist setting throughout session
+  Sys.setenv(TAXALD_HOME=dbdir)
+  
+  ## FIXME Should we also set this more persistently, e.g. to an Renviron file?
+  
+  dbdir
+}

### 2018-09-21 tests/testthat/test-all-dbs.R

@@ -12,8 +12,21 @@ test_that("we can set up a db and call basic functions", {
   skip_on_cran()
   
   
-  create_taxadb("all")
 
+  td_create(authorities = "itis")
+  td_create(authorities = "ncbi")
+  td_create(authorities = "col")
+  td_create(authorities = "tbl")
+  td_create(authorities = "gbif")
+  td_create(authorities = "fb")
+  td_create(authorities = "slb")
+  #td_create(authorities = "wd")
+  
+  
+  td_create(authorities = "all")
+
+  
+  
   df <- taxa_tbl(authority = "col", schema = "hierarchy")
   chameleons <- df %>% filter(family == "Chamaeleonidae") %>% collect()
   

### 2018-09-21 tests/testthat/test-taxald.R

@@ -8,9 +8,7 @@ library(dplyr)
 test_that("we can set up a db and call basic functions", {
   
   tmp <- tempdir()
-  Sys.setenv(TAXALD_HOME=tmp)
-  
-  create_taxadb()
+  td_create(tmp)
 
   df <- taxa_tbl(authority = "itis", schema = "hierarchy")
   chameleons <- df %>% filter(family == "Chamaeleonidae") %>% collect()

### 2018-09-28 README.Rmd
rebuild docs
clean up
lighter dependencies, cleaner flow.
refactor
- no default disk location
- td_home() sets home location
- file names by function name
- more portable an consistent fn names
progress to simplifying and abstracting db connection

@@
 The database is installed on your local harddisk and will persist between R sessions.
-Download and install the database like so:
+By default, the database will be installed in your user's application data directory, as detected by the `rappdirs` package.
+(Set a custom location instead using `dbdir` argument.)
 
-```{r eval = TRUE, message=FALSE}
-tic()
-create_taxadb()
-toc()
+```{r }
+td_create()
 ```
 
-The default behavior installs only the ITIS database using the hierachy schema.
+The default behavior installs only the ITIS database.
 While this should be sufficient for getting started on basic tasks like looking up classification or translating between IDs and scientific names, users will want to install additional authorities and schemas for advanced use.

### 2018-09-28 R/classification.R

@@ -51,10 +51,9 @@ classification <- function(species = NULL,
                           null_tibble(id, species), 
                           copy = TRUE)
   
-  if(collect && inherits(db, "DBIConnection")){ 
+  if(collect && inherits(out, "tbl_lazy")){ 
     ## Return an in-memory object
     out <- dplyr::collect(out)
-    DBI::dbDisconnect(db)
   }
   
   out

### 2018-09-28 R/descendants.R

@@ -22,21 +22,22 @@ descendants <- function(name = NULL,
   ## technically could guess rank from name most but not all time
   ## could still do this as join rather than a filter with appropriate table construction
   
-  if(schema == "hierarchy"){
     df <- data.frame(setNames(list(name),  rank), stringsAsFactors = FALSE)
     df$id <- id
-    out <- dplyr::semi_join(
-      taxa_tbl(authority = authority,
-               schema = "hierarchy", 
-               db = db),
-      df,
-      copy = TRUE, by = rank)
-  }
+    
+    taxa <- taxa_tbl(authority = authority,
+                     schema = "hierarchy", 
+                     db = db)
+    
+    out <- dplyr::semi_join(taxa, 
+                            df, 
+                            copy = TRUE, 
+                            by = rank)
   
-  if(collect && inherits(db, "DBIConnection")){
+  
+  if(collect && inherits(out, "tbl_lazy")){
     ## Return an in-memory object
     out <- dplyr::collect(out)
-    DBI::dbDisconnect(db)
   }
   
   out

### 2018-09-28 R/ids.R

@@ -26,10 +26,9 @@ ids <- function(name = NULL,
   out <- dplyr::select(out, "id", "species")
   
   
-  if(collect && inherits(db, "DBIConnection")){
+  if(collect && inherits(out, "tbl_lazy")){
     ## Return an in-memory object
     out <- dplyr::collect(out)
-    DBI::dbDisconnect(db)
   }
   
   out

### 2018-09-28 R/synonyms.R

@@ -29,10 +29,9 @@ synonyms <- function(name = NULL,
   
   ## ITIS seems to map synonyms that are obviously species names into higher ranks??
   
-  if(collect && inherits(db, "DBIConnection")){ 
+  if(collect && inherits(out, "tbl_lazy")){ 
     ## Return an in-memory object
     out <- dplyr::collect(out)
-    DBI::dbDisconnect(db)
   }
   
   out

### 2018-09-28 R/td_create.R

@@ -1,8 +1,7 @@
 
 #' create a local taxonomic database
 #'
-#' @param dbdir a location on your computer where the database should be installed.
-#'  By default, will look for a location given by `Sys.getenv("TAXALD_HOME")` if not specified.
+
 #' @param authorities a list (character vector) of authorities to be included in the
 #'  database. By default, will install `itis`.  See details for a list of recognized
 #'  authorities. Use `authorities="all"` to install all available authorities automatically.
@@ -12,6 +11,10 @@
 #' @param overwrite Should we overwrite existing tables? Default is `FALSE`.  Change
 #'  to "ask" for interactive interface, or `TRUE` to force overwrite (i.e. updating
 #'  a local database upon new release.)
+#' @param dbdir a location on your computer where the database should be installed.
+#'  Defaults to user data directory given by [rappdirs::user_data_dir]().
+#' @param db connection to a database.  By default, taxald will set up its own
+#' fast [MonetDBLite::MonetDBLite]() connection.  
 #' @details 
 #'  Authorities recognized by taxald are:
 #'  - `itis` 
@@ -37,10 +40,11 @@
 td_create <- function(authorities = "itis", 
                       schema = "hierarchy",
                       overwrite = FALSE,
-                      dbdir = td_home(),
-                      lines = 1e6
+                      lines = 1e6,
+                      dbdir =  rappdirs::user_data_dir("taxald"),
+                      db = td_connect(dbdir)
                       ){
-  
+
   recognized_authorities = c("itis", 
                              "ncbi", 
                              "col", 
@@ -49,56 +53,51 @@ td_create <- function(authorities = "itis",
                              "fb", 
                              "slb", 
                              "wd")
+  recognized_schema = c("hierarchy", "taxonid", "synonyms")
   if(authorities == "all"){
     authorities <- recognized_authorities
   }
-  
   stopifnot(all(authorities %in% recognized_authorities))
+  stopifnot(all(schema %in% recognized_schema))
   
-  ## FIXME include some messaging about the large downloads etc?
-  
-  files <- vapply(schema, 
-                  function(s) paste0(authorities, "_", s, ".tsv.bz2"),
+  ## supports vectorized schema and authorities lists.  
+  files <- vapply(schema, function(s) 
[…]

### 2018-09-28 R/td_home.R

@@ -1,29 +1,0 @@
-
-#' td_home
-#' 
-#' Set taxald home directory where database will be stored
-#' @param dbdir a location on your computer where the database should be installed.
-#' By default, will look for a location given by `Sys.getenv("TAXALD_HOME")` if not specified.
-#' @param create should the location be created if no directory exits? 
-#' @return dbdir path that will be used.
-#' @export
-#' @importFrom rappdirs user_data_dir
-td_home <- function(dbdir = rappdirs::user_data_dir("taxald"), create = TRUE){
-  if(dbdir == ""){
-    stop(paste0(
-      'Please set a location for the taxald database by running:\n\n\t',
-      'td_home("~/.taxald")', '\n\n',
-      'or setting the env variable TAXALD_HOME to the appropriate ',
-      'location'), call. = FALSE)
-  }
-  if(!dir.exists(dbdir) && create){
-    dir.create(dbdir, showWarnings = FALSE, recursive = TRUE)
-  }
-
-  ## Persist setting throughout session
-  Sys.setenv(TAXALD_HOME=dbdir)
-  
-  ## FIXME Should we also set this more persistently, e.g. to an Renviron file?
-  
-  dbdir
-}

### 2018-09-28 tests/testthat/test-connections.R

@@ -1,0 +1,41 @@
+context("test connections")
+
+
+
+test_that("we can use alternative DBs, such as SQLite", {
+  
+  skip_if(TRUE)
+  skip_on_travis()
+  skip_on_cran()
+  
+## NOTE: SQLite joins are waaaay slower than MonetDBLite
+dbdir <- rappdirs::user_data_dir("taxald")
+dir.create(dbdir, showWarnings = FALSE)
+dbname <- file.path(dbdir, "taxald.sqlite")
+
+db <- DBI::dbConnect(RSQLite::SQLite(), dbname = dbname)
+                     
+  
+  td_create(db = db)
+  td_create("ncbi", db = db)
+
+  itis <- taxa_tbl(authority = "itis", 
+                 schema = "hierarchy", 
+                 db = db)
+  ncbi <- taxa_tbl(authority = "ncbi", 
+                 schema = "hierarchy", 
+                 db = db)
+
+  birds <- descendants(name = "Aves", 
+                    rank = "class", 
+                    db = db)
+  mammals <- descendants(name = "Mammalia", 
+                    rank = "class", 
+                    db = db)
+  
+  ids <- ids(birds$species, 
+                 db = db)
+  hier <- classification(birds$species, 
+                         db = db)
+
+})

### 2018-09-28 tests/testthat/test-fast.R

@@ -7,8 +7,6 @@ library(dplyr)
 
 test_that("setup-free calls to basic functions", {
   
-  td_home(tempdir()) # Sets a location for peristent cache
-  
   df <- taxa_tbl(authority = "itis", schema = "hierarchy", db = NULL)
   chameleons <- filter(df, family == "Chamaeleonidae")
   

### 2018-09-28 tests/testthat/test-taxald.R

@@ -8,12 +8,14 @@ library(dplyr)
 test_that("we can set up a db and call basic functions", {
   
   tmp <- file.path(tempdir(), "taxald")
+  dir.create(tmp, showWarnings = FALSE)
+  
   td_create(dbdir = tmp)
   db <- td_connect(tmp)
 
   df <- taxa_tbl(authority = "itis", 
                  schema = "hierarchy", 
-                 db = td_connect(tmp))
+                 db = db)
   
   chameleons <- df %>% 
     filter(family == "Chamaeleonidae") %>% 
@@ -21,11 +23,11 @@ test_that("we can set up a db and call basic functions", {
   
   df <- descendants(name = "Aves", 
                     rank = "class", 
-                    db = td_connect(tmp))
+                    db = db)
   species <- ids(df$species, 
                  db = td_connect(tmp))
   hier <- classification(df$species, 
-                         db = td_connect(tmp))
+                         db = db)
   
   expect_is(df, "data.frame")
   expect_is(species, "data.frame")

### 2018-10-10 R/classification.R
WIP

@@ -1,61 +1,47 @@
 
 
 #' Return the full classification hierarchy for requested species or ids
-#' 
-#' @param species a character vector of species names, typically 
+#'
+#' @param species a character vector of species names, typically
 #'  specified as (`Genus species` or `Genus species epithet`)
 #' @param id alternately users can provide a vector of species ids.
 #'  IDs must be prefixed matching the requested authority.  See `id`
 #'  column returned by most `taxald` functions for examples.
-#' @param authority from which authority should the hierachy be returned?
-#'  Default is 'itis'.  
-#' @param collect logical, default `TRUE`. Should we return an in-memory
-#' data.frame (default, usually the most convenient), or a reference to
-#' lazy-eval table on disk (useful for very large tables on which we may
-#' first perform subsequent filtering operations.) 
-#' @param db a connection to the taxald database. See details.
-#' @return a data.frame with one row for each requested species, 
+#' @inheritParams ids
+#' @return a data.frame with one row for each requested species,
 #'  giving the species id, and column for each of the unique-rank
 #'  levels of the species.  Note that the different authorities recognize
 #'  a range of different ranks.
-#'  
+#'
 #' @details Some authorities recognize multiple values of the same rank
-#' level, (i.e. a species may be assigned to two different "suborders").  
+#' level, (i.e. a species may be assigned to two different "suborders").
 #' In this case, only one has been included in the "hierarchy" schema.
-#' Likewise, some authorities (i.e. NCBI) use some unnamed rank levels 
-#' (i.e. a scientific name is associated with a clade, but the clade is 
-#' not associated with any traditionally recognized rank). In these cases, 
-#' see the "long" schema for more complete classification.  
-#' 
-#' If setting `collect = FALSE`, consider calling `db = td_connect()`
-#' in a separate call and passing the resulting connection object, `db`
-#' explicitly to all subsequent `taxald` functions.  This will allow them
-#' to re-use the existing connection, which can also be used in conjunction
-#' with the returned results for further on-disk queries. In general, this will
-#' only be desirable when tables are extremely large or availability memory is 
-#' extremely limited. Under most use cases, the defaults for `collect` and 
-#' `db` should be appropriate.  
+#' Likewise, some authorities (i.e. NCBI) use some unnamed rank levels
+#' (i.e. a scientific name is associated with a clade, but the clade is
+#' not associated with any traditionally recognized rank). In these cases,
+#' see the "long" schema for more complete classification.
+#'
 #' @export
[…]

### 2018-10-10 R/td_create.R

@@ -7,17 +7,17 @@
 #'  authorities. Use `authorities="all"` to install all available authorities automatically.
 #' @param schema format of the database to import.
 #' @param lines number of lines that can be safely read in to memory at once. Leave
-#' at default or increase for faster importing if you have plenty of spare RAM. 
+#' at default or increase for faster importing if you have plenty of spare RAM.
 #' @param overwrite Should we overwrite existing tables? Default is `FALSE`.  Change
 #'  to "ask" for interactive interface, or `TRUE` to force overwrite (i.e. updating
 #'  a local database upon new release.)
 #' @param dbdir a location on your computer where the database should be installed.
 #'  Defaults to user data directory given by [rappdirs::user_data_dir]().
 #' @param db connection to a database.  By default, taxald will set up its own
-#' fast [MonetDBLite::MonetDBLite]() connection.  
-#' @details 
+#' fast [MonetDBLite::MonetDBLite]() connection.
+#' @details
 #'  Authorities recognized by taxald are:
-#'  - `itis` 
+#'  - `itis`
 #'  - `ncbi`
 #'  - `col`
 #'  - `tpl`
@@ -38,43 +38,43 @@
 #'   # create_db(authorities = "itis", dbdir = tmp)
 #'
 #' }
-td_create <- function(authorities = "itis", 
-                      schema = "hierarchy",
+td_create <- function(authorities = "itis",
+                      schema = c("hierarchy", "taxonid", "synonyms"),
                       overwrite = FALSE,
                       lines = 1e6,
                       dbdir =  rappdirs::user_data_dir("taxald"),
                       db = td_connect(dbdir)
                       ){
 
-  recognized_authorities = c("itis", 
-                             "ncbi", 
-                             "col", 
-                             "tpl", 
-                             "gbif", 
-                             "fb", 
-                             "slb", 
+  recognized_authorities = c("itis",
+                             "ncbi",
+                             "col",
+                             "tpl",
+                             "gbif",
+                             "fb",
+                             "slb",
                              "wd")
   recognized_schema = c("hierarchy", "taxonid", "synonyms")
-  if(authorities == "all"){
+  if (authorities == "all") {
     authorities <- recognized_authorities
   }
   stopifnot(all(authorities %in% recognized_authorities))
[…]

### 2018-10-10 tests/testthat/test-all-dbs.R

@@ -6,34 +6,35 @@ library(dplyr)
 
 
 test_that("we can set up a db and call basic functions", {
-  
+
   skip_if(TRUE)
   skip_on_travis()
   skip_on_cran()
-  
+
   schema <- c("hierarchy", "taxonid", "synonyms")
+  schema2 <- c("hierarchy", "taxonid")
   overwrite = TRUE
-  td_create(authorities = "itis", schema = schema, overwrite = overwrite)
-  td_create(authorities = "ncbi", schema = schema, overwrite = overwrite)
-  td_create(authorities = "col", schema = schema)
-  td_create(authorities = "tpl")
-  td_create(authorities = "gbif", schema = c("hierarchy"))
-  td_create(authorities = "fb")
+  td_create(authorities = "itis")
+  td_create(authorities = "ncbi")
+  td_create(authorities = "col")
+  td_create(authorities = "gbif")
+  td_create(authorities = "fb",)
   td_create(authorities = "slb")
   td_create(authorities = "wd")
-  
-  
+  td_create(authorities = "tpl")
+
+
   td_create(authorities = "all")
 
-  
-  
+
+
   df <- taxa_tbl(authority = "col", schema = "hierarchy")
   chameleons <- df %>% filter(family == "Chamaeleonidae") %>% collect()
-  
+
   df <- descendants(name = "Aves", rank = "class")
   species <- ids(df$species)
   hier <- classification(df$species)
-  
+
   expect_is(df, "data.frame")
   expect_is(species, "data.frame")
   expect_is(hier, "data.frame")

### 2018-10-10 tests/testthat/test-taxald.R

@@ -6,35 +6,35 @@ library(dplyr)
 
 
 test_that("we can set up a db and call basic functions", {
-  
+
+  taxald:::td_disconnect()
+
   tmp <- file.path(tempdir(), "taxald")
   dir.create(tmp, showWarnings = FALSE)
-  
+
   td_create(dbdir = tmp)
   db <- td_connect(tmp)
 
-  df <- taxa_tbl(authority = "itis", 
-                 schema = "hierarchy", 
+  df <- taxa_tbl(authority = "itis",
                  db = db)
-  
-  chameleons <- df %>% 
-    filter(family == "Chamaeleonidae") %>% 
+
+  chameleons <- df %>%
+    filter(family == "Chamaeleonidae") %>%
     collect()
-  
-  df <- descendants(name = "Aves", 
-                    rank = "class", 
+
+  df <- descendants(name = "Aves",
+                    rank = "class",
                     db = db)
-  species <- ids(df$species, 
+  species <- ids(df$species,
                  db = td_connect(tmp))
-  hier <- classification(df$species, 
+  hier <- classification(df$species,
                          db = db)
-  
+
   expect_is(df, "data.frame")
-  expect_is(species, "data.frame")
+  expect_is(species, "character")
   expect_is(hier, "data.frame")
   expect_is(chameleons, "data.frame")
   expect_gt(dim(df)[1], 1)
-  expect_gt(dim(species)[1], 1)
   expect_gt(dim(hier)[1], 1)
   expect_gt(dim(chameleons)[1], 1)
 

### 2018-10-10 schema.Rmd
rename schema into Rmd, extend

# Database schema for taxald

`taxald` relies on a set of pre-assembled tables following a set of standardized schema layouts outlined below.  The database dumps provided by authorities supported in `taxald` at this time are:

- `itis`: [The Integrated Taxonomic Information System](https://www.itis.gov/)
- `col`: [The Catalogue of Life](http://www.catalogueoflife.org/)
- `ncbi`: [The National Center for Biotechnology Information](https://www.ncbi.nlm.nih.gov/)
- `gbif`: [The Global Biodiversity Information Facility](https://www.gbif.org/)
- `tpl`: [The Plant List](http://www.theplantlist.org/)
- `fb`: [FishBase](https://fishbase.ca)
- `slb`: [SeaLifeBase](https://www.sealifebase.ca/)
- `wd`: [WikiData](https://wikidata.org)

These authorities provide taxonomic data in a wide range of database formats using a wide range of data layouts (schemas), not all of which are particularly easy to use or interpret (e.g. hierarchies are often but not always specified in `taxon_id,parent_id` pairs.)  To make it faster and easier to work across these authorities, `taxald` defines a common set of table schemas outlined below that are particularly suited for efficient computation of common tasks.  `taxald` pre-processes and publicly archives compressed, flat tables corresponding to each of these schema for each of these authorities.

Using the schema defined below, most common operations can be expressed in terms of very standard operations, such as simple filtering joins in SQL.  to implement these, `taxald` imports the compressed flat files into a local, column-oriented database, [`MonetDBLite`](), which can be installed entirely as an R package with no additional server setup required.  This provides a persistant store, and ensures that operations can be performed on disk since the taxonomic tables considered here are frequently too large to store in active memory.  The columnar structure enables blazingly fast joins.  Once the database is created, `taxald` simply wraps a set of user-friendly R functions around common `SQL` queries, implemented in the popular `dplyr` syntax.  By default, `taxald` will always collect the results of these queries to return familiar, in-memory objects to the R user.  Optional arguments allow more direct access the database queries.

This vignette summarizes the table schema defined by `taxald`.  Pre-processing of the original database dumps from each authority into the format described here can be found in the corresponding scripts in `data-raw` directory of the R package source code.

```{r}
library(taxald)
td_create("all")
```

## taxonid schema

A taxonid schema is provided for every authority, containing the columns:

- `id`
- `name`
- `rank`

Some authorities may report additional optional columns, such as the update date.

```{r}
taxa_tbl("itis", "taxonid")
```

Note that `taxald` tables report all identifiers using authority prefixes.  Certain authorities only provide taxanomic identifiers to species names (rank is always species).  Currently this includes `gbif`, `fb`, `slb`.  `ncbi`, `col` and `itis` provide taxonomic identifers to scientific names at any rank.

## Hierarchy table

Any valid rank is a column, and a taxon `id` is a unique key.

First column is named `id`, all subsequent columns are corresponding rank names.  Note that this wide format does not support unnamed rank levels, which are common in certain classifications such as NCBI.  Additional issues arise in some databases (e.g. also in NCBI) where a rank name is duplicated.

Example:

```{r}
taxa_tbl("col", "hierarchy")
```

## Synonyms table

A synonyms table can be build for most but not all authorities. Columns in the synonyms table are:

- `name`: Synonym or any other type (misspelling, alternate name) that is not the accepted name but can be unambiguously matched to an accepted name.   
- `accepted_name`: The corresponding accepted name according to the authority
- `id`: the id of the *accepted name*
- `rank`: the rank of the accepted name.

And optional fields:

- `synonym_id`: Some authorities (e.g. ITIS, FB, COL) provide unique identifiers corresponding to the synonym name.  
- `name_type`: the type of name given in `name`, e.g. `synonym`, `misspelling`, etc.

Example:

```{r}
taxa_tbl("gbif", "synonyms")
```

## Common names

Common names are avialable from several authorities, but tidy tables for `taxald` have not yet been implemented.  Common names tables are expected to follow the following schema:

- `id` The taxonomic identifier for the species (or possibly other rank)
- `name` The common name / vernacular name
- `language` The language in which the common name is given, if known. (all lowercase)
- `language_code` the two-letter language code.

## Long format

Single-table representation of core taxonomic and hierarchical data.  Format includes followings required columns:

id | name | rank  | path    |  path_rank  | path_id
---|------|-------|---------|-------------|---------

And the following optional columns:

 rank_id  | path_rank_id  | name_type   | date
----------|---------------|-------------|---------

(Most databases do not define `rank_id`, e.g. ids corresponding to the ranks themselves)

These tables are considerably larger and longer than other formats, which can make them slow to query.  Note that for each taxonomic name, `id` and the corresponding `name`, and `rank` of the `id` are repeated across the full path hierarchy.  This format is not installed and not used by functions of `taxald` at this time.

A fundamental advantage of this format is that it can be defined identically for every table, meaning that all authorities could be combined into a single table using the long format. both the `taxonid` and `hierarchy` formats can be derived directly from this format, and this format can represent information such as unnamed or duplicate ranks that cannot be represented in the `hierarchy` format.

Example:

```{r}
taxa_tbl("itis", "long", NULL)
```

## Conventions

- `species` is always given as the scientific name associated with the species; e.g. `Homo sapiens` not `sapiens`.  (ITIS convention)
- Identifiers use the integer identifier defined by the authority, prefixed by the authority abbreviation in all capital letters: `ITIS:`, `GBIF:`, etc.
- Rank names are always lower case without hyphens or spaces. Rank names should be mapped
  to a table of standard accepted rank names (i.e. those recognized by ITIS, NCBI, Wikidata),
  and rank names should have 
- Encoding is UTF-8

### 2018-10-11 R/classification.R
vignettes, pkgdown

@@ -31,7 +31,7 @@ classification <- function(species = NULL,
                            collect = TRUE,
                            db = td_connect()){
 
-  out <- dplyr::semi_join(taxa_tbl(authority = authority,
+  out <- dplyr::right_join(taxa_tbl(authority = authority,
                                    schema = "hierarchy",
                                    db = db),
                           null_tibble(id, species),

### 2018-10-11 README.Rmd
fix links

@@
 ```
+
 # taxald
 
-**This package is purley exploratory at this stage!**
+The goal of `taxald` is to provide fast access to taxonomic data and manipulations, such as resolving taxonomic names to ids, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
+These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records.
 
-Central to `taxald` design is to create a *local* database of all available taxonomic authority data that we can query quickly, even when working with tens of thousands of species.
-The second key feature of `taxald` design is to return queries from each authority using a *consistent* and convenient table structure, facilitating queries that compare, combine, or work across names from multiple authorities.
+Existing approaches to these problems typically rely on web APIs, which can make them impractical for work with large numbers of species or in more complex pipelines.
+Queries and returned formats also differ across the different taxonomic authorities, making tasks that query multiple authorities particularly complex.
+`taxald` creates a *local* database of most readily available taxonomic authorities, each of which is transformed into consistent, standard, and researcher-friendly tabular formats.
 
@@
 
-`taxald` should now be installable.  To get started, try:
+To get started, install the development version directly from GitHub:
 
@@
 
-```{r}
-library(taxald)
-library(tictoc) # let's display timing of queries for reference
-```
-
 Before we can use most `taxald` functions, we need to do a one-time installation of the database `taxald` uses for almost all commands.
 This can take a while to run, but needs only be done once.
-The database is installed on your local harddisk and will persist between R sessions.
+The database is installed on your local hard-disk and will persist between R sessions.
 By default, the database will be installed in your user's application data directory, as detected by the `rappdirs` package.
@@
 ```{r }
+library(taxald)
 td_create()
@@
 
-The default behavior installs only the ITIS database.
-While this should be sufficient for getting started on basic tasks like looking up classification or translating between IDs and scientific names, users will want to install additional authorities and schemas for advanced use.
-See the `authorities` vignette for details.
-By default, `taxald` will install the database into a hidden `.taxald` folder in your home directory (`~/.taxald`).
-You can control this by passing the path to `create_taxadb()` or setting the environmental variable `TAXALD_HOME` to a different location.
+The default behavior installs only the ITIS database.  You can also specify a list of authorities to install, or install every authority using `td_create("all")`.
 
-This function downloads each of the individual tables and loads them into a [MonetDBLite](https://www.monetdb.org) database.
-(MonetDBLite is much like SQLite, requiring no external server setup.
-However, for our purposes, MonetDB's columnar design provides significantly faster performance than SQLite, Postgres, or even some in-memory operations in `dplyr`.)
+## Test drive
 
-Currently tables are pulled from the development cache on GitHub.
-Many of these tables are still being cleaned up and standardized, see [schema.md](schema.md).
-Multiple tablesare provided for each authority, though some of these contain redundant information, different orientations may prove more convenient or computationally efficient.
-
-## Package API
-
-Once the database is installed, we can start to make some queries.
-
-First, let's get a nice big species list, say, all the birds (known to ITIS):
+Once the databases have been set up, we're ready to explore.  
+Here's a list of all the birds species known to ITIS:
 
 ```{r}
-tic()
-df <- descendants(name = "Aves", rank = "class")
-toc()
+descendants(name = "Aves", rank = "class")
 ```
 
-How many species did we get?
+## Learn More
 
-```{r}
-length(df$species)
-```
+### [An introduction to taxald](https://cboettig.github.io/taxald/articles/articles/taxald.html)
 
-Generally, we have a species list could have come from some particular research project, and the names may not match those of any one particular authority.
+The [taxald introduction](https://cboettig.github.io/taxald/articles/articles/taxald.html) 
+provides an overview showing how `taxald` functions can help us synthesize data across a given list of species by resolving synonyms and identifiers.
 
-```{r}
-tic()
-ids(df$species)
-toc()
-```
+### [taxald schemas](https://cboettig.github.io/taxald/articles/articles/schema.html)
 
-We can get the full hierachical classification for this list of species:
-
-```{r}
-tic()
-classification(df$species)
-toc()
-```
-
-Design sketch/spec for package API:
-
-- Given vector of taxonomic names, return taxonomic identifiers
-- Given taxonomic identifiers, return heirarchical classification.
-
-(Note that for some authorities, e.g. GBIF and TPL, taxonomic identifiers are only assigned to species names.  So given the name of a higher-level rank, no id could be returned)
-
-- Given any higher-order rank name (e.g. `infraorder`) return identifiers of all member species.
-
-- Given common names, resolve to accepted scientific name
-
-- Resolve any known miss-spelling, recognized synonymous name, generic name or name-part to corresponding taxonomic name.
-
-- Map between identifiers
-
-- Normalize rank names
-
-- Resolve a list of names across multiple authorities to attain higher coverage, 
-- provide merged tables
-
-See [schema.md](schema.md) for a sketch of the underlying database architecture imposed on the data source from each authority.
-
-Note that all inputs and outputs should be vectorized.
+See the [schemas](https://cboettig.github.io/taxald/articles/articles/schema.html) vignette for an overview of the underlying tables used by `taxald` functions, and more about the different authorities accessed by taxald.

### 2018-10-19 R/ids.R
patch ids function
pull method did not return ids in original order!
also, pull=FALSE will now return accepted_name.
Other minor tweaks and updated vignette

@@ -24,7 +24,7 @@
 #' accepted name and the ID associated with the accepted name and return that.
 #'
 #' @export
-#' @importFrom dplyr quo
+#' @importFrom dplyr quo tibble filter right_join
 #' @importFrom rlang !!
 #' @importFrom magrittr %>%
 ids <- function(name = NULL,
@@ -35,31 +35,28 @@ ids <- function(name = NULL,
                 collect = TRUE,
                 db = td_connect()){
 
+  input_table <- dplyr::tibble(name)
+
   ## Use right_join, so unmatched names are kept, with NA
+  ## Using right join, names appear in order of authority
   out <- dplyr::right_join(
       taxa_tbl(authority, "taxonid", db),
-      dplyr::tibble(name),
+      input_table,
       by = "name",
-      copy = TRUE)
+      copy = TRUE) %>%
+    select("id", "name", "rank")
   # input table will be copied in, cleaned on disconnect
-  out <- select(out, "id", "name", "rank")
 
-## SQL execution with larger tables is SO MUCH slower using filter!
-#  system.time({
-#  taxon_names <- name # must not use protected term
-#  out <- dplyr::filter(
-#    taxa_tbl("itis", "taxonid", db),
-#    name %IN% taxon_names)
-#  out <- collect(out)
-#  })
+  ## SQL with larger tables is MUCH slower using filter!
 
   id <- "id"
-  id_column <- quo(id)
+  id_column <- dplyr::quo(id)
 
   if (synonyms_check) {
 
     not_missing <-
-      dplyr::filter(out, !is.na(!!id_column))
+      dplyr::filter(out, !is.na(!!id_column)) %>%
+      dplyr::mutate(accepted_name = name)
 
     missing <-
       filter(out, is.na(!!id_column)) %>%
@@ -69,17 +66,22 @@ ids <- function(name = NULL,
       taxa_tbl(authority, "synonyms", db),
       missing,
       by = "name"
-      ) %>%
-      dplyr::select("name", "id", "rank")
+      )
 
-    out <- dplyr::union(not_missing, syn)
+    out <- dplyr::union(
+      not_missing,
+      dplyr::select(syn, "name", "id", "accepted_name", "rank"))
+
 
   }
 
 
   if (pull) {
-
-    return( dplyr::pull(out, !!id_column) )
+    ## Sort by input if returning a vector
+    id <- input_table %>%
[…]

### 2018-12-06 README.Rmd
a little testing + appveyor

@@
 
-[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
+[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
 [![Travis build status](https://travis-ci.org/cboettig/taxald.svg?branch=master)](https://travis-ci.org/cboettig/taxald)
+[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/taxald?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxald)
 [![Coverage status](https://codecov.io/gh/cboettig/taxald/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxald?branch=master)

### 2018-12-06 tests/testthat/test-get_ids.R

@@ -1,0 +1,12 @@
+context("get_ids")
+
+test_that("we can use get_ids options", {
+
+  bare <- get_ids("Homo sapiens")
+  expect_identical(bare, "ITIS:180092")
+  some_ids <- get_ids(c("Homo sapiens", "Mammalia"), format = "prefix")
+  expect_identical(some_ids,  c("ITIS:180092", "ITIS:179913"))
+  uri <- get_ids("Homo sapiens", db= "ncbi", format = "uri")
+  expect_identical("https://www.ncbi.nlm.nih.gov/Taxonomy/Browser/wwwtax.cgi?id=9606",
+                   uri)
+})

### 2018-12-11 R/ids.R
Unique ids (#20)
* starting to remap ids again

* a slow work in progress

* remaining dbs with syns

also update tests, docs

* data

* testing vignettes

* update pkgdown

@@ -5,8 +5,6 @@
 #' @param name a character vector of species names, e.g. "Homo sapiens"
 #' (Most but not all authorities can also return ids for higher-level
 #'  taxonomic names).
-#' @param synonyms_check Should we check if names are recognized synonyms? Default FALSE.
-#' @param pull logical, should we pull out the id column or return the full table?
 #' @param authority from which authority should the hierachy be returned?
 #'  Default is 'itis'.
 #' @param collect logical, default `TRUE`. Should we return an in-memory
@@ -15,13 +13,17 @@
 #' first perform subsequent filtering operations.)
 #' @param db a connection to the taxald database. See details.
 #' @return a data.frame with columns of `id`, scientific
-#' `name`, and `rank` and a row for each species name queried.
+#' `name`, and `rank`, and `accepted_id` (if data includes synonyms)
+#'  and a row for each species name queried.
 #'
 #' @details
-#' NOTE on matching synonyms: Some authorities (like ITIS) issue separate synonym_ids
-#' corresponding to synonym names. The `ids()` function *does not* return synonym ids.
-#' Rather, if a name is recognized as a synonym, we will look up the appropriate
-#' accepted name and the ID associated with the accepted name and return that.
+#' Some authorities (ITIS, COL, FB/SLB, NCBI) provide synonyms as well.  These
+#' are included in the name list search.  The data.frame returned will then
+#' include an `accepted_id` column, providing the synonym for the requested table.
+#' In this case, the `id` column corresponds to the id for either the name column --
+#' that is, if the name is a synonym, this is the id for the synonym; if the name
+#' is accepted, this id is the same as the accepted id.  NCBI does not issue ids
+#' for known synonyms, so the id is missing for synonym names in this case.
 #'
 #' @export
 #' @importFrom dplyr quo tibble filter right_join
@@ -30,8 +32,6 @@
 ids <- function(name = NULL,
                 authority = c("itis", "ncbi", "col", "tpl",
                               "gbif", "fb", "slb", "wd"),
-                synonyms_check = TRUE,
-                pull = TRUE,           # Should we avoid options that change return type?
                 collect = TRUE,
                 db = td_connect()){
 
@@ -43,46 +43,7 @@ ids <- function(name = NULL,
[…]

### 2018-12-11 data-raw/fishbase.R

@@ -1,33 +1,29 @@
 library(rfishbase) # 3.0
 library(tidyverse)
 
-
-
-
+#### Fishbase
 fb <- as_tibble(rfishbase::load_taxa())
-fb_wide <- fb %>% 
-  select( id = SpecCode, 
-          species = Species, 
-          genus = Genus, 
-          subfamily = Subfamily, 
+fb_wide <- fb %>%
+  select( id = SpecCode,
+          species = Species,
+          genus = Genus,
+          subfamily = Subfamily,
           family = Family,
-          order = Order, 
+          order = Order,
           class = Class,
           superclass = SuperClass) %>%
-  mutate(phylum = "Chorodata", 
+  mutate(phylum = "Chorodata",
          kingdom = "Animalia",
          id = paste0("FB:", id))
 
 write_tsv(fb_wide, "data/fb_hierarchy.tsv.bz2")
 
-fb_taxonid <- fb_wide %>% select(id, species) %>% gather(rank, name, -id)
-write_tsv(fb_taxonid, "data/fb_taxonid.tsv.bz2")
-
+accepted <- fb_wide %>% select(id, species) %>% gather(rank, name, -id)
 
 species <- rfishbase:::fb_species()
 synonym_table <- rfishbase::synonyms(NULL) %>%
-  left_join(species) %>% 
-  rename(id = SpecCode)  %>% 
+  left_join(species) %>%
+  rename(id = SpecCode)  %>%
   mutate(id = paste0("FB:", id),
          SynCode = paste0("FB:", SynCode),
          TaxonLevel = tolower(TaxonLevel)) %>%
@@ -38,7 +34,7 @@ synonym_table <- rfishbase::synonyms(NULL) %>%
          synonym_id = SynCode,
          rank = TaxonLevel)
 
-#find accepted names that are actually useful synonym maps, create table with just synonyms 
+#find accepted names that are actually useful synonym maps, create table with just synonyms
 synonyms <- synonym_table %>%
   filter(type == "accepted name", accepted_name != name) %>%
   select(-type) %>%
@@ -46,25 +42,30 @@ synonyms <- synonym_table %>%
   rename(name = "accepted_name", accepted_name = "name") %>%
   bind_rows(synonym_table %>% filter(type != "accepted name"))
 
+
[…]

### 2018-12-11 data-raw/gbif.R

@@ -11,7 +11,7 @@ library(tidyverse)
 #gbif_taxa <- tbl(gbif_db, "gbif") %>% collect()
 
 
-piggyback::pb_download(repo="cboettig/taxadb")
+#piggyback::pb_download(repo="cboettig/taxadb")
 gbif_taxa <- read_tsv("taxizedb/gbif/gbif.tsv.bz2")
 
 gbif_taxa <- gbif_taxa %>%
@@ -53,14 +53,15 @@ write_tsv(gbif_long, "data/gbif_long.tsv.bz2")
 gbif_taxonid <- gbif_long %>%
   filter(taxonomicStatus == "accepted") %>%
   select(id, name, rank) %>%
-  distinct()
+  distinct() %>%
+  mutate(accepted_id = id) # for consistency, even though we lack synonyms
 write_tsv(gbif_taxonid, "data/gbif_taxonid.tsv.bz2")
 
 # gbif_synonyms <- gbif_long %>%
 #   filter(taxonomicStatus != "accepted") %>%
 #   select(synonym_id = id, accepted_name = name, rank, name = genericName, type = taxonomicStatus) %>%
 #   distinct()# %>%
-#   
+#
 # syn_temp <- left_join(gbif_synonyms, select(gbif_synonyms, id = synonym_id, name), by = c("accepted_name" = "name"))
 
 

### 2018-12-21 README.Rmd
migrate name to taxadb

--- 2018-10-11 README.Rmd
@@
 
-[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
-[![Travis build status](https://travis-ci.org/cboettig/taxald.svg?branch=master)](https://travis-ci.org/cboettig/taxald)
-[![Coverage status](https://codecov.io/gh/cboettig/taxald/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxald?branch=master)
-[![CRAN status](https://www.r-pkg.org/badges/version/taxald)](https://cran.r-project.org/package=taxald)
+[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
+[![Travis build status](https://travis-ci.org/cboettig/taxadb.svg?branch=master)](https://travis-ci.org/cboettig/taxadb)
+[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxadb)
+[![Coverage status](https://codecov.io/gh/cboettig/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxadb?branch=master)
+[![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
 
@@
 
-# taxald
+# taxadb
 
-The goal of `taxald` is to provide fast access to taxonomic data and manipulations, such as resolving taxonomic names to ids, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
+The goal of `taxadb` is to provide fast access to taxonomic data and manipulations, such as resolving taxonomic names to ids, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
 These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records.
@@
 Queries and returned formats also differ across the different taxonomic authorities, making tasks that query multiple authorities particularly complex.
-`taxald` creates a *local* database of most readily available taxonomic authorities, each of which is transformed into consistent, standard, and researcher-friendly tabular formats.
+`taxadb` creates a *local* database of most readily available taxonomic authorities, each of which is transformed into consistent, standard, and researcher-friendly tabular formats.
 
@@
 ```{r eval=FALSE}
-devtools::install_github("cboettig/taxald")
+devtools::install_github("cboettig/taxadb")
 ```
 
-Before we can use most `taxald` functions, we need to do a one-time installation of the database `taxald` uses for almost all commands.
+Before we can use most `taxadb` functions, we need to do a one-time installation of the database `taxadb` uses for almost all commands.
 This can take a while to run, but needs only be done once.
@@
 ```{r }
-library(taxald)
+library(taxadb)
 td_create()
@@
 
-### [An introduction to taxald](https://cboettig.github.io/taxald/articles/articles/taxald.html)
+### [An introduction to taxadb](https://cboettig.github.io/taxadb/articles/articles/taxadb.html)
 
-The [taxald introduction](https://cboettig.github.io/taxald/articles/articles/taxald.html) 
-provides an overview showing how `taxald` functions can help us synthesize data across a given list of species by resolving synonyms and identifiers.
+The [taxadb introduction](https://cboettig.github.io/taxadb/articles/articles/taxadb.html) 
+provides an overview showing how `taxadb` functions can help us synthesize data across a given list of species by resolving synonyms and identifiers.
 
-### [taxald schemas](https://cboettig.github.io/taxald/articles/articles/schema.html)
+### [taxadb schemas](https://cboettig.github.io/taxadb/articles/articles/schema.html)
 
-See the [schemas](https://cboettig.github.io/taxald/articles/articles/schema.html) vignette for an overview of the underlying tables used by `taxald` functions, and more about the different authorities accessed by taxald.
+See the [schemas](https://cboettig.github.io/taxadb/articles/articles/schema.html) vignette for an overview of the underlying tables used by `taxadb` functions, and more about the different authorities accessed by taxadb.
+
+```{r include=FALSE}
+taxadb:::td_disconnect()
+```

### 2018-12-21 R/get_ids.R

@@ -11,14 +11,14 @@
 #'   - `prefix` (e.g. `NCBI:9606`), or
 #'   - `uri` (e.g.
 #'   `http://ncbi.nlm.nih.gov/taxonomy/9606`).
-#' @param taxald_db Connection to from `[td_connect]()`.
+#' @param taxadb_db Connection to from `[td_connect]()`.
 #' @param ... additional arguments passed to `ids()`
 #' @details Note that some taxize authorities: `nbn`, `tropicos`, and `eol`,
-#' are not recognized by taxald and will throw an error here. Meanwhile,
-#' taxald recognizes several authorities not known to `[taxize::get_ids()]`.
+#' are not recognized by taxadb and will throw an error here. Meanwhile,
+#' taxadb recognizes several authorities not known to `[taxize::get_ids()]`.
 #' Both include `itis`, `ncbi`, `col`, and `gbif`.
 #'
-#' Like all taxald functions, this function will run
+#' Like all taxadb functions, this function will run
 #' fastest if a local copy of the authority is installed in advance
 #' using `[td_create()]`.
 #' @examples \donttest{
@@ -31,7 +31,7 @@ get_ids <- function(names,
                     db = c("itis", "ncbi", "col", "tpl",
                            "gbif", "fb", "slb", "wd"),
                     format = c("bare", "prefix", "uri"),
-                    taxald_db = td_connect(),
+                    taxadb_db = td_connect(),
                     ...){
   format <- match.arg(format)
   # be compatible with common space delimiters
@@ -39,7 +39,7 @@ get_ids <- function(names,
   df <- ids(name = names,
              authority = db,
              collect = TRUE,
-             db = taxald_db)
+             db = taxadb_db)
 
   if("accepted_id" %in% names(df))
    out <- df %>% pull("accepted_id")

### 2018-12-21 R/ids.R

@@ -11,7 +11,7 @@
 #' data.frame (default, usually the most convenient), or a reference to
 #' lazy-eval table on disk (useful for very large tables on which we may
 #' first perform subsequent filtering operations.)
-#' @param db a connection to the taxald database. See details.
+#' @param db a connection to the taxadb database. See details.
 #' @return a data.frame with columns of `id`, scientific
 #' `name`, and `rank`, and `accepted_id` (if data includes synonyms)
 #'  and a row for each species name queried.

### 2018-12-21 tests/testthat.R

@@ -1,4 +1,4 @@
 library(testthat)
-library(taxald)
+library(taxadb)
 
-test_check("taxald")
+test_check("taxadb")

### 2018-12-21 tests/testthat/test-connections.R

@@ -19,7 +19,7 @@ test_that("we can use alternative DBs, such as SQLite", {
 
   ## NOTE: SQLite joins are waaaay slower than MonetDBLite
   dbdir <- tempdir()
-  dbname <- file.path(dbdir, "taxald.sqlite")
+  dbname <- file.path(dbdir, "taxadb.sqlite")
 
   db <- DBI::dbConnect(RSQLite::SQLite(), dbname = dbname)
 

### 2018-12-21 tests/testthat/test-fast.R

@@ -1,7 +1,7 @@
 context("fast")
 
 library(testthat)
-library(taxald)
+library(taxadb)
 library(dplyr)
 
 

### 2018-12-21 R/classification.R
migrate name to taxadb (#21)

@@ -6,7 +6,7 @@
 #'  specified as (`Genus species` or `Genus species epithet`)
 #' @param id alternately users can provide a vector of species ids.
 #'  IDs must be prefixed matching the requested authority.  See `id`
-#'  column returned by most `taxald` functions for examples.
+#'  column returned by most `taxadb` functions for examples.
 #' @inheritParams ids
 #' @return a data.frame with one row for each requested species,
 #'  giving the species id, and column for each of the unique-rank

### 2018-12-21 R/td_connect.R

@@ -1,9 +1,9 @@
-#' Connect to the taxald database
+#' Connect to the taxadb database
 #'
 #' @param dbdir Path to the database.
 #' @return Returns a `src_dbi` connection to the database
 #' @details Primarily useful when a lower-level interface to the
-#' database is required.  Most `taxald` functions will connect
+#' database is required.  Most `taxadb` functions will connect
 #' automatically without the user needing to call this function.
 #' @importFrom DBI dbConnect
 #' @importFrom MonetDBLite MonetDBLite
@@ -13,7 +13,7 @@
 #' db <- connect_db()
 #'
 #' }
-td_connect <- function(dbdir = rappdirs::user_data_dir("taxald")){
+td_connect <- function(dbdir = rappdirs::user_data_dir("taxadb")){
 
   dbname <- file.path(dbdir, "monetdblite")
 
@@ -21,7 +21,7 @@ td_connect <- function(dbdir = rappdirs::user_data_dir("taxald")){
   ## CANNOT Just check if this file exists -- that breaks every connection!
   ## Stop if monetdb is pointing to another location from the same session
 
-  db <- mget("td_db", envir = taxald_cache, ifnotfound = NA)[[1]]
+  db <- mget("td_db", envir = taxadb_cache, ifnotfound = NA)[[1]]
   if (inherits(db, "DBIConnection")) {
     if (DBI::dbIsValid(db)) {
       return(db)
@@ -30,20 +30,20 @@ td_connect <- function(dbdir = rappdirs::user_data_dir("taxald")){
 
   dir.create(dbname, FALSE)
   db <- DBI::dbConnect(MonetDBLite::MonetDBLite(), dbname)
-  assign("td_db", db, envir = taxald_cache)
+  assign("td_db", db, envir = taxadb_cache)
 
   db
 }
 
-td_disconnect <- function(env = taxald_cache){
+td_disconnect <- function(env = taxadb_cache){
   db <- mget("td_db", envir = env, ifnotfound = NA)[[1]]
   if (inherits(db, "DBIConnection")) {
     DBI::dbDisconnect(db)
   }
 }
 
-taxald_cache <- new.env()
-reg.finalizer(taxald_cache, td_disconnect, onexit = TRUE)
+taxadb_cache <- new.env()
+reg.finalizer(taxadb_cache, td_disconnect, onexit = TRUE)
 
 
 ## dyplr automatially drops temporary table on disconnect

### 2018-12-21 data-raw/taxizedb-raw.R

@@ -1,41 +1,50 @@
+## Use taxizedb to download all data from original sources and establish in native database formats
+## Then, use arkdb to extract all tables as tsv.bz2 files
 
-## apt -y install mariadb-client postgresql-client
+## This script requires a postgres database and mariadb database be already set up
+## and available using the user name, password, and connection addresses given below.
+
+## Instead of running this code, cached copies of the extracted databases are avialable
+## as a piggyback from the github repo.  Simply run:
+##     piggyback::pb_download(repo = "cboettig/taxadb")
 
 library(arkdb)
-library(taxizedb) # remotes::install_github("ropensci/taxizedb")
+library(taxizedb)
 
 #### ITIS ###########
 itis_store <- db_download_itis()
-db_load_itis(itis_store, user = "postgres", pwd = "password", host = "postgres") # locale issues
+#db_load_itis(itis_store, user = "postgres", pwd = "password", host = "postgres") # locale issues
 itis_db <- src_itis(user = "postgres", password = "password", host = "postgres")
-ark(itis_db, fs::dir_create("taxizedb/itis"), lines = 1e6L, overwrite = TRUE)
+ark(itis_db, fs::dir_create("taxizedb/itis"), streamable_table = streamable_readr_tsv(), lines = 1e5L)
 
 #### TPL ##############
 tpl <- db_download_tpl()
 db_load_tpl(tpl, user = "postgres", pwd = "password", host = "postgres")
 tpl_db <- src_tpl(user = "postgres", password = "password", host = "postgres")
-ark(tpl_db, fs::dir_create("taxizedb/tpl"), lines = 1e6L, overwrite = TRUE)
+ark(tpl_db, fs::dir_create("taxizedb/tpl"), streamable_table = streamable_readr_tsv(), lines = 1e5L)
 
 ##### NCBI ################
 ncbi_store <- db_download_ncbi()
 db_load_ncbi() ## not needed for ncbi
 ncbi_db <- src_ncbi(ncbi_store)
-ark(ncbi_db, fs::dir_create("taxizedb/ncbi"), lines = 1e6L, overwrite = TRUE)
+ark(ncbi_db, fs::dir_create("taxizedb/ncbi"), streamable_table = streamable_readr_tsv(), lines = 1e5L)
 
 #### GBIF ############
 gbif <- db_download_gbif()
 db_load_gbif()## not needed
[…]

### 2018-12-21 tests/testthat/setup-test_db.R

@@ -1,5 +1,5 @@
 
 
-taxald:::td_disconnect()
-test_db <- file.path(tempdir(), "taxald")
+taxadb:::td_disconnect()
+test_db <- file.path(tempdir(), "taxadb")
 dir.create(test_db, showWarnings = FALSE)

### 2018-12-21 tests/testthat/teardown-tests.R

@@ -1,0 +1,1 @@
+taxadb:::td_disconnect()

### 2018-12-21 tests/testthat/test-get_ids.R

@@ -4,14 +4,14 @@ test_that("we can use get_ids options", {
 
   db <- td_connect(test_db)
 
-  bare <- get_ids("Homo sapiens", taxald_db = db)
+  bare <- get_ids("Homo sapiens", taxadb_db = db)
   expect_identical(bare, "180092")
 
   some_ids <- get_ids(c("Homo sapiens", "Mammalia"),
-                      format = "prefix", taxald_db = db)
+                      format = "prefix", taxadb_db = db)
 
   expect_identical(some_ids,  c("ITIS:180092", "ITIS:179913"))
-  uri <- get_ids("Homo sapiens", db= "ncbi", format = "uri", taxald_db = db)
+  uri <- get_ids("Homo sapiens", db= "ncbi", format = "uri", taxadb_db = db)
   expect_identical("http://ncbi.nlm.nih.gov/taxonomy/9606",
                    uri)
 })
@@ -21,7 +21,7 @@ test_that("we can get ids without a DB", {
 
   db <- td_connect(test_db)
 
-  bare <- get_ids("Homo sapiens", taxald_db = NULL)
+  bare <- get_ids("Homo sapiens", taxadb_db = NULL)
   expect_identical(bare, "180092")
 
 })

### 2018-12-28 README.Rmd
update pkgdown
try white-on-green logo
patch ids ordering
also updates pkgdown

@@
 ---
-
-[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
-[![Travis build status](https://travis-ci.org/cboettig/taxadb.svg?branch=master)](https://travis-ci.org/cboettig/taxadb)
-[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxadb)
-[![Coverage status](https://codecov.io/gh/cboettig/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxadb?branch=master)
-[![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
 
@@
 )
+taxadb:::td_disconnect()
+MonetDBLite::monetdblite_shutdown()
 ```
 
-# taxadb
+# taxadb  <img src="man/figures/logo.svg" align="right" alt="" width="120" />
+
+[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
+[![Travis build status](https://travis-ci.org/cboettig/taxadb.svg?branch=master)](https://travis-ci.org/cboettig/taxadb)
+[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxadb)
+[![Coverage status](https://codecov.io/gh/cboettig/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxadb?branch=master)
+[![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
 
@@
 taxadb:::td_disconnect()
+MonetDBLite::monetdblite_shutdown()
 ```

### 2018-12-28 paper/manuscript.Rmd
patch ids ordering
also updates pkgdown

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari Norman"
    affiliation: ucb
  - name: "Jorrit Poelen"
    affiliation: jorritt
  - name: "Scott Chamberlain"
    affiliation: ropensci      
  - name: "Carl Boettiger"
    affiliation: ucb, 1
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: jorrit
    address: "Independent consultant, Oakland, CA, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.
  3) `taxadb` provides a consistent and intuitive data format
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.

journal: "Methods in Ecology & Evolution"
date: "`r Sys.Date()`"
bibliography: refs.bib
output: rticles::elsevier_article
---

The purpose and strengths of `taxadb` can be most easily seen through an introductory example.

```{r}
library(tidyverse)
library(taxadb)
```

```{r}
td_create("gbif")
```

```{r}
url <- "ftp://ftpext.usgs.gov/pub/er/md/laurel/BBS/DataFiles/SpeciesList.txt"
col_names <- c("Seq", "AOU", "English_Common_Name", "French_Common_Name",
               "Spanish_Common_Name", "ORDER", "Family", "Genus", "Species")
bbs <- readr::read_table("bbs.txt", 
                         skip = 9, 
                         col_names = col_names, 
                         col_types = "ccccccccc")

```

```{r}
species_list = paste(bbs$Genus, bbs$Species)
length(species_list)
```

```{r}
bbs_ids <- 
data_frame(species = species_list) %>% 
  mutate(gbif = get_ids(species, "gbif", "prefix"),
         col = get_ids(species, "col", "prefix"),
         itis = get_ids(species, "itis", "prefix"),
         ncbi = get_ids(species, "ncbi", "prefix"),
         wikidata = get_ids(species, "wd", "prefix")) 
```

1. A problem already solved  including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List

2. Performance.  A fundamental challenge of most existing services is the use of internet-based Application Programming Interfaces, (web API)s to resolve individual names.  The motivation for an API-based approach is reasonable: (1) Web APIs are a ubiquitous standard of data exchange (2) make very minimal assumptions about the language, software, or hardware avaialable to the user, (3) users access a single centrally managed database that is maintained and upgraded by the provider, and (4) the provider can easily control access to the data, measure usage statistics, and enforce rate limits.  Many existing tools provide bindings of these APIs as functions to a specific language [@chamberlain2013; @taxize; @itis; @rotl; @worrms; @wikitaxa]

Instead of binding existing web APIs, `taxadb` is built around a set compressed text files following a simple and consistent schema (discussed below). These files are automatically downloaded and imported and stored on a local database by `taxadb`.  By default `taxadb` uses MonetDBLite [@monetdblite], a columnar-oriented relational database requiring no additional installation and providing persistent disk-based storage which can be orders of magnitude faster than traditional database servers such as Postgres [@monetdblite]

3. Consistency

### 2018-12-28 R/get_ids.R

@@ -21,7 +21,7 @@
 #' Like all taxadb functions, this function will run
 #' fastest if a local copy of the authority is installed in advance
 #' using `[td_create()]`.
-#' @examples \donttest{
+#' @examples \dontrun{
 #' get_ids("Homo sapiens")
 #' get_ids(c("Homo sapiens", "Mammalia"), format = "prefix")
 #' get_ids("Homo sapiens", db= "ncbi", format = "uri")

### 2018-12-28 R/ids.R

@@ -34,16 +34,19 @@ ids <- function(name = NULL,
                               "gbif", "fb", "slb", "wd"),
                 collect = TRUE,
                 db = td_connect()){
-
-  input_table <- dplyr::tibble(name)
+  sort <- TRUE # dummy name
+  input_table <- dplyr::tibble(name, sort = 1:length(name))
 
   ## Use right_join, so unmatched names are kept, with NA
-  ## Using right join, names appear in order of authority
-  out <- dplyr::right_join(
+  ## Using right join, names appear in order of authority!
+  out <-
+    dplyr::right_join(
       taxa_tbl(authority, "taxonid", db),
       input_table,
       by = "name",
-      copy = TRUE)
+      copy = TRUE) %>%
+    dplyr::arrange(sort) %>%
+    select(-sort)
 
   if (collect && inherits(out, "tbl_lazy")) {
     ## Return an in-memory object

### 2018-12-28 R/td_create.R

@@ -32,8 +32,7 @@
 #' @importFrom arkdb unark streamable_readr_tsv
 #' @importFrom MonetDBLite MonetDBLite
 #' @importFrom readr cols
-#' @examples
-#' \dontrun{
+#' @examples \dontrun{
 #'   # tmp <- tempdir()
 #'   # create_db(authorities = "itis", dbdir = tmp)
 #'

### 2018-12-28 paper/manuscript.Rmd
various exploration:
looking a BBS data example for manuscript.
need to deal with fuzzy matching

@@
 
+```{r}
+## Consider: Working from a focal authority:
+## select all rows where focal authority did not match but another authority did
+## for each matched authority, access species name associated with id, and requery
+
+## OR consider: access accepted species name associated with any authority, requery that in original authority?
+##
+
+## OR consider? generate a combined taxonid table for all authorities
+```
+
+```{r}
+bbs_ids %>% 
+  select(-species) %>% 
+  purrr::map_dbl(function(x) sum(!is.na(x)))
+```
+
+ITIS resolves the most names, but no authority resolves all the names provided.
+
+```{r}
+unmatched <- species_list[is.na(bbs_ids$itis)]
+
+first_two <- function(x){ tmp <- strsplit(x, " ")[[1]]; paste(tmp[[1]], tmp[[2]]) }
+drop_sp <- function(x) gsub("(\\w+)\\s+sp\\.", "\\1", x)
+
+matched <- unmatched %>% 
+  map_chr(first_two) %>% 
+  drop_sp() %>%
+  get_ids("itis", "prefix")
+
+matched %>% is.na() %>% sum()
+
+still_umatched <- unmatched[is.na(matched)]
+
+```
+
+```{r}
+first_two_fuzzy <- function(x){ 
+  tmp <- strsplit(x, " ")[[1]]
+  paste0(tmp[[1]], " ", tmp[[2]], "%") 
+}
+
+only_genus <- function(x){ 
+  purrr::map_chr(x, function(x) strsplit(x, " ")[[1]][[1]])
+}
+name_to_regex <- function(x){ 
+  x <- gsub("(\\w+)\\s+sp\\..*", "\\1.*", x)
+  # x <- gsub("(\\w+)\\s+(\\w+).*", "\\1 \\2.*", x)
+  x <- gsub("(\\w+)\\s+(\\w{1,3}).*", "\\1 \\2.*", x)
+  x
+}
+name_to_regex(still_umatched)
+
+## subset by genus first
+tmp <- 
+  right_join(taxa_tbl("itis", "hierarchy"), 
+           data_frame(genus = only_genus(still_umatched)),
+           copy = TRUE) %>% select(id) %>% 
+  left_join(taxa_tbl("itis", "taxonid"), by="id") %>%
+  distinct() %>% 
+  collect()
+
+df <- data_frame(regex_name = name_to_regex(still_umatched))
+i <- vapply(df$regex_name, 
+            function(x) which(grepl(x, tmp$name)),
+            integer(1))
+tmp[i,] %>%  select(id, name) %>% distinct()
+
+tmp[which(grepl("Rallus.*", tmp$name)), ] %>%  select(id, name) %>% distinct()
+
+```
+
+```{r}
+matched <- tmp %>%   
+  fuzzyjoin::regex_inner_join(
+    data.frame(regex_name = name_to_regex(still_umatched)),
+    by = c(name = "regex_name")
+  )
+```
+
+```{r}
+matches <- 
+  right_join(taxa_tbl("itis", "hierarchy"), 
+           data_frame(genus = only_genus(species_list)),
+           copy = TRUE) %>% 
+  distinct() %>% 
+  collect() 
+# uses smaller table filtered by genus
+library(fuzzyjoin)
+left_join(data_frame(species = species_list), matches)
+
+stringdist_left_join(data_frame(species = species_list), matches)
+
+  right_join(data_frame(species = first_two(unmatched))) %>% 
+  select(id, species) %>% 
+  distinct()
+
+taxa_tbl("itis", "taxonid") %>%
+  filter(name %in% c("Rallus%", "Vermivora%"), rank == "species")
+
+taxa_tbl("itis", "hierarchy") %>% filter(genus == "Rallus") %>% select(genus, species)
+```
+
+```{r}
+itis <- ids(species_list, "itis") 
+col <- accepted_name(bbs_ids$itis, "itis") %>% pull(name) %>% get_ids("col")
+sum(!is.na(col))
+```
+
+```{r}
+unmatched <- bbs_ids %>%  
+  select(-species) %>%
+  purrr::map_dfc(function(x) !is.na(x)) %>% 
+  rowSums() < 1
+
+sum(!unmatched)
+```
+
 1. A problem already solved  including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List

### 2018-12-28 R/ids.R

@@ -58,3 +58,34 @@ ids <- function(name = NULL,
 
 
 
+
+
+## FIXME abstract this to filter on id / name / generic column?
+accepted_name <- function(id = NULL,
+                authority = c("itis", "ncbi", "col", "tpl",
+                              "gbif", "fb", "slb", "wd"),
+                collect = TRUE,
+                db = td_connect()){
+  sort <- TRUE # dummy name
+  input_table <- dplyr::tibble(id, sort = 1:length(id))
+
+  ## Use right_join, so unmatched names are kept, with NA
+  ## Means names appear in order of authority, so we must arrange
+  ## after-the-fact to match the query order
+  out <-
+    dplyr::right_join(
+      taxa_tbl(authority, "taxonid", db),
+      input_table,
+      by = "id",
+      copy = TRUE) %>%
+    dplyr::arrange(sort) %>%
+    select(-sort)
+
+  if (collect && inherits(out, "tbl_lazy")) {
+    ## Return an in-memory object
+    return( dplyr::collect(out) )
+  }
+
+  out
+}
+

### 2018-12-28 data-raw/ott.R

@@ -1,0 +1,1 @@
+download.file("http://files.opentreeoflife.org/ott/ott3.0/ott3.0.tgz", "ott3.0.tgz")

### 2018-12-30 paper/manuscript.Rmd
support ott

@@
          wikidata = get_ids(species, "wd", "prefix"))
+
+```
+
+```{r}
+ott_ids <- ids(species_list, "ott")
 ```

### 2018-12-30 R/taxa_tbl.R

@@ -1,4 +1,5 @@
-
+KNOWN_AUTHORITIES <- c("itis", "ncbi", "col", "tpl",
+                       "gbif", "fb", "slb", "wd", "ott")
 #' Return a reference to a given table in the taxadb database
 #'
 #' @param db a connection to the taxadb database. Default will
@@ -8,8 +9,7 @@
 #' @inheritParams classification
 #' @export
 taxa_tbl <- function(
-  authority = c("itis", "ncbi", "col", "tpl",
-                "gbif", "fb", "slb", "wd"),
+  authority = KNOWN_AUTHORITIES,
   schema = c("hierarchy", "taxonid", "synonyms", "common", "long"),
   db = td_connect()){
 

### 2018-12-30 R/td_create.R

@@ -17,14 +17,15 @@
 #' fast [MonetDBLite::MonetDBLite]() connection.
 #' @details
 #'  Authorities recognized by taxadb are:
-#'  - `itis`
-#'  - `ncbi`
-#'  - `col`
-#'  - `tpl`
-#'  - `gbif`
+#'  - `itis`: Integrated Taxonomic Information System, <https://www.itis.gov/>
+#'  - `ncbi`:  National Center for Biotechnology Information, <https://www.ncbi.nlm.nih.gov/taxonomy>
+#'  - `col`: Catalogue of Life, <http://www.catalogueoflife.org/>
+#'  - `tpl`: The Plant List, <http://www.theplantlist.org/>
+#'  - `gbif`: Global Biodiversity Information Facility, <https://www.gbif.org/>
 #'  - `fb` FishBase, <http://fishbase.org>
 #'  - `slb`, SeaLifeBase, <http://sealifebase.org>
-#'  - `wd`, wikidata
+#'  - `wd`, Wikidata: <https://www.wikidata.org/>
+#'  - `ott` OpenTree Taxonomy: <https://github.com/OpenTreeOfLife/reference-taxonomy>
 #' @return path where database has been installed (invisibly)
 #' @export
 #' @importFrom utils download.file
@@ -45,14 +46,7 @@ td_create <- function(authorities = "itis",
                       db = td_connect(dbdir)
                       ){
 
-  recognized_authorities = c("itis",
-                             "ncbi",
-                             "col",
-                             "tpl",
-                             "gbif",
-                             "fb",
-                             "slb",
-                             "wd")
+  recognized_authorities = KNOWN_AUTHORITIES
   recognized_schema = c("hierarchy", "taxonid", "synonyms")
   if (authorities == "all") {
     authorities <- recognized_authorities

### 2018-12-30 data-raw/redlist.R

@@ -1,7 +1,7 @@
 library(tidyverse)
 library(httr)
 links <- paste0("http://apiv3.iucnredlist.org/api/v3/species/page/",
-                0:9,
+                0:10,
                 "?token=9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee")
 system.time({
 
@@ -16,8 +16,24 @@ system.time({
     })
 })
 
+
+sentence_case <- function(x) {
+  Hmisc::capitalize(str_to_lower(x))
+  #gsub("(. )([A-Z])(.+)", "\\1\\U\\2\\L\\3", x)
+}
+hierarchy <- full %>% select(id = taxonid, kingdom = kingdom_name, phylum = phylum_name,
+                class = class_name, order = order_name, family = family_name,
+                genus = genus_name, species = scientific_name) %>%
+  mutate_if(is.character, sentence_case) %>%
+  mutate(id = paste0("IUCN:", id))
+
+dir.create("data", FALSE)
+write_tsv(hierarchy, "data/iucn_hierarchy.tsv.bz2")
+
+#, category   infra_rank    infra_name      population   )
+
 ## Or just read in result to save time
-full <- read_csv("https://espm-157.github.io/extinction-module/all_species.csv")
+full2 <- read_csv("https://espm-157.github.io/extinction-module/all_species.csv")
 
 #install.packages('rredlist')
 library(rredlist)
@@ -29,3 +45,10 @@ synonyms <- full$scientific_name %>%
       return$result
     })
 
+
+
+common <- full$scientific_name %>%
+  map_dfr(function(name){
+    return <- rredlist::rl_common_names(name, key = key)
+    return$result
+  })

### 2019-01-02 paper/manuscript.Rmd

@@
                "Spanish_Common_Name", "ORDER", "Family", "Genus", "Species")
-bbs <- readr::read_table("bbs.txt", 
+bbs <- readr::read_table(url, 
                          skip = 9, 

### 2019-01-04 notebook/fuzzy-matching.Rmd
notebook

---
title: "fuzzy-matching.Rmd"
author: "Carl Boettiger"
date: "1/3/2019"
output: html_document
---

```{r}
library(tidyverse)
library(taxadb)
```

```{r}
td_create("itis")
```

```{r}
url <- "ftp://ftpext.usgs.gov/pub/er/md/laurel/BBS/DataFiles/SpeciesList.txt"
col_names <- c("Seq", "AOU", "English_Common_Name", "French_Common_Name",
               "Spanish_Common_Name", "order", "family", "genus", "species")
bbs <- readr::read_table(url, 
                         skip = 9, 
                         col_names = col_names, 
                         col_types = "ccccccccc")
bbs <- bbs %>% mutate(species = paste(genus, species))
#readr::write_tsv(bbs, "inst/extdata/bbs.tsv")

```

```{r}
name <- bbs$species
authority <- "itis"
db <- td_connect()
match <- "contains"
```

# Strategy 1: SQL-based fuzzy match

```{r}
name_pattern <- switch(match,
                         starts_with = paste0(name, "%"),
                         contains =  paste0("%", name, "%")
  )

  system.time({
  out <- purrr::map_dfr(name_pattern,
          function(pattern)
            taxa_tbl(authority, "taxonid", db) %>%
            filter(name %like% pattern) %>% collect()
         )
  })
out
```

# Strategy 2: Extract a smaller table

```{r}
  ## Strategy: extract all potential matches by Genus alone.  assumes first name is a genus name!
  only_genus <- function(name)  stringi::stri_extract_first_words(name)
  id_tbl <- ids(only_genus(name), authority = authority, db = db, collect = FALSE) %>%
    select(name) %>%
    inner_join(select(taxa_tbl(authority, "hierarchy"), id, genus), by = c(name = "genus")) %>%
    select(id) %>% inner_join(taxa_tbl(authority, "taxonid"), by = "id") %>%
    distinct() %>%
    collect()

  name_regex <- switch(match,
                         starts_with = paste0(name, ".*"),
                         contains =  paste0(".*", name, ".*")
  )

  id_tbl <- collect(taxa_tbl(authority, "taxonid"))

  ## Using the genus subset -- a much smaller list of matches -- is this good or bad?
  system.time({
    out2 <- purrr::map_dfr(name_regex, function(pattern)
      filter(id_tbl, grepl(pattern, name))
    )
  })

```

```{r}   
  ## In memory, even slower!! 
  system.time({
    id_tbl <-  collect(taxa_tbl(authority, "taxonid"))
    out2 <- purrr::map_dfr(name_regex, function(pattern)
      filter(id_tbl, grepl(pattern, name))
    )
  })

```

### 2019-01-04 notebook/fuzzy-matching.Rmd
WIP on fuzzy matching

@@
 ```{r}
-url <- "ftp://ftpext.usgs.gov/pub/er/md/laurel/BBS/DataFiles/SpeciesList.txt"
-col_names <- c("Seq", "AOU", "English_Common_Name", "French_Common_Name",
-               "Spanish_Common_Name", "order", "family", "genus", "species")
-bbs <- readr::read_table(url, 
-                         skip = 9, 
-                         col_names = col_names, 
-                         col_types = "ccccccccc")
-bbs <- bbs %>% mutate(species = paste(genus, species))
-#readr::write_tsv(bbs, "inst/extdata/bbs.tsv")
+read_tsv(system.file("extdata/bbs.tsv"))
 

### 2019-01-04 paper/manuscript.Rmd
WIP on fuzzy matching

@@
 ```{r}
-td_create("gbif")
+td_create("all")
 ```
@@
 col_names <- c("Seq", "AOU", "English_Common_Name", "French_Common_Name",
-               "Spanish_Common_Name", "ORDER", "Family", "Genus", "Species")
+               "Spanish_Common_Name", "order", "family", "genus", "species")
 bbs <- readr::read_table(url, 
@@
                          col_types = "ccccccccc")
+bbs <- bbs %>% mutate(species = paste(genus, species))
+readr::write_tsv(bbs, "inst/extdata/bbs.tsv")
 
@@
 ```{r}
-species_list = paste(bbs$Genus, bbs$Species)
+species_list = paste(bbs$genus, bbs$species)
 length(species_list)
@@
          ncbi = get_ids(species, "ncbi", "prefix"),
-         wikidata = get_ids(species, "wd", "prefix"))
+     wikidata = get_ids(species, "wd", "prefix"),
+         #iucn = get_ids(species, "iucn", "prefix"),
+          ott = get_ids(species, "ott","prefix") 
+  )
 
+```
+
+```{r}
+bbs_ids %>% 
+  select(-species) %>% 
+  purrr::map_dbl(function(x) sum(!is.na(x)))
+```
+
+```{r}
 ```
@@
 ## OR consider? generate a combined taxonid table for all authorities
-```
-
-```{r}
-bbs_ids %>% 
-  select(-species) %>% 
-  purrr::map_dbl(function(x) sum(!is.na(x)))
 ```

### 2019-01-04 R/get_ids.R

@@ -31,8 +31,8 @@
 #' @importFrom tibble column_to_rownames
 get_ids <- function(names,
                     db = KNOWN_AUTHORITIES,
-                    taxadb_db = td_connect(),
                     format = c("bare", "prefix", "uri"),
+                    taxadb_db = td_connect(),
                     ...){
   format <- match.arg(format)
   # be compatible with common space delimiters
@@ -44,7 +44,7 @@ get_ids <- function(names,
 
   ## More aggressive logic to enforce
   ## we get precisely one ID or NA for each name queried
-  df <- tibble::column_to_rownames(df, "name")
+  # df <- tibble::column_to_rownames(df, "name")
   if("accepted_id" %in% names(df)){
     #out <- df[names, "accepted_id"]
     out <- pull(df, "accepted_id")

### 2019-01-07 paper/manuscript.Rmd

--- 2019-01-02 paper/manuscript.Rmd
@@
 ```{r}
-url <- "ftp://ftpext.usgs.gov/pub/er/md/laurel/BBS/DataFiles/SpeciesList.txt"
-col_names <- c("Seq", "AOU", "English_Common_Name", "French_Common_Name",
-               "Spanish_Common_Name", "ORDER", "Family", "Genus", "Species")
-bbs <- readr::read_table(url, 
-                         skip = 9, 
-                         col_names = col_names, 
-                         col_types = "ccccccccc")
+bbs <- read_tsv(system.file("extdata/bbs.tsv", package="taxadb"))
+```
+
+```{r}
+dim(bbs)
+```
+
+```{r}
+bbs_ids <- 
+bbs %>% select(species) %>%
+  mutate(gbif = get_ids(species, "gbif", "prefix"),
+          col = get_ids(species, "col", "prefix"),
+         itis = get_ids(species, "itis", "prefix"),
+         ncbi = get_ids(species, "ncbi", "prefix"),
+     wikidata = get_ids(species, "wd", "prefix"),
+          ott = get_ids(species, "ott","prefix") 
+  )
 
@@
 ```{r}
-species_list = paste(bbs$Genus, bbs$Species)
-length(species_list)
+bbs_ids %>% 
+  select(-species) %>% 
+  purrr::map_dbl(function(x) sum(!is.na(x)))
 ```
@@
 ```{r}
-bbs_ids <- 
-data_frame(species = species_list) %>% 
+bbs_ids <- bbs %>% 
+  select(species) %>% 
+  mutate(species = clean_names(species)) %>%
   mutate(gbif = get_ids(species, "gbif", "prefix"),
@@
          ncbi = get_ids(species, "ncbi", "prefix"),
-         wikidata = get_ids(species, "wd", "prefix"))
+     wikidata = get_ids(species, "wd", "prefix"),
+          ott = get_ids(species, "ott","prefix") 
+  )
 
+bbs_ids %>% 
+  select(-species) %>% 
+  purrr::map_dbl(function(x) sum(!is.na(x)))
+```
+
+```{r}
+ids <- get_ids(clean_names(bbs$species), "ott", "prefix")
+unmatched <- bbs$species[is.na(ids)]
+clean_names(unmatched)
+```
+
+```{r}
+unmatched <- bbs$species[is.na(bbs_ids$ott)]
+length(unmatched)
+
+drop_sp.(unmatched) %>%
+  binomial_names()
+
+clean_names(unmatched)
+```
+
+```{r}
+bbs_ids %>% 
+  select(-species) %>% 
+  purrr::transpose() %>%
+  purrr::map_dbl(function(x) sum(!is.na(x)) < 1) %>% sum()
 ```

### 2019-01-09 paper/manuscript.Rmd
fuzzy name matching, name cleaning (WIP)

@@
 
+`get_ids` by default looks only for exact matches.  We can get more matches by cleaning the names first.  `clean_names()`
+
 ```{r}
-bbs_ids <- bbs %>% 
-  select(species) %>% 
-  mutate(species = clean_names(species)) %>%
-  mutate(gbif = get_ids(species, "gbif", "prefix"),
-          col = get_ids(species, "col", "prefix"),
-         itis = get_ids(species, "itis", "prefix"),
-         ncbi = get_ids(species, "ncbi", "prefix"),
-     wikidata = get_ids(species, "wd", "prefix"),
-          ott = get_ids(species, "ott","prefix") 
+clean_names(bbs$species, 
             )
-
-bbs_ids %>% 
-  select(-species) %>% 
-  purrr::map_dbl(function(x) sum(!is.na(x)))
 ```

### 2019-01-09 R/fuzzy.R

@@ -27,7 +27,35 @@ search <- function(name = NULL,
 }
 
 
-#
+#' Clean taxonomic names
+#'
+#' A utility to sanitize taxonomic names to increase probability of resolving names.
+#'
+#' @param names a character vector of taxonomic names (usually species names)
+#' @param fix_delim Should we replace separators `.`, `_`, `-`
+#' with spaces? e.g. 'Homo.sapiens' becomes 'Homo sapiens'.
+#' logical, default TRUE.
+#' @param binomial_only Attempt to prune name to a binomial name, e.g.
+#'  Genus and species (specific epithet), e.g. `Homo sapiens sapiens`
+#'  becomes `Homo sapiens`. logical, default TRUE.
+#' @param remove_sp Should we drop unspecified species epithet designations?
+#' e.g. `Homo sp.` becomes `Homo` (thus only matching against genus level ids).
+#' logical, default TRUE.
+#' @details Current implementation is limited to handling a few common cases.
+#' Additional extensions may be added later. A goal of the `clean_names` function
+#' is that any modification rule of the name strings be precise, atomic, and
+#' toggle-able, rather than relying on clever but more opaque rules and
+#' arbitrary scores. This utility should always be used with care, as
+#' indiscriminant modification of names may result in successful but inaccurate
+#' name matching. A good pattern is to only apply this function to the subset
+#' of names that cannot be directly matched.
+#'
+#'
+#' @importFrom stringi stri_replace_all_regex stri_extract_all_words
+#' @importFrom stringi stri_trim stri_split_regex
+#' @export
+#' @examples
+#' clean_names(c("Homo sapiens sapiens", "Homo.sapiens", "Homo sp."))
 clean_names <-
   function(names,
            fix_delim = TRUE,
@@ -44,8 +72,7 @@ clean_names <-
 }
 
 ## Name cleaning utilities
-#' @importFrom stringi stri_replace_all_regex stri_extract_all_words
-#' @importFrom stringi stri_trim stri_split_regex
+
 set_space_delim <- function(x)
   stringi::stri_replace_all_regex(x, "(_|-|\\.)", " ") %>%
   stringi::stri_trim()
@@ -59,7 +86,9 @@ binomial_names <- function(x){
     stringi::stri_extract_all_words(simplify = TRUE)
   stringi::stri_trim(paste(s[,1], s[,2]))
 }
-
+drop_author_year <- function(x){
[…]

### 2019-01-10 paper/manuscript.Rmd
wip

@@
      wikidata = get_ids(species, "wd", "prefix"),
+  #       iucn = get_ids(species, "iucn", "prefix"),
           ott = get_ids(species, "ott","prefix") 
@@
 
-`get_ids` by default looks only for exact matches.  We can get more matches by cleaning the names first.  `clean_names()`
+`get_ids` by default looks only for exact matches.
+We can get more matches by cleaning the names first.
+The `clean_names()` function provides several simple string manipulations to tidy names to improve the probability of a match: missing species (specific eptithets) such as `Accipiter sp.` drop the `sp.`, allowing matches against Genus names, and intraspecific epithets such as `Colaptes auratus cafer` are dropped to binomial names, `Colaptes auratus`.
+These transformations may not be appropriate for certain use cases, and should be used with care, and only once as many untransformed names as possible have already been matched.
+
+In this example, applying this transformation on the unmatched names we can bring matches against OTT identifiers up from 665 to 740, missing only ten names:
 
 ```{r}
-clean_names(bbs$species, 
-            )
+bbs_ott <- bbs$species %>% get_ids("ott", "prefix")
 ```
@@
 ```{r}
-ids <- get_ids(clean_names(bbs$species), "ott", "prefix")
-unmatched <- bbs$species[is.na(ids)]
-clean_names(unmatched)
+cleaned <- 
+  bbs$species[is.na(bbs_ott)] %>% 
+  clean_names() %>% 
+  get_ids("ott", "prefix")
+
+## use ID from cleaned name only if no exact match.
+bbs_ott[is.na(bbs_ott)] <- cleaned
 ```
 
+We are still missing 10 species:
+
 ```{r}
-unmatched <- bbs$species[is.na(bbs_ids$ott)]
-length(unmatched)
-
-drop_sp.(unmatched) %>%
-  binomial_names()
-
-clean_names(unmatched)
+still_missing <- bbs$species[is.na(bbs_ott)]
+still_missing
 ```
 
+Three of these are recognizably common names, which we can resolve against the common names table:
+
 ```{r}
-bbs_ids %>% 
-  select(-species) %>% 
-  purrr::transpose() %>%
-  purrr::map_dbl(function(x) sum(!is.na(x)) < 1) %>% sum()
+clean_names(still_missing) %>% ids("gbif")
 ```
 
-```{r}
-ott_ids <- ids(species_list, "ott")
-```
-
-```{r}
-## Consider: Working from a focal authority:
-## select all rows where focal authority did not match but another authority did
-## for each matched authority, access species name associated with id, and requery
-
-## OR consider: access accepted species name associated with any authority, requery that in original authority?
-##
-
-## OR consider? generate a combined taxonid table for all authorities
-```
-
-ITIS resolves the most names, but no authority resolves all the names provided.
-
-```{r}
-unmatched <- species_list[is.na(bbs_ids$itis)]
-
-first_two <- function(x){ tmp <- strsplit(x, " ")[[1]]; paste(tmp[[1]], tmp[[2]]) }
-drop_sp <- function(x) gsub("(\\w+)\\s+sp\\.", "\\1", x)
-
-matched <- unmatched %>% 
-  map_chr(first_two) %>% 
-  drop_sp() %>%
-  get_ids("itis", "prefix")
-
-matched %>% is.na() %>% sum()
-
-still_umatched <- unmatched[is.na(matched)]
-
-```
-
-```{r}
-first_two_fuzzy <- function(x){ 
-  tmp <- strsplit(x, " ")[[1]]
-  paste0(tmp[[1]], " ", tmp[[2]], "%") 
-}
-
-only_genus <- function(x){ 
-  purrr::map_chr(x, function(x) strsplit(x, " ")[[1]][[1]])
-}
-name_to_regex <- function(x){ 
-  x <- gsub("(\\w+)\\s+sp\\..*", "\\1.*", x)
-  # x <- gsub("(\\w+)\\s+(\\w+).*", "\\1 \\2.*", x)
-  x <- gsub("(\\w+)\\s+(\\w{1,3}).*", "\\1 \\2.*", x)
-  x
-}
-name_to_regex(still_umatched)
-
-## subset by genus first
-tmp <- 
-  right_join(taxa_tbl("itis", "hierarchy"), 
-           data_frame(genus = only_genus(still_umatched)),
-           copy = TRUE) %>% select(id) %>% 
-  left_join(taxa_tbl("itis", "taxonid"), by="id") %>%
-  distinct() %>% 
-  collect()
-
-df <- data_frame(regex_name = name_to_regex(still_umatched))
-i <- vapply(df$regex_name, 
-            function(x) which(grepl(x, tmp$name)),
-            integer(1))
-tmp[i,] %>%  select(id, name) %>% distinct()
-
-tmp[which(grepl("Rallus.*", tmp$name)), ] %>%  select(id, name) %>% distinct()
-
-```
-
-```{r}
-matched <- tmp %>%   
-  fuzzyjoin::regex_inner_join(
-    data.frame(regex_name = name_to_regex(still_umatched)),
-    by = c(name = "regex_name")
-  )
-```
-
-```{r}
-matches <- 
-  right_join(taxa_tbl("itis", "hierarchy"), 
-           data_frame(genus = only_genus(species_list)),
-           copy = TRUE) %>% 
-  distinct() %>% 
-  collect() 
-# uses smaller table filtered by genus
-library(fuzzyjoin)
-left_join(data_frame(species = species_list), matches)
-
-stringdist_left_join(data_frame(species = species_list), matches)
-
-  right_join(data_frame(species = first_two(unmatched))) %>% 
-  select(id, species) %>% 
-  distinct()
-
-taxa_tbl("itis", "taxonid") %>%
-  filter(name %in% c("Rallus%", "Vermivora%"), rank == "species")
-
-taxa_tbl("itis", "hierarchy") %>% filter(genus == "Rallus") %>% select(genus, species)
-```
-
-```{r}
-itis <- ids(species_list, "itis") 
-col <- accepted_name(bbs_ids$itis, "itis") %>% pull(name) %>% get_ids("col")
-sum(!is.na(col))
-```
-
-```{r}
-unmatched <- bbs_ids %>%  
-  select(-species) %>%
-  purrr::map_dfc(function(x) !is.na(x)) %>% 
-  rowSums() < 1
-
-sum(!unmatched)
-```
+(GBIF lists *Antigone canadensis* as a synonym for *Grus canadensis*; this appears backwards as the Sandhill Crane was moved out of *Grus* after molecular analyses showed it made that taxon polyphyletic; https://en.wikipedia.org/wiki/Sandhill_crane).
 

### 2019-01-14 drafts/crosswalk.Rmd
update docs

---
title: "Crosswalking The Authorities"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{crosswalk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

taxadb:::td_disconnect()
MonetDBLite::monetdblite_shutdown()

```

```{r message = FALSE}
library(taxadb)
library(dplyr)
library(tidyr)
```

Work in progress -- the mismatches of `itis`, `col` and `wd` here illustrate the complexity of name matching.  Some names resolve multiple times.  Some authorities (ITIS) recognize certain synonyms that they still haven't mapped to accepted id.

```{r}
fish <- taxa_tbl("fb") %>%  
  select(fb = id, species) %>% 
  collect() 
species <- pull(fish, species)

itis <- ids(species, "itis")
wd <- ids(species, "wd")
col <- ids(species, "col")

## Note each of these authorities all return more rows than we had in the input table!
sapply(list(fb = fish, itis = itis, wd = wd, col = col), function(x) length(x[[1]]))

```

```{r}
my_taxa <- fish %>%

  mutate(#itis = get_ids(species, "itis"),
         ncbi = get_ids(species, "ncbi"),
         #col = get_ids(species, "col"),
         gbif = get_ids(species, "gbif"),
         #wd = get_ids(species, "wd"),
         tpl = get_ids(species, "tpl"),
        # fb = get_ids(species, "fb"),
         slb = get_ids(species, "slb")) 
```

```{r}
my_taxa %>% 
  select(-species) %>% 
  purrr::map_dbl(function(x) sum(!is.na(x)))
```

Looks like three plants have matching scientific names to some of our fish:

```{r}
dup <- fish %>% pull(species) %>% ids(authority = "tpl") %>% filter(!is.na(id))
dup
```

This also probably explains why `col` and `wd` are returning the wrong-length matches:

```{r}
species <- pull(fish, species)

col_fb <- ids(species, "col") 
dim(col_fb)[1] - length(species)

wd_fb <- ids(species, "wd") 
dim(wd_fb)[1] - length(species)

```

```{r}

#col_hierarchy <- classification(id = col_fb$id) %>% filter(kingdom == "Animalia")

taxa_tbl("col", "hierarchy") %>%
  filter(kingdom == "Animalia") %>%
  semi_join(select(col_fb, id), copy = TRUE) %>%
  select(id, species) %>% 
  collect()
```

```{r}
has_match <- my_taxa %>% 
  select(-species, -fb) %>%
  purrr::map_dfc(function(x) !is.na(x)) %>% 
  rowSums() > 0

my_taxa %>% filter(!has_match)
```

```{r include=FALSE}
taxadb:::td_disconnect()
```

### 2019-02-12 notebook/algae-explore.Rmd
starting to poke around algae names

```{r}
library(tidyverse)
library(taxadb)
```

Populate the `taxadb` database (only required right after installing `taxadb`, otherwise this will just note the previously installed tables).

```{r}
td_create("all")
```

I'm using names given for Red, Green, and Brown Algae from [Guiry (2012)](https://doi.org/10.1111/j.1529-8817.2012.01222.x):

Open Tree Taxonomy (OTT) has 32,347 names recognized as belonging to one the three phyla:

```{r}
ott_phyla <- bind_rows(
  descendants(name = "Cyanobacteria", rank = "phylum", authority = "ott"),
  descendants(name = "Rhodophyta", rank = "phylum", authority = "ott"),
  descendants(name = "Phaeophyceae", rank = "phylum", authority = "ott")
)

ott_phyla
```

GBIF has 7,412 recognized names

```{r}
gbif_phyla <- bind_rows(
  descendants(name = "Cyanobacteria", rank = "phylum", authority = "gbif"),
  descendants(name = "Rhodophyta", rank = "phylum", authority = "gbif"),
  descendants(name = "Phaeophyceae", rank = "phylum", authority = "gbif")
)

gbif_phyla
```

How many GBIF names also match the names given in OTT? Looks like only 2,923 exact matches.

```{r}
gbif_in_ott <- gbif_phyla %>% 
  select(gbif_id = taxonID, scientificName, taxonRank) %>%
  inner_join(ott_phyla)
```

### 2019-02-19 notebook/algae.Rmd
keep sort column; work on algae

Load libraries and data to get started

```{r}
library(tidyverse)
library(taxadb)
```

```{r}
taxadb::td_create() # only needed once.
algae <- read_csv("~/projects/algae-names/algae_uncleanednames_NAdropped.csv")
```

Before we begin, let's take a look around the data to get a sense of what we have:

- `r dim(algae)[[1]]` rows, `r names(algae)` columns.

```{r}
dim(algae)[[1]] # 57,700 rows. 
names(algae) # cols are: source, family, species, is_source_herbaria
```

```{r}
algae %>% count(family) %>% arrange(desc(n))  # 1,126 families (uncleaned)
```

```{r}
# 25 sources, including MACROALGAE, iDigBio, GBIF, OBIS, ...
algae %>% count(source) %>% arrange(desc(n))
```

lots of duplicate names across & sources: only 37,505 unique species.

```{r}
algae %>% count(species) %>% arrange(desc(n))
```

However, note that some of these duplicated names have different families though:

```{r}
unique_sp_family <- algae %>% count(species, family) %>% arrange(desc(n))
unique_sp_family  # 42,242 rows
```

We can see this is sometimes due to missing data or differences in capitalization, but also see cases where the same species name is assigned to different families:

```{r}
species_multiple_families <- 
  unique_sp_family %>% 
  filter(species != "sp.", species != "Indet. sp.") %>%
  select(species) %>% 
  count(species) %>% 
  arrange(desc(n)) %>%
  left_join(unique_sp_family %>% select(-n))
species_multiple_families %>%  arrange(desc(n))
```

We will assume these differences correspond to changes in taxonomic group assignment at the family level, and not to two distinct species with the same scientific name (genus+specific epithet) belonging to different families.

So we'll begin by focusing on resolving the unique species names:

```{r}
names <- unique(algae$species) # 37,505
```

Okay, without futher ado, let's start matching names.  We'll resolve against the Open Tree Taxonomy (OTT) first, because it is a assembly that already includes names from WORMS, GBIF, and others.

```{r}
ott_ids <- names %>% ids("ott")
```

The resulting data frame has at least one row per name provided, though possibily all NA.

This database includes synonyms, which do not have their own OTT identifiers in `taxonID`, but can be resolved to accepted names with accepted identifiers.  Thus, we want to look at `acceptedNameUsageID` column to see who matched and who didn't.  (This nomenclature for column names probably feels clumsy, but it comes from the Darwin Core standard and thus lets us be explicitly consistent with how these terms are used elsewhere.)

The `ids` function also automatically normalizes all strings to lowercase before matching against the (lowercased) scientificNames in the database.  `ids` returns these strings in the `input` column.  Standardizing between upper and lowercase reveals that many of the 37,505 names are still duplicates:

```{r}
length(ott$input)
length(unique(ott$input))
```

The `sort` column indicates the position of the original input data. Without this column, input names that are identical after correcting for capitalization result in identical rows.  We can drop these by filtering for only the distinct columns:

```{r}
clean_ott_ids <- ott_ids %>% select(-sort) %>% distinct()
clean_ott_ids # 29,878 rows
```

We can now check how many unique names were matched:

```{r}
matched <- clean_ott_ids %>% filter(!is.na(acceptedNameUsageID))
length(unique(matched$input)) # 18,568
length(unique(matched$acceptedNameUsageID)) #17,766

matched %>% filter(taxonomicStatus != "accepted") # 1,766
```

We matched 18,568 unique names (out of `r length(unique(clean_ott_ids$input))`), which resolved to 17,766 unique IDs in OTT, since 1,766 names were recognized synonyms to OTT.  Meanwhile, in the unmatched names set, this leaves us with:

```{r}
unmatched <- clean_ott_ids %>% filter(is.na(acceptedNameUsageID)) %>% pull(input)
length(unique(unmatched))
```

So 10,911 unmatched names still to resolve (a little more than one third of the unique names).

```{r}
wordcount <- 
  data.frame(input = unmatched) %>% 
  mutate(n = stringi::stri_count_regex(input, "\\s")) %>% 
  arrange(desc(n))

wordcount
```

Wow, so some of our entries have a whole lot more words than a species name: the top 6 entries have dozens of words:

```{r}
head(wordcount$input)
```

many more (1,551) have 4 words (three spaces), indicating subspecies and varities:

```{r}
wordcount %>% filter(n==3)
```

If we restrict ourselves to matching on the first two-word names, we have a reasonable chance of resolving names to the genus or species level.  `clean_names` does up to three transformations: we drop missing specific epithet indication `sp`, leaving only the genus name, we resolve the first two names, and we standardize delimiters between names.

```{r}
clean_unmatched <- unmatched %>% clean_names() %>% unique()
length(clean_unmatched)
```
This gives us 9303 unmatched names to resolve.

```{r}
ott_ids2 <- clean_unmatched %>% ids("ott")
unmatched2 <- ott_ids2 %>% filter(is.na(acceptedNameUsageID)) %>% pull(input)
```

We can bind the new matches to our existing matched names table:

```{r}
matched2 <- ott_ids2 %>% filter(!is.na(acceptedNameUsageID)) %>% bind_rows(matched)
```

With these clean names, another 1,927 can be matched to OTT ids, leaving 7,526 unmatched:

```{r}
length(unmatched2)
head(unmatched2, 10)
```

We can find at least some matches for these remaining names in most of the other databases:

```{r}
unmatched2 %>% ids("slb") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))
unmatched2 %>% ids("col") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))
unmatched2 %>% ids("gbif") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))
unmatched2 %>% ids("itis") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))
unmatched2 %>% ids("ncbi") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))
unmatched2 %>% ids("wd") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))
unmatched2 %>% ids("iucn") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))
unmatched2 %>% ids("tpl") %>% summarise(matched = sum(!is.na(acceptedNameUsageID)))

```
GBIF does the best among these, resolving another 1614 names exactly.

```{r}
```

### 2019-02-19 notebook/algae.Rmd
improving clean names utility
update example, docs

@@
 ```{r}
-taxadb::td_create() # only needed once.
+taxadb::td_create("all") # only needed once.
 algae <- read_csv("~/projects/algae-names/algae_uncleanednames_NAdropped.csv")
@@
 ```{r}
-length(ott$input)
-length(unique(ott$input))
+length(ott_ids$input)
+length(unique(ott_ids$input))
 ```
@@
 
-Wow, so some of our entries have a whole lot more words than a species name: the top 6 entries have dozens of words:
-
-```{r}
-head(wordcount$input)
-```
-
-many more (1,551) have 4 words (three spaces), indicating subspecies and varities:
+Wow, so some of our entries have a whole lot more words than a species name: the top 6 entries have dozens of words.  Many more (1,551) have 4 words (three spaces), indicating subspecies and varities:
 
@@
 ```
-This gives us 9303 unmatched names to resolve.
+This gives us `r length(clean_unmatched)` unmatched names to resolve.
 
@@
 
-With these clean names, another 1,927 can be matched to OTT ids, leaving 7,526 unmatched:
+With these clean names, another `r length(clean_unmatched) - length(umatched2)` can be matched to OTT ids, leaving `r length(umatched2)` unmatched:
 
@@
 head(unmatched2, 10)
+
+ott <- taxa_tbl("ott") %>% collect()
+ott_mash <- ott %>% mutate(input = clean_names(scientificName))
 ```
@@
 ```{r}
+gbif_ids <- unmatched2 %>% ids("gbif")
+unmatched3 <- gbif_ids %>% filter(is.na(acceptedNameUsageID)) %>% pull(input)
+length(names(umatched3))
 ```
+
+FIXME also try matching against `clean_names` versions of scientificName column on the database side; in particular, on the synonyms.

### 2019-02-19 R/clean_names.R

@@ -24,7 +24,7 @@
 #'
 #'
 #' @importFrom stringi stri_replace_all_regex stri_extract_all_words
-#' @importFrom stringi stri_trim stri_split_regex
+#' @importFrom stringi stri_trim stri_split_regex stri_trans_general
 #' @export
 #' @examples
 #' clean_names(c("Homo sapiens sapiens", "Homo.sapiens", "Homo sp."))
@@ -35,7 +35,7 @@ clean_names <-
            remove_sp = TRUE,
            ascii_only = TRUE){
     if(ascii_only)
-      names <- stri_trans_general(names,"latin-ascii")
+      names <- stringi::stri_trans_general(names,"latin-ascii")
     if(fix_delim)
       names <- set_space_delim(names)
     if(binomial_only)
@@ -67,20 +67,5 @@ drop_author_year <- function(x){
   stringi::stri_replace_all_regex(x, "\\(.+)", "")
 }
 
-## Function to strip ascii characters
-find_non_ascii <- function(string){
-  grep("I_WAS_NOT_ASCII",
-       iconv(string, "latin1", "ASCII", sub="I_WAS_NOT_ASCII"))
-
-}
-replace_non_ascii <-function(string){
-  i <- find_non_ascii(string)
-  non_ascii <- "áéíóúÁÉÍÓÚñÑüÜ’åôö"
-  ascii <- "aeiouAEIOUnNuU'aoo"
-  translated <- sapply(string[i], function(x)
-    chartr(non_ascii, ascii, x))
-  string[i] <- unname(translated)
-  string
-}
 
 

### 2019-02-20 notebook/algae.Rmd
exploring matching on split names

@@
 head(unmatched2, 10)
+```
 
-ott <- taxa_tbl("ott") %>% collect()
-ott_mash <- ott %>% mutate(input = clean_names(scientificName))
+Working from the names directly:
+
+```{r}
+ott <- taxa_tbl("ott") %>% 
+  mutate(name1 = splitpart(tolower(scientificName), " ", 1L),
+         name2 = splitpart(tolower(scientificName), " ", 2L)
+         )
+ott
+```
+
+```{r}
+x <- unmatched2 %>% stringi::stri_extract_all_words(simplify = TRUE) 
+colnames(x) <- c("name1", "name2")
+x <- as.data.frame(x)
+x$sort <- 1:length(x$name1)
+
+sp <-  x %>% filter(name2 != "")
+genus_only <- x %>% filter(name2 == "") %>% select(name1, sort)
+```
+
+```{r}
+match_genus <- right_join(ott, genus_only, copy=TRUE, by = "name1") %>% arrange(sort)
+match_genus %>% filter(!is.na(acceptedNameUsageID))
+```
+
+```{r}
+match_binomial <- right_join(ott, sp, copy=TRUE, by = c("name1", "name2")) %>% arrange(sort)
+match_binomial %>% filter(!is.na(acceptedNameUsageID))
 ```

### 2019-02-26 R/clean_names.R
WIP

@@ -61,7 +61,10 @@ binomial_names <- function(x){
   s <-
     stringi::stri_split_regex(x, "/", simplify = TRUE)[,1] %>%
     stringi::stri_extract_all_words(simplify = TRUE)
-  stringi::stri_trim(paste(s[,1], s[,2]))
+  if(dim(s)[2] > 1)
+    stringi::stri_trim(paste(s[,1], s[,2]))
+  else
+    stringi::stri_trim(s[,1])
 }
 drop_author_year <- function(x){
   stringi::stri_replace_all_regex(x, "\\(.+)", "")

### 2019-02-26 R/db_mutate.R

@@ -1,0 +1,55 @@
+#' Add new variables to a database
+#'
+#' [dplyr::mutate()] cannot pass arbitrary R functions over a database connection.
+#' This function provides a way to work around this, by querying the data in chunks
+#' and applying the function to each chunk, which is then appended back out to a
+#' temporary table.
+#' @param db a database connection, [DBI::DBIConnection()]
+#' @param tbl the name of a table
+#' @param r_fn any R function that can be called on a vector (column) of the table
+#' @param col the name of the column to which the R function is applied.
+#' (Note, [dplyr::mutate()] can operate on an arbitrary list of columns, this function
+#' only operates on a single column at this time...)
+#' @param new_column column name for the new column.
+#' @param n the number of rows included in each chunk, see [DBI::dbFetch()]
+#' @importFrom DBI dbCreateTable dbGetQuery dbSendQuery dbFetch
+#' @importFrom DBI dbWriteTable dbClearResult
+#' @importFrom progress progress_bar
+#'
+#' @export
+db_mutate <- function(db, tbl, r_fn, col, new_column, n = 5000L){
+
+  ## Create a temporary table which will store our data, including new column
+  tmp_tbl <- paste0("tmp_", paste0(sample(letters, 10, replace = TRUE), collapse = ""))
+  schema <- DBI::dbGetQuery(db, paste("SELECT * FROM", tbl, "LIMIT 1"))
+  schema[[new_column]] = r_fn(schema[[col]])
+  DBI::dbCreateTable(db, tmp_tbl, schema, temporary = TRUE)
+
+
+  ## Send the query -- we'll then page over the results in chunks.
+  res <- DBI::dbSendQuery(db, paste("SELECT * FROM", tbl))
+
+  ## Read table in by chunks & write out with mutated column
+  p <- progress::progress_bar$new("[:spin] chunk :current", total = 100000)
+  while (TRUE) {
+    p$tick()
+    chunk <- DBI::dbFetch(res, n = n)
+    if (nrow(chunk) == 0) break
+    chunk[[new_column]] = r_fn(chunk[[col]])
+    DBI::dbWriteTable(db, tmp_tbl, chunk, append=TRUE)
+  }
+  DBI::dbClearResult(res)
+
+  ## no need to join, new table is full copy of old table...
+  #fields <- DBI::dbListFields(db, tbl)
+  #dplyr::inner_join(dplyr::tbl(db,tmp_tbl), dplyr::tbl(db,tbl),
[…]

### 2019-02-26 R/ids.R

@@ -120,7 +120,6 @@ db_table <-
   mutate(input = tolower(scientificName),
          name1 = splitpart(input, " ", 1L),
          name2 = splitpart(input, " ", 2L))
-)
 
 }
 

### 2019-02-26 notebook/algae.Rmd
WIP

--- 2019-02-19 notebook/algae.Rmd
@@
 taxadb::td_create("all") # only needed once.
+```
+
+```{r}
 algae <- read_csv("~/projects/algae-names/algae_uncleanednames_NAdropped.csv")
@@
 head(unmatched2, 10)
+```
 
-ott <- taxa_tbl("ott") %>% collect()
-ott_mash <- ott %>% mutate(input = clean_names(scientificName))
+Working from the names directly:
+
+```{r}
+
+df <- db_mutate(r_fn = clean_names,
+          tbl = "ott",
+          db = td_connect(),
+          col = "scientificName",
+          new_column = "input")
+
+```
+
+```{r}
+sp <- tibble(input = unmatched2)
+
+match_binomial <- right_join(df, sp, copy=TRUE, by = "input") 
+match_binomial %>% filter(!is.na(acceptedNameUsageID))
 ```

### 2019-02-26 R/clean_names.R

@@ -33,7 +33,10 @@ clean_names <-
            fix_delim = TRUE,
            binomial_only = TRUE,
            remove_sp = TRUE,
-           ascii_only = TRUE){
+           ascii_only = TRUE,
+           lowercase = TRUE){
+    if(lowercase)
+      names <- stringi::stri_trans_tolower(names)
     if(ascii_only)
       names <- stringi::stri_trans_general(names,"latin-ascii")
     if(fix_delim)

### 2019-02-26 notebook/algae-explore.Rmd
cleaner mutate_db method
also tighter algae example in algae-explore

@@
 ```
+
+```{r}
+algae <- read_csv("~/projects/algae-names/algae_uncleanednames_NAdropped.csv")
+```
+
+```{r}
+## Input table with clean names
+algae <- algae %>% mutate(input = clean_names(species), sort = 1:length(input))
+
+## Let's get some matches
+taxa <- taxa_tbl("ott") %>% 
+        mutate_db(clean_names, "scientificName", "input") %>%
+        right_join(algae, copy=TRUE, by="input") %>% 
+        arrange(sort)  %>% 
+        collect()
+
+## lots of duplicate matches, pick the first one for now:
+matched <- taxa %>% select(acceptedNameUsageID, sort) %>% distinct() %>% 
+  group_by(sort) %>% top_n(1, acceptedNameUsageID)
+
+# 46,045 / 57,700 have been matched!
+
+## Who is unmatched? (sort id is in algae table but not in the matched table)
+unmatched <- anti_join(algae, matched, by="sort")
+
+# 11,655 are still unmatched.  Many appear to be known synonyms to Algaebase...
+unmatched %>% count(source) %>% arrange(desc(n))
+
+```
+
+-------
 

### 2019-02-26 R/mutate_db.R

@@ -4,23 +4,52 @@
 #' This function provides a way to work around this, by querying the data in chunks
 #' and applying the function to each chunk, which is then appended back out to a
 #' temporary table.
-#' @param db a database connection, [DBI::DBIConnection-class]
-#' @param tbl the name of a table
+#' @param .data A [dplyr::tbl] that uses a database connection, a [dplyr::tbl_dbi].
 #' @param r_fn any R function that can be called on a vector (column) of the table
 #' @param col the name of the column to which the R function is applied.
 #' (Note, [dplyr::mutate()] can operate on an arbitrary list of columns, this function
 #' only operates on a single column at this time...)
 #' @param new_column column name for the new column.
 #' @param n the number of rows included in each chunk, see [DBI::dbFetch()]
+#' @return a dplyr tbl connection to the temporary table in the database
+#' @importFrom dplyr tbl
+#' @export
+#'
+mutate_db <- function(.data,
+                      r_fn,
+                      col,
+                      new_column,
+                      n = 5000L){
+
+  if(!inherits(.data, "tbl_dbi"))
+    stop(paste("input must be a table from remote database connection"))
+
+  db <- .data$src$con
+  tmp_tbl <-
+    dbi_mutate(db = db,
+               tbl = as.character(.data$ops$x),
+               r_fn = r_fn, col = col, new_column = new_column,
+               n = 5000L, tmp_tbl = tmp_tablename())
+  dplyr::tbl(db, tmp_tbl)
+}
+
+
+#' DBI Mutate
+#'
+#' @inheritParams mutate_db
+#' @param db a database connection, [DBI::DBIConnection-class]
+#' @param tbl the name of a table
+#' @param tmp_tbl a name for the temporary table created
+#' @return the name of the temporary table created (invisibly).
+#' @noRd
 #' @importFrom DBI dbCreateTable dbGetQuery dbSendQuery dbFetch
 #' @importFrom DBI dbWriteTable dbClearResult
 #' @importFrom progress progress_bar
 #'
-#' @export
-db_mutate <- function(db, tbl, r_fn, col, new_column, n = 5000L){
+dbi_mutate <- function(db, tbl, r_fn, col, new_column, n = 5000L,
+                       tmp_tbl = tmp_tablename()){
 
[…]

### 2019-02-27 paper/manuscript.Rmd
exploring bird data in manuscript
and tweaking code along the way.

@@
 
+```{r include=FALSE}
+knitr::opts_chunk$set(message=FALSE)
+```
+
 The purpose and strengths of `taxadb` can be most easily seen through an introductory example.
 
-```{r}
+```{r message=FALSE}
 library(tidyverse)
@@
 
-```{r}
+```{r message=FALSE}
 td_create("all")
@@
 
-```{r}
+```{r message=FALSE}
 bbs <- read_tsv(system.file("extdata/bbs.tsv", package="taxadb"))
@@
 dim(bbs)
+
 ```
@@
 bbs_ids <- 
-bbs %>% select(species) %>%
-  mutate(gbif = get_ids(species, "gbif", "prefix"),
+bbs %>% 
+  select(species) %>%
+  mutate(
+         gbif = get_ids(species, "gbif", "prefix"),
           col = get_ids(species, "col", "prefix"),
@@
      wikidata = get_ids(species, "wd", "prefix"),
-  #       iucn = get_ids(species, "iucn", "prefix"),
+         iucn = get_ids(species, "iucn", "prefix"),
           ott = get_ids(species, "ott","prefix") 
@@
 ```{r}
-bbs_ott <- bbs$species %>% get_ids("ott", "prefix")
+ott_ids <- bbs %>% 
+  mutate(names = clean_names(species)) %>% 
+  pull(names) %>% 
+  ids("ott")
+
+unmatched <- ott_ids %>% 
+  filter(is.na(acceptedNameUsageID)) %>% 
+  select(input, sort)
+```
+
+```{r}
+unmatched$input %>% ids("iucn") %>% 
+  select(scientificName, taxonomicStatus, input, taxonID, acceptedNameUsageID)
+
+syn <- taxa_tbl("iucn") %>% 
+  right_join(
+    unmatched$input %>% 
+    ids("iucn") %>% 
+    select(acceptedNameUsageID),
+    copy = TRUE) %>%
+  select(scientificName, acceptedNameUsageID, taxonomicStatus) %>%
+  collect()
+
+## Some names have multiple synonyms, do not do this as a spread...    
+syn %>% mutate(n = 1:length(scientificName)) %>% spread(taxonomicStatus, scientificName)
+```
+
+```{r}
+
+## a name that resolves to multiple possible IDs is not considered a match by `get_ids()`
+ott_ids <- bbs %>% 
+  mutate(names = clean_names(species)) %>%
+  pull(names) %>% 
+  get_ids("ott", "prefix")
+
+length(!is.na(ott_ids))
 ```

### 2019-02-27 inst/examples/docker.R

@@ -1,0 +1,16 @@
+
+create_monetdb <- function(datadir = rappdirs::user_data_dir("taxadb")) {
+  if (!requireNamespace("stevedore", quietly = TRUE)) {
+    stop("Package stevedore must be installed to use this function")
+  }
+
+  dir.create(datadir, FALSE)
+  stopifnot(stevedore::docker_available())
+  docker <- stevedore::docker_client()
+  docker$container$run("monetdb/monetdb",
+    detach = TRUE,
+#    volumes = paste0(normalizePath(datadir), ":/var/monetdb5/dbfarm/taxadb"),
+#    env = c("DBA_PASSWORD" = "dba"),
+    ports = "127.0.0.1:5000:5000"
+  )
+}

### 2019-02-27 inst/examples/docker.R
WIP

@@ -3,14 +3,21 @@ create_monetdb <- function(datadir = rappdirs::user_data_dir("taxadb")) {
   if (!requireNamespace("stevedore", quietly = TRUE)) {
     stop("Package stevedore must be installed to use this function")
   }
+  if (!requireNamespace("MonetDB.R", quietly = TRUE)) {
+    stop("Package stevedore must be installed to use this function")
+  }
 
   dir.create(datadir, FALSE)
   stopifnot(stevedore::docker_available())
   docker <- stevedore::docker_client()
   docker$container$run("monetdb/monetdb",
     detach = TRUE,
-#    volumes = paste0(normalizePath(datadir), ":/var/monetdb5/dbfarm/taxadb"),
+    volumes = paste0(normalizePath(datadir), ":/var/monetdb5/dbfarm/taxadb"),
 #    env = c("DBA_PASSWORD" = "dba"),
-    ports = "127.0.0.1:5000:5000"
+    ports = "50000:50000"
+#    ports = "127.0.0.1:50000:50000"
   )
 }
+
+
+#conn <- dbConnect(MonetDB.R(), host="localhost", dbname="taxadb", user="monetdb", password="monetdb")

### 2019-02-27 notebook/synonyms.R

@@ -1,0 +1,39 @@
+globalVariables(c("taxonomicStatus", "scientificName", "taxonID",
+                  "taxonRank", "acceptedNameUsageID"))
+## A mapping in which synonym and accepted names are listed in the same row
+
+#' @importFrom dplyr full_join filter select
+syn_table <- function(taxon, accepted = "accepted"){
+
+  suppress_msg({
+    synonyms <- dplyr::full_join(
+      taxon %>%
+        dplyr::filter(taxonomicStatus != accepted) %>%
+        dplyr::select(synonym = scientificName,
+                      synonym_id = taxonID,
+                      taxonomicStatus,
+                      acceptedNameUsageID),
+      taxon %>%
+        dplyr::filter(taxonomicStatus == accepted) %>%
+        dplyr::select(acceptedNameUsage = scientificName,
+                      acceptedNameUsageID,
+                      taxonRank,
+                      taxonomicStatus))
+  })
+
+}
+
+
+#' synonyms
+#'
+#' Resolve provided list of names against all known synonyms
+#' @inheritParams ids
+#' @export
+synonyms <- function(name = NULL,
+                     provider = KNOWN_AUTHORITIES,
+                     collect = TRUE,
+                     db = td_connect()){
+  ids(name, provider, collect, db) %>%
+    syn_table()
+
+}

### 2019-02-28 paper/manuscript.Rmd
WIP
support for synonyms

@@
 bbs <- read_tsv(system.file("extdata/bbs.tsv", package="taxadb"))
-```
-
-```{r}
 dim(bbs)
-
 ```
@@
          iucn = get_ids(species, "iucn", "prefix"),
-          ott = get_ids(species, "ott","prefix") 
+          ott = get_ids(species, "ott",  "prefix") 
   )
@@
 ```{r}
-unmatched$input %>% ids("iucn") %>% 
-  select(scientificName, taxonomicStatus, input, taxonID, acceptedNameUsageID)
+syn <- unmatched$input %>% synonyms("iucn")
+syn
+```
 
-syn <- taxa_tbl("iucn") %>% 
-  right_join(
-    unmatched$input %>% 
-    ids("iucn") %>% 
-    select(acceptedNameUsageID),
-    copy = TRUE) %>%
-  select(scientificName, acceptedNameUsageID, taxonomicStatus) %>%
-  collect()
+Three of these synonyms are considered accepted names in OTT:
 
-## Some names have multiple synonyms, do not do this as a spread...    
-syn %>% mutate(n = 1:length(scientificName)) %>% spread(taxonomicStatus, scientificName)
+```{r}
+syn$synonym %>% na.omit() %>% ids("ott")
+```
+
+```{r}
+syn$synonym %>% na.omit() %>% get_ids("ott")
+syn %>% na.omit() %>% mutate(
+  #ott = get_ids(synonym, "ott"),
+                             col = get_ids(synonym, "col"),
+                             itis = get_ids(synonym, "itis"),
+                             gbif = get_ids(synonym, "gbif"))
+```
+
+```{r}
+
+syns <- unmatched$input %>% ids("iucn") %>%
+        dplyr::filter(taxonomicStatus != "accepted") %>%
+        dplyr::select(synonym = scientificName,
+                      synonym_id = taxonID,
+                      taxonomicStatus,
+                      acceptedNameUsageID)
+accepted
+
+syn <- unmatched$input %>% synonyms("iucn")
+syn
 ```

### 2019-02-28 R/synonyms.R

@@ -1,0 +1,68 @@
+
+#' synonyms
+#'
+#' Resolve provided list of names against all known synonyms
+#' @inheritParams ids
+#' @export
+synonyms <- function(name = NULL,
+                     provider = KNOWN_AUTHORITIES,
+                     collect = TRUE,
+                     db = td_connect()){
+
+
+  the_id_table <- ids(name, provider = provider, db = db)
+
+  ## Get both accepted names & synonyms for anything with an acceptedNameUsageID
+  taxadb:::suppress_msg({
+    syn <-
+      taxa_tbl(provider = provider, db = db) %>%
+      right_join(the_id_table %>%
+                   select(acceptedNameUsageID),
+                 by = "acceptedNameUsageID",
+                 copy = TRUE) %>%
+      select(scientificName, acceptedNameUsageID,
+             taxonomicStatus, taxonRank) %>%
+      syn_table()
+
+  })
+  ## Join that back onto the id table
+  out <- the_ids_table %>%
+    select(input, sort, acceptedNameUsageID) %>%
+    left_join(syn, by = "acceptedNameUsageID", copy = TRUE) %>%
+    # reorder
+    select("input", "acceptedNameUsage", "synonym",
+           "acceptedNameUsageID","taxonRank", "sort")
+
+  if (collect && inherits(out, "tbl_lazy")) {
+    return( dplyr::collect(out) )
+  }
+
+  out
+
+}
+
+
+globalVariables(c("taxonomicStatus", "scientificName", "taxonID",
+                  "taxonRank", "acceptedNameUsageID", "synonym", "input"))
+## A mapping in which synonym and accepted names are listed in the same row
+
+#' @importFrom dplyr full_join filter select
+syn_table <- function(taxon, accepted = "accepted"){
+
+  taxadb:::suppress_msg({
+
+    dplyr::full_join(
+      taxon %>%
+        dplyr::filter(taxonomicStatus != "accepted") %>%
+        dplyr::select(synonym = scientificName,
+                      acceptedNameUsageID),
+      taxon %>%
+        dplyr::filter(taxonomicStatus == "accepted") %>%
+        dplyr::select(acceptedNameUsage = scientificName,
+                      acceptedNameUsageID,
+                      taxonRank),
+      by = "acceptedNameUsageID")
+  })
+
+}
+

### 2019-02-28 tests/testthat/test-taxadb.R

@@ -43,6 +43,7 @@ test_that("we can set up a db and call basic functions", {
   expect_gt(dim(hier)[1], 1)
   expect_gt(dim(chameleons)[1], 1)
 
+  ## we can opt out of ignore_case on ids():
   species <- ids(df$scientificName,
                  db = db,
                  ignore_case = FALSE) %>%
@@ -51,6 +52,7 @@ test_that("we can set up a db and call basic functions", {
   expect_gt(dim(species)[1], 1)
 
 
+  ## descendants() can take ids instead of names:
   names <- descendants(id = df$taxonID,
                  db = db)
   expect_is(names, "data.frame")

### 2019-02-28 paper/manuscript.Rmd
WIP on manuscript

@@
 ---
+
+1. A problem already solved  including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List
+
+2.
+Performance.
+A fundamental challenge of most existing services is the use of internet-based Application Programming Interfaces, (web API)s to resolve individual names.
+The motivation for an API-based approach is reasonable: (1) Web APIs are a ubiquitous standard of data exchange (2) make very minimal assumptions about the language, software, or hardware avaialable to the user, (3) users access a single centrally managed database that is maintained and upgraded by the provider, and (4) the provider can easily control access to the data, measure usage statistics, and enforce rate limits.
+Many existing tools provide bindings of these APIs as functions to a specific language [@chamberlain2013; @taxize; @itis; @rotl; @worrms; @wikitaxa]
+
+Instead of binding existing web APIs, `taxadb` is built around a set compressed text files following a simple and consistent schema (discussed below).
+These files are automatically downloaded and imported and stored on a local database by `taxadb`.
+By default `taxadb` uses MonetDBLite [@monetdblite], a columnar-oriented relational database requiring no additional installation and providing persistent disk-based storage which can be orders of magnitude faster than traditional database servers such as Postgres [@monetdblite]
+
+3. Consistency
 
@@
 
+Of the remaining 10 names, 3 are clearly common names (`gull`, `tern`, `woodpecker`), two (`trochilid` and `ardeid`) are references to family names, *Trochilidae* and *Ardeidae*, and the remaining five are unresolved.
+Let us attempt to resolve these five before we consider the cases of common names and family names.
+
+Consulting IUCN Redlist names, we can resolve acceptedNameUsageIDs for four of these five:
+
+```{r}
+unmatched %>% mutate(iucn = get_ids(input, "iucn", "prefix"))
+```
+
+By consulting the synonyms table, we can see what other names these species might be known by:
+
 ```{r}
 syn <- unmatched$input %>% synonyms("iucn")
-syn
 ```
 
-Three of these synonyms are considered accepted names in OTT:
+```{r}
+x <- syn %>% 
+  mutate(ott_id = get_ids(synonym, "ott", "prefix")) %>% 
+  select(input, ott_id) %>%
+  na.omit()
+
+syn$synonym %>% get_ids("ott", "prefix")
+
+descendants(id = x$ott_id, provider = "ott") %>% select(scientificName, taxonomicStatus, acceptedNameUsageID)
+```
+
+```{r}
+x
+```
+
+Note that three of the four synonyms given by IUCN are considered accepted names in OTT:
 
@@
 
+We can look for these synonyms in other tables too. ITIS recognizes some of these unmatched names as synonyms and some as accepted names.
+
 ```{r}
-syn$synonym %>% na.omit() %>% get_ids("ott")
-syn %>% na.omit() %>% mutate(
-  #ott = get_ids(synonym, "ott"),
-                             col = get_ids(synonym, "col"),
-                             itis = get_ids(synonym, "itis"),
-                             gbif = get_ids(synonym, "gbif"))
+unmatched$input %>% synonyms("itis")
+
+unmatched$input %>% ids("itis")
+
+## 
+syn$synonym %>% na.omit() %>% unique() %>% synonyms("itis")
+
 ```
 
+We can confirm the family names can be resolved:
+
 ```{r}
+c("Trochilidae",  "Ardeidae") %>% ids("ott")
+```
 
-syns <- unmatched$input %>% ids("iucn") %>%
-        dplyr::filter(taxonomicStatus != "accepted") %>%
-        dplyr::select(synonym = scientificName,
-                      synonym_id = taxonID,
-                      taxonomicStatus,
-                      acceptedNameUsageID)
-accepted
-
-syn <- unmatched$input %>% synonyms("iucn")
-syn
-```
+We can use the common names table to resolve common names:
 
@@
 (GBIF lists *Antigone canadensis* as a synonym for *Grus canadensis*; this appears backwards as the Sandhill Crane was moved out of *Grus* after molecular analyses showed it made that taxon polyphyletic; https://en.wikipedia.org/wiki/Sandhill_crane).
-
-1. A problem already solved  including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List
-
-2.
-Performance.
-A fundamental challenge of most existing services is the use of internet-based Application Programming Interfaces, (web API)s to resolve individual names.
-The motivation for an API-based approach is reasonable: (1) Web APIs are a ubiquitous standard of data exchange (2) make very minimal assumptions about the language, software, or hardware avaialable to the user, (3) users access a single centrally managed database that is maintained and upgraded by the provider, and (4) the provider can easily control access to the data, measure usage statistics, and enforce rate limits.
-Many existing tools provide bindings of these APIs as functions to a specific language [@chamberlain2013; @taxize; @itis; @rotl; @worrms; @wikitaxa]
-
-Instead of binding existing web APIs, `taxadb` is built around a set compressed text files following a simple and consistent schema (discussed below).
-These files are automatically downloaded and imported and stored on a local database by `taxadb`.
-By default `taxadb` uses MonetDBLite [@monetdblite], a columnar-oriented relational database requiring no additional installation and providing persistent disk-based storage which can be orders of magnitude faster than traditional database servers such as Postgres [@monetdblite]
-
-3. Consistency

### 2019-02-28 paper/manuscript.Rmd

@@
 
-```{r}
-x
-```
-
 Note that three of the four synonyms given by IUCN are considered accepted names in OTT:
@@
 syn$synonym %>% na.omit() %>% ids("ott")
-```
-
-We can look for these synonyms in other tables too. ITIS recognizes some of these unmatched names as synonyms and some as accepted names.
-
-```{r}
-unmatched$input %>% synonyms("itis")
-
-unmatched$input %>% ids("itis")
-
-## 
-syn$synonym %>% na.omit() %>% unique() %>% synonyms("itis")
-
 ```
@@
 
-```{r}
-
-## a name that resolves to multiple possible IDs is not considered a match by `get_ids()`
-ott_ids <- bbs %>% 
-  mutate(names = clean_names(species)) %>%
-  pull(names) %>% 
-  get_ids("ott", "prefix")
-
-length(!is.na(ott_ids))
-```
-
-```{r}
-cleaned <- 
-  bbs$species[is.na(bbs_ott)] %>% 
-  clean_names() %>% 
-  get_ids("ott", "prefix")
-
-## use ID from cleaned name only if no exact match.
-bbs_ott[is.na(bbs_ott)] <- cleaned
-```
-
-We are still missing 10 species:
-
-```{r}
-still_missing <- bbs$species[is.na(bbs_ott)]
-still_missing
-```
-
-Three of these are recognizably common names, which we can resolve against the common names table:
-
-```{r}
-clean_names(still_missing) %>% ids("gbif")
-```
-
 (GBIF lists *Antigone canadensis* as a synonym for *Grus canadensis*; this appears backwards as the Sandhill Crane was moved out of *Grus* after molecular analyses showed it made that taxon polyphyletic; https://en.wikipedia.org/wiki/Sandhill_crane).

### 2019-03-04 paper/manuscript.Rmd
WIP

@@
 
-1. A problem already solved  including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List
+1. A problem already "solved" including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List
 
@@
 
+Outline strategy for name resolution.
+
+ID resolution including synonym mapping.
+
 ```{r}
@@
 
-`get_ids` by default looks only for exact matches.
+`get_ids` looks only for exact matches.
+Certain names may not uniquely resolve to a single ID, such as a name which is a recognized synonym for two separate taxonomic groups (such as may happen if a species is split.)
+
 We can get more matches by cleaning the names first.
 The `clean_names()` function provides several simple string manipulations to tidy names to improve the probability of a match: missing species (specific eptithets) such as `Accipiter sp.` drop the `sp.`, allowing matches against Genus names, and intraspecific epithets such as `Colaptes auratus cafer` are dropped to binomial names, `Colaptes auratus`.
-These transformations may not be appropriate for certain use cases, and should be used with care, and only once as many untransformed names as possible have already been matched.
+These transformations may not be appropriate for certain use cases, and should be used with care.
 
-In this example, applying this transformation on the unmatched names we can bring matches against OTT identifiers up from 665 to 740, missing only ten names:
+Observe that truncating names would result in a handful of subspecies being resolved only to the species level identifiers
 
 ```{r}
-ott_ids <- bbs %>% 
-  mutate(names = clean_names(species)) %>% 
-  pull(names) %>% 
-  ids("ott")
+bbs_cleaned_ids <- 
+  bbs_ids %>% 
+  select(species, ott) %>% 
+  mutate(ids_post_clean = 
+           get_ids(clean_names(species, binomial_only = TRUE), 
+                   "ott", "prefix"))
+bbs_cleaned_ids %>% na.omit() %>% filter(ott != ids_post_clean)
+x <- bbs_cleaned_ids %>% na.omit() %>% filter(ott != ids_post_clean)
 
+x
+descendants(id = x$ids_post_clean, provider = "ott")
+```
+
+```{r}
+bbs_ids %>% filter(is.na(ott) & !is.na(ids_post_clean))
+```
+
+```{r}
+pick <- function(A, B){ A[is.na(A)] <- B[is.na(A)]; A }
+
+ott_ids <- bbs_cleaned_ids %>% mutate(ids = pick(ott, ids_post_clean))
+```
+
+```{r}
 unmatched <- ott_ids %>% 
-  filter(is.na(acceptedNameUsageID)) %>% 
-  select(input, sort)
+  filter(is.na(ids)) %>% 
+  select(species, ids) %>%
+  mutate(species = clean_names(species))
+
+unmatched
 ```
@@
 ```{r}
-unmatched %>% mutate(iucn = get_ids(input, "iucn", "prefix"))
+
+unmatched %>% mutate(iucn = get_ids(species, "iucn", "prefix"))
 ```
@@
 ```{r}
-syn <- unmatched$input %>% synonyms("iucn")
+syn <- unmatched$species %>% synonyms("iucn")
+syn
 ```
@@
   na.omit()
+x
+```
 
+```{r}
 syn$synonym %>% get_ids("ott", "prefix")
+```
 
+```{r}
 descendants(id = x$ott_id, provider = "ott") %>% select(scientificName, taxonomicStatus, acceptedNameUsageID)
@@
 ```{r}
-syn$synonym %>% na.omit() %>% ids("ott")
+syn$synonym %>% na.omit() %>% ids("ott") %>% select(input, scientificName, acceptedNameUsageID, taxonomicStatus)
 ```

### 2019-03-06 paper/manuscript.Rmd
WIP

@@
 ---
+
+Synonyms vs accepted names. How many species of chameleons are described?
+
+Inconsistencies: errors or disputes? Taxon concepts.
+
+-----
 
@@
 ```{r}
-bbs_ids %>% filter(is.na(ott) & !is.na(ids_post_clean))
+bbs_cleaned_ids %>% filter(is.na(ott) & !is.na(ids_post_clean))
 ```
@@
 
+Alternately, sticking with the standard `dwc` schema:
+
+```{r}
+iucn_ids <- unmatched$species %>% ids("iucn") %>% select(acceptedNameUsageID) %>% distinct()
+right_join(taxa_tbl("iucn"), iucn_ids, copy=TRUE, by ="acceptedNameUsageID") %>% 
+  select(scientificName, taxonomicStatus, acceptedNameUsageID, taxonRank, taxonID)
+```
+
 ```{r}

### 2019-03-06 R/clean_names.R

@@ -12,8 +12,10 @@
 #' @param remove_sp Should we drop unspecified species epithet designations?
 #' e.g. `Homo sp.` becomes `Homo` (thus only matching against genus level ids).
 #' logical, default [TRUE].
-#' @param ascii_only should we coerce strings to ascii characters? (see [stringi::stri_trans_general()])
-#' @param lowercase should names be coerced to lower-case to provide case-insensitive matching?
+#' @param ascii_only should we coerce strings to ascii characters?
+#' (see [stringi::stri_trans_general()])
+#' @param lowercase should names be coerced to lower-case to provide
+#'  case-insensitive matching?
 #' @details Current implementation is limited to handling a few common cases.
 #' Additional extensions may be added later. A goal of the `clean_names` function
 #' is that any modification rule of the name strings be precise, atomic, and

### 2019-03-10 paper/manuscript.Rmd

@@
 ```{r}
-iucn_ids <- unmatched$species %>% ids("iucn") %>% select(acceptedNameUsageID) %>% distinct()
+iucn_ids <- unmatched$species %>% by_name("iucn") %>% select(acceptedNameUsageID) %>% distinct()
 right_join(taxa_tbl("iucn"), iucn_ids, copy=TRUE, by ="acceptedNameUsageID") %>% 
@@
 ```{r}
-descendants(id = x$ott_id, provider = "ott") %>% select(scientificName, taxonomicStatus, acceptedNameUsageID)
+by_id(id = x$ott_id, provider = "ott") %>% select(scientificName, taxonomicStatus, acceptedNameUsageID)
 ```
@@
 ```{r}
-syn$synonym %>% na.omit() %>% ids("ott") %>% select(input, scientificName, acceptedNameUsageID, taxonomicStatus)
+syn$synonym %>% na.omit() %>% by_name("ott") %>% select(input, scientificName, acceptedNameUsageID, taxonomicStatus)
 ```
@@
 ```{r}
-c("Trochilidae",  "Ardeidae") %>% ids("ott")
+c("Trochilidae",  "Ardeidae") %>% by_name("ott")
 ```

### 2019-03-14 paper/manuscript.Rmd
WIP

@@
 
-Synonyms vs accepted names. How many species of chameleons are described?
+```{r message=FALSE, include = FALSE}
+library(tidyverse)
+```
 
-Inconsistencies: errors or disputes? Taxon concepts.
+As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names.
+How many species are in the combined data?
+Do the studies use the same names for the same species, or do they use different synonyms for the same species?
+Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
+These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names.
+While these issues have long been recognized in the literature [], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
+@itis; @ncbi; @col], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
+Here, we present an R package, `taxadb`, which seeks to address this gap.
 
------
+Databases of taxonomic names such as the Integrated Taxonomic Information System [ITIS; @itis], the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi], the Catalogue of Life [COL; @col], and over one hundred other providers have sought to address these problems by providing expert-curated lists of accepted taxonomic names, synonyms, associated taxonomic rank, hierarchical classification, and scientific authority (e.g.
+author and date) establishing a scientific name.
+The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2012] has become a popular way R users to interact with naming providers and name resolution services.
+`taxize` implements bindings to the web APIs (Application Programming Interface) hosted by many popular taxonomic name providers.
+Unfortunately, this means that functions in the `taxize` are impacted by several major drawbacks that are inherent in the implementation of these central API servers, such as:
+
+- Queries require internet access at all times
+- Queries are slow and inefficient to implement and perform; frequently requiring separate API calls for each taxonomic name
+- The type of query is highly limited by the API design.
+For instance, it is impossible usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?"
+- Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
+- Most queries are not reproducible, as the results depend on the state of the central server (and potentially the quality of the internet connection).
+Many names providers update the server data either continuously or at regular intervals, including both revising existing names and adding new names.
+<!-- For instance, @predicts2014 observes that 2013 Catalogue of Life includes the genus "Notiophilus" in two separate families, contrary to the rules of nomenclature.
+This error does not occur in the 2018 edition.
+-->
+
+Instead of binding existing web APIs, `taxadb` is built around a set compressed text files following a consistent, standardized layout or schema (discussed below).
+These files are automatically downloaded and imported and stored on a local database by `taxadb`.
+The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes under a GB, and can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection.
+In contrast, querying a single name over the web API, requiring the server to respond, execute the query, and serialize the response, can take several seconds.
+Thus it does not take many taxa before transferring the entire data set to query locally is more efficient.
+Moreover, this local copy can be cached on the user's machine, requiring only the one-time setup and enabling offline use.
+
+After installing the `taxadb` R package, users can create local copies of data from any of the providers using the `td_create()` function and specifying the provider abbreviation (see Table 1), or `all` to install all available providers:
+
+```{r}
+library(taxadb)
+td_create("all")
+```
+
+<!--  Next 2 paragraphs should possibly be left to a manual, but are likely to be overlooked aspects there... -->
+This one-time download will download and import local copies of the provider data.
+By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage, and data is stored in the appropriate location specified for applications by the operating sytstem [@rappdirs].
+Users can customize this location, or opt for any alternative relational database backend by providing alternative arguments, though the default MonetDBLite system can be an order of magnitude faster than popular alternatives such as Postgres [@monetdblite].
+`taxadb` will automatically handle opening, caching, and closing the MonetDBLite connection, but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time.
+Users who need concurrent access from multiple sessions may wish to use a standalone MonetDB server or other database server instead.
+
+All `taxadb` functions can also operate in one-time-use mode without first installing a standalone server.
+If a function requests a provider which has not been installed, the relevant table can be downloaded and read directly into memory.
+This requires sufficient memory be available to hold the entire table (typically 1-2 GB) and will only be cached for the R session.
+This can be useful in special cases such as execution on remote servers where this will be faster than importing the data locally, or to allow other packages to use methods such as `get_ids()` internally without assuming users will have first created a local copy with `td_create()`.
+Specify `db=NULL` on any `taxadb` function to force this in-memory dispatch.
+
+Functions in `taxadb` are organized into several families:
+
+- database functions  `td_create()`, `td_connect()` and `taxa_tbl()`
+- queries that return vectors: `get_ids()` and it's complement, `get_names()`,
+- queries that filter the underlying taxonomic data frames: `by_name()`, `by_rank()`, `by_id()`, and `by_common()`,
+- and helper utilities, such as `clean_names()`.
+
+## Returning IDs
+
+Resolving names to IDs
+
+```{r}
+birds <- data.frame(species = c(
+"Leucocarbo colensoi",
+"Saxicola macrorhynchus",
+"Peneothello sigillata",
+"Ichthyaetus melanocephalus",
+"Phalacrocorax colensoi",
+"Saxicola macrorhyncha",
+"Peneothello sigillatus",
+"Poospiza lateralis lateralis",
+"Larus melanocephalus",
+"Myadestes elizabeth"
+), stringsAsFactors = FALSE)
+```
+
+Resolving names to identifiers (using the default provider, ITIS) reveals that we have only 5 unique ids, as many of our names were known synonyms:
+
+```{r}
+birds %>% mutate(id = get_ids(species)) %>% arrange(id)
+```
+This step has also revealed one name that could not be resolved.
+This could be a synonym not known to ITIS, or a species not known to ITIS, but often indicates a misspelling.
+A peek at other members of this genus starting with `"e"` might help us spot the issue:
+
+```{r}
+by_rank("Myadestes", "genus") %>% filter(grepl("^e", specificEpithet))
+```
+
+```{r}
+ids <- birds$species %>% get_ids() %>% unique() 
+data.frame(ids, species = get_names(ids))
+```
+
+```{r}
+birds %>%
+  mutate(id = get_ids(species), 
+         accepted = get_names(id)) %>%
+  select(id, accepted) %>%
+  distinct()
+
+```
+
+## Filtering operations
+
+```{r}
+by_name()
+```
+
+```{r}
+by_rank()
+```
+
+```{r}
+by_id()
+```
+
+## Database operations
+
+The `taxa_tbl()` function is the low-level interface at the heart of most `taxadb` functions, returning a connection to a single table in the database.
+
+## Subsetting the database
+
+Working directly with `taxa_tbl()` requires some familarity with `dplyr` and the way in which it operates when using remote connections.
+For many users this would create a significant hurdle: for instance, while typical operations like `select()`, `group_by()`, `summarise()` and `filter()` work in the same way on remote connections and in-memory data, operations that rely on specific R functions that `dplyr` cannot translate to SQL, such as
+
+"Which bird families starting with the letter 'B'?"
+
+```r
+taxa_tbl("itis") %>% filter(class=="Aves", grepl("B.*", family)) %>% count(family)
+```
+
+will result in an error, becuase `grepl` is an R-specific function.  Using the `by_*` functions, which `collect()` results into memory by default, will avoid this:
+
+```{r}
+by_rank("Aves", "class") %>% 
+  filter(grepl("B.*", family)) %>% 
+  count(family, sort = TRUE)
+```
+
+<!--
+The motivation for an API-based approach is reasonable: (1) Web APIs are a ubiquitous standard of data exchange (2) make very minimal assumptions about the language, software, or hardware avaialable to the user, (3) users access a single centrally managed database that is maintained and upgraded by the provider, and (4) the provider can easily control access to the data, measure usage statistics, and enforce rate limits.
+Many existing tools provide bindings of these APIs as functions to a specific language, including `taxize`, `tits`, `rotl`, `worms`, `redlist` and `wikitaxa` [@Chamberlain2013; @itis; @rotl; @worrms; @wikitaxa]
+-->
+
+Darwin Core schema.
+
+### Supported providers
+
+<!--
+comment that taxonomic backbones aren't really the same thing:
+In addition to these projects, large biodiversity informatics efforts such as the Global Biodiversity Information Facility [GBIF; @gbif], publish the 'taxonomic backbone' -->
+
+- Resolve names to identifiers
+- Resolve synonyms to accepted names
+- Associate names with taxonomic rank and heirarchical classification
+
+discuss use Scientific names vs identifiers.  taxonConcepts?
+
+several major ongoing efforts seek to address this problem, including:
+
+- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
+- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
+- Catalogue of Life annual Species list
+
+- `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.
+Instead, it fills an important gap between existing tools and typical research patterns.
+
+In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
+
+- `taxadb` works from versioned snapshots of the data providers.
+
+- All `taxadb` functions are specific to a provider, while operating consistently across different providers.  
+- `taxadb` does not attempt to present any unified taxonomic backbone or synthesis from existing naming providers, or make any assertions, numerical scores or other inferences about the data or matches to the data -- it is merely a tool for accessing this infromation.
+This contrasts from other approaches such as the Global Names Resolver, which
+
+Many taxonomic name providers are explicitly based on other existing naming assembeges.
+
+There is an important objection to this approach that must also be addressed.
+
+@Franz2018
+
+>  no self-respecting biodiversity researcher would or should trust aggregated data blindly, that indeed careful data cleaning is almost always necessary and expected to render the downloaded data fit for purpose, and that therefore aggregator services need to be understood mainly or merely as data discovery tools.
+
+> We reject this deflationary view for four reasons.
+For one, aggregators frequently blur the lines between advertizing their services just as a data discovery tool or as a more powerful data signal tool.
+Second, the biases inherent in using unitary backbones remain in place even if users are only interested in discovering all relevant data for their research purpose.
+If the backbone-based data record modulations are not easily retrievable through primary on-line interfaces, then users are significantly constrained in their ability to design search queries with high rates of precision and recall (81).
+Third, let us assume that labor-intensive off-line data quality review and correction efforts are indeed the norm, prior to publishing.
+Then why must the fruits of these efforts remain outside of the aggregator’s environment?
+Why can they not immediately flow back into the same aggregation domain, while recording the provenance of expert changes?
+In other words, if the workflow of rendering data fit for purpose flows only in one direction, i.e.
+from the on-line aggregate to the off-line quality review and publication, then our criticism of the design stands.
+
+The inability to make automate use and manipulation
+
+(Figure 1: schematic example of data join?)
+
+Assemble data across multiple studies
+
+How many species of Chameleons are known? There are over
+
+```{r}
+taxa_tbl("col") %>% count(taxonomicStatus, sort = TRUE)
+```
+
+```{r}
+taxa_tbl("col") %>% count(acceptedNameUsageID, sort=TRUE) %>% head()
+```
+
+```{r}
+taxa_tbl("iucn") %>% count()
+
+taxa_tbl("itis") %>%
+  filter(taxonomicStatus == "accepted") %>%
+  count()
+```
+
+131,042 of the 580,304 of *accepted names* in ITIS are not recognized by the Catalogue of Life as either accepted names or known synonyms.
+27,507 names in the IUCN Redlist are not in the 2018 Catalogue of Life.
+
+```{r}
+taxa_tbl("iucn") %>%
+  select(scientificName) %>% 
+  left_join(taxa_tbl("col")) %>% 
+  filter(is.na(acceptedNameUsageID)) %>% 
+  count()
+
+```
+
+```{r}
+most_synonyms <- taxa_tbl("col") %>% count(acceptedNameUsageID, sort=TRUE) %>% head() %>% collect()
+by_id(most_synonyms$acceptedNameUsageID, "col")
+most_synonyms %>% mutate(accepted = get_names(acceptedNameUsageID, "col"))
+
+synonyms("Mentha longifolia subsp. longifolia", "col")
+```
 
 1. A problem already "solved" including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List
-
-2.
-Performance.
-A fundamental challenge of most existing services is the use of internet-based Application Programming Interfaces, (web API)s to resolve individual names.
-The motivation for an API-based approach is reasonable: (1) Web APIs are a ubiquitous standard of data exchange (2) make very minimal assumptions about the language, software, or hardware avaialable to the user, (3) users access a single centrally managed database that is maintained and upgraded by the provider, and (4) the provider can easily control access to the data, measure usage statistics, and enforce rate limits.
-Many existing tools provide bindings of these APIs as functions to a specific language [@chamberlain2013; @taxize; @itis; @rotl; @worrms; @wikitaxa]
-
-Instead of binding existing web APIs, `taxadb` is built around a set compressed text files following a simple and consistent schema (discussed below).
-These files are automatically downloaded and imported and stored on a local database by `taxadb`.
-By default `taxadb` uses MonetDBLite [@monetdblite], a columnar-oriented relational database requiring no additional installation and providing persistent disk-based storage which can be orders of magnitude faster than traditional database servers such as Postgres [@monetdblite]
 
@@
   mutate(
-         gbif = get_ids(species, "gbif", "prefix"),
-          col = get_ids(species, "col",  "prefix"),
-         itis = get_ids(species, "itis", "prefix"),
-         ncbi = get_ids(species, "ncbi", "prefix"),
-     wikidata = get_ids(species, "wd",   "prefix"),
-         iucn = get_ids(species, "iucn", "prefix"),
-          ott = get_ids(species, "ott",  "prefix") 
+         gbif = get_ids(species, "gbif"),
+          col = get_ids(species, "col" ),
+         itis = get_ids(species, "itis"),
+         ncbi = get_ids(species, "ncbi"),
+           wd = get_ids(species, "wd"  ),
+         iucn = get_ids(species, "iucn"),
+          ott = get_ids(species, "ott" ) 
   )
@@
 x
-descendants(id = x$ids_post_clean, provider = "ott")
+by_id(x$ids_post_clean, provider = "ott")
 ```
@@
 ```{r}
-x <- syn %>% 
-  mutate(ott_id = get_ids(synonym, "ott", "prefix")) %>% 
-  select(input, ott_id) %>%
-  na.omit()
-x
-```
-
-```{r}
 syn$synonym %>% get_ids("ott", "prefix")
-```
-
-```{r}
-by_id(id = x$ott_id, provider = "ott") %>% select(scientificName, taxonomicStatus, acceptedNameUsageID)
 ```

### 2019-03-14 R/get_names.R

@@ -69,7 +69,7 @@ as_prefix <- function(x, provider){
 
 id_to_prefix <- function(x, provider){
   ## NAs
-  if(is.na(x)) return(NA)
+  if(is.na(x)) return(as.character(NA))
   ## Already prefix format
   if(grepl(paste0("^", toupper(provider), ":"),  x)) return(x)
   ## bare ids

### 2019-03-17 README.Rmd
update readme
updating README
also includes work on vignettes, articles

@@
 
-The goal of `taxadb` is to provide fast access to taxonomic data and manipulations, such as resolving taxonomic names to ids, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
+The goal of `taxadb` is to provide *fast*, *consistent* access to taxonomic data, supporting commont tasks such as resolving taxonomic names to identifiers, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
 These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records.
@@
 
-Before we can use most `taxadb` functions, we need to do a one-time installation of the database `taxadb` uses for almost all commands.
-This can take a while to run, but needs only be done once.
-The database is installed on your local hard-disk and will persist between R sessions.
-By default, the database will be installed in your user's application data directory, as detected by the `rappdirs` package.
-(Set a custom location instead using `dbdir` argument.)
+```{r}
+library(taxadb)
+library(dplyr)
+library(readr)
+```
+
+Create a local copy of the Catalogue of Life (2018) database:
 
 ```{r }
-library(taxadb)
-td_create()
+td_create("col")
 ```
 
-The default behavior installs only the ITIS database.  You can also specify a list of authorities to install, or install every authority using `td_create("all")`.
+Read in the species list used by the Breeding Bird Survey:
 
-## Test drive
+```{r, message = FALSE}
+bbs_species_list <- system.file("extdata/bbs.tsv", package="taxadb")
+bbs <- read_tsv(bbs_species_list)
+```
 
-Once the databases have been set up, we're ready to explore.  
-Here's a list of all the birds species known to ITIS:
+## Getting names and ids
+
+Two core functions are `get_ids()` and `get_names()`.
+These functions take a vector of names or ids (respectively), and return a vector of ids or names (respectively).
+For instance, we can use this to attempt to resolve all the bird names in the Breeding Bird Survey against the Catalogue of Life:
 
 ```{r}
-descendants(name = "Aves", rank = "class")
+birds <- bbs %>% 
+  select(species) %>% 
+  mutate(id = get_ids(species, "col"))
+
+head(birds, 10)
 ```
 
-## Learn More
+Note that some names cannot be resolved to an identifier.
+This can occur because of mis-spellings, non-standard formatting, or the use of a synonym not recognized by the naming provider.
+Names that cannot be uniquely resolved because they are known synonyms of multiple different species will also return `NA`.
+The `by_name` filtering functions can help us resolve this last case (see below).
 
-### [An introduction to taxadb](https://cboettig.github.io/taxadb/articles/articles/taxadb.html)
+`get_ids()` returns the IDs of accepted names, that is `dwc:AcceptedNameUsageID`s.  We can resolve the IDs into accepted names:
 
-The [taxadb introduction](https://cboettig.github.io/taxadb/articles/articles/taxadb.html) 
-provides an overview showing how `taxadb` functions can help us synthesize data across a given list of species by resolving synonyms and identifiers.
+```{r}
+birds %>% 
+  mutate(accepted_name = get_names(id, "col"))
+```
 
-### [taxadb schemas](https://cboettig.github.io/taxadb/articles/articles/schema.html)
+This illustrates that some of our names, e.g.
+*Dendrocygna bicolor* are accepted in the Catalogue of Life, while others, *Anser canagicus* are **known synonyms** of a different accepted name: **Chen canagica**.
+Resolving synonyms and accepted names to identifiers helps us avoid the possible miss-matches we could have when the same species is known by two different names.
 
-See the [schemas](https://cboettig.github.io/taxadb/articles/articles/schema.html) vignette for an overview of the underlying tables used by `taxadb` functions, and more about the different authorities accessed by taxadb.
+## Taxonomic Data Tables
+
+Local access to taxonomic data tables lets us do much more than look up names and ids.
+A family of `by_*` functions in `taxadb` help us work directly with subsets of the taxonomic data.
+As we noted above, this can be useful in resolving certain ambiguous names.
+
+For instance, *Trochalopteron henrici gucenense* does not resolve to an identifier in ITIS:
+
+```{r}
+get_ids("Trochalopteron henrici gucenense") 
+```
+
+Using `by_name()`, we find this is because the name resolves not to zero matches, but to more than one match:
+
+```{r}
+by_name("Trochalopteron henrici gucenense") 
+```
+
+```{r}
+by_name("Trochalopteron henrici gucenense")  %>%
+  mutate(acceptedNameUsage = get_names(acceptedNameUsageID)) %>% 
+  select(scientificName, taxonomicStatus, acceptedNameUsage, acceptedNameUsageID)
+```
+
+Similar functions `by_id`, `by_rank`, and `by_common` take IDs, scientific ranks, or common names, respectively.  Here, we can get taxonomic data on all bird names in the Catalogue of Life:
+
+```{r}
+by_rank(name = "Aves", rank = "class", provider = "col")
+```
+
+Combining these with `dplyr` functions can make it easy to explore this data: for instance, which families have the most species?
+
+```{r}
+by_rank(name = "Aves", rank = "class", provider = "col") %>%
+  filter(taxonomicStatus == "accepted", taxonRank=="species") %>% 
+  group_by(family) %>%
+  count(sort = TRUE) %>% 
+  head()
+```
+
+## Using the database connection directly
+
+`by_*` functions by default return in-memory data frames.
+Because they are filtering functions, they return a subset of the full data which matches a given query (names, ids, ranks, etc), so the returned data.frames are smaller than the full record of a naming provider.
+Working directly with the SQL connection to the MonetDBLite database gives us access to all the data.
+The `taxa_tbl()` function provides this connection:
+
+```{r}
+taxa_tbl("col")
+```
+
+We can still use most familiar `dplyr` verbs to perform common tasks.  For instance: which species has the most known synonyms?
+
+```{r}
+most_synonyms <- taxa_tbl("col") %>% 
+  group_by(acceptedNameUsageID) %>% 
+  count(sort=TRUE)
+most_synonyms
+```
+However, unlike the `by_*` functions which return convienent in-memory tables, this is still a remote connection.
+This means that direct access using the `taxa_tbl()` function (or directly accessing the database connection using `td_connect()`) is more low-level and requires greater care.
+For instance, we cannot just add a `%>% mutate(acceptedNameUsage = get_names(acceptedNameUsageID))` to the above, becuase `get_names` does not work on a remote collection.
+Instead, we would first need to use a `collect()` to pull the summary table into memory.
+Users familiar with remote databases in `dplyr` will find using `taxa_tbl()` directly to be convenient and fast, while other users may find the `by_*` approach to be more intuitive.
+
+So which species had those 456 names?
+
+```{r}
+most_synonyms %>% 
+  head(1) %>% 
+  pull(acceptedNameUsageID) %>% 
+  by_id("col") %>%
+  select(scientificName)
+```
 
@@
 MonetDBLite::monetdblite_shutdown()
+
+if(require(codemetar)) codemetar::write_codemeta()
 ```

### 2019-03-17 paper/manuscript.Rmd

@@
 As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names.
-How many species are in the combined data?
-Do the studies use the same names for the same species, or do they use different synonyms for the same species?
+How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
 Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
-These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names.
+These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
 While these issues have long been recognized in the literature [], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
@@
 library(taxadb)
+```
+
+```{r eval=FALSE, message=FALSE}
 td_create("all")
@@
 
-Resolving names to IDs
+Taxonomic identifiers provide a fundamental abstraction which lies at the heart of managing taxonomic names.
+For instance, by resolving scientific names to identifiers, we can idetnify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized.
+Unmatched names may indicate an error in data entry or otherwise warrant further investigation.
+Taxon identifiers area also easily resolved to the original authority (scientific publication) establishing the name.
+(The historical practice of appending an author and year to a scientific name, e.g.
+*Poa annua ssp.
+annua* (Smith 1912), serves a valuable role in disambiguating different uses of the same name but can be notoriously harder to resolve to the appropriate reference, while variation in this convention creates many distinct versions of the same name [@Patterson2010]).
+
+To illustrate this, we consider the following ten species names of different birds:
+
+<!-- Would this example be more compelling if we were joining two tables of data?  -->
 
@@
 
-Resolving names to identifiers (using the default provider, ITIS) reveals that we have only 5 unique ids, as many of our names were known synonyms:
+The `get_ids()` function in `taxadb` can quickly resolve thousands of species names to unique identifiers.
+Here, we see that resolving names to identifiers in ITIS reveals that we have only 5 unique ids, as many of our names were known synonyms:
+
+---
+
+Box: `get_ids()` returns the `acceptedNameUsageID`, the identifier associated with the *accepted* name.
+Some naming providers, such as ITIS and NCBI, provide taxonomic identifiers to both synonyms and accepted names.
+Other providers, such as COL and GBIF, only provide identifiers for accepted names.
+Common practice in Darwin Core archives is to provide an `acceptedNameUsageID` only for names which are synonyms, and otherwise to provide a `taxonID`.
+For accepted names, the `acceptedNameUsageID` is then given as missing (`NA`), while for synonyms, the `taxonID` may be missing (`NA`).
+In contrast, `taxadb` lists the `acceptedNameUsageID` for accepted names (where it matches the `taxonID`), as well as known synonyms.
+This is semantically identical, but also more convenient for database interfaces, since it allows a name to mapped to it's accepted identifier (or an identifier to map to it's accepted name usage) without the additional logic.
+For consistency, we will use the term "identifier" to mean the `acceptedNameUsageID` rather than the more ambiguous `taxonID` (which is undefined for synonyms listed by many providers), unless explicitly stated otherwise.
+
+---
 
 ```{r}
-birds %>% mutate(id = get_ids(species)) %>% arrange(id)
+birds <- birds %>% mutate(id = get_ids(species, "itis")) %>% arrange(id)
+birds
 ```
+
 This step has also revealed one name that could not be resolved.
@@
 
+Here we see that  `"Myadestes elizabeth"` was likely a miss-spelling of the ITIS-accepted name `"Myadestes elisabeth"`.
+
+```{r}
+get_ids("Myadestes elisabeth", provider = "itis", format = "uri")
+```
+
+It is equally important to be able to resolve scientific names from these identifiers.
+
 ```{r}
@@
 
+We can get a richer picture of the our matched names by using the table-functions such as `by_id` to return a table of taxonomic information about each identifier.
+For instance, with this information we can quickly verify that no name has been mistakenly matched to a taxon outside of the target group (*Aves*):
+
+ <!-- show multiple-matched cases -->
+
 ```{r}
-birds %>%
-  mutate(id = get_ids(species), 
-         accepted = get_names(id)) %>%
-  select(id, accepted) %>%
-  distinct()
+df <- by_id(birds$id, "itis") 
+df 
+```
 
+The column headings (though not necessarily the order) are consistent across `taxadb`, with all column names using terms defined by the Darwin Core ontology, [@dwc].
+Some of these (`acceptedNameUsageID`, `taxonRank`, `taxonomicStatus`) may be cumbersome, but are at least consistent and unambiguous.
+`taxadb` promises a core set of these columns are provided for (see Table 3)
+
+```{r}
+df %>% filter(class != "Aves")
 ```
@@
 ```{r}
-by_name()
+by_name(birds$species)
 ```
@@
 ```{r}
-by_rank()
+by_rank("Aves", rank="class")
 ```
@@
 ```{r}
-by_id()
+by_id(birds$id)
 ```

### 2019-03-22 paper/manuscript.Rmd
some small progress

@@
 ```{r message=FALSE, include = FALSE}
-library(tidyverse)
+pkgconfig::set_config("dplyr::na_matches" = "never")
 ```
@@
 
-To illustrate this, we consider the following ten species names of different birds:
-
-<!-- Would this example be more compelling if we were joining two tables of data?  -->
+These issues are best illustrated using a minimal example.
+We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@IUCN] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example consise enough for for visual presentation we will focus on a subset involving just 10 species:
 
 ```{r}
-birds <- data.frame(species = c(
-"Leucocarbo colensoi",
-"Saxicola macrorhynchus",
-"Peneothello sigillata",
-"Ichthyaetus melanocephalus",
-"Phalacrocorax colensoi",
-"Saxicola macrorhyncha",
-"Peneothello sigillatus",
-"Poospiza lateralis lateralis",
-"Larus melanocephalus",
-"Myadestes elizabeth"
-), stringsAsFactors = FALSE)
+trait_data <- read_tsv(system.file("extdata", "trait_data.tsv"))
+status_data <- read_tsv(system.file("extdata", "status_data.tsv"))
+```
+
+```{r}
+status_data
+```
+
+```{r}
+trait_data
+```
+
+If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved:
+
+```{r}
+full_join(trait_data, status_data, by = c("name_A" = "name_B"))
+```
+
+If we first resolve names used in each data set into shared identifiers, (for instance, using the Catalogue of Life), we discover that there is far more overlap in the species coverage than we might have realized at first.
+
+```{r}
+
+traits <- trait_data %>% mutate(id = get_ids(name_A, "col"))
+status <- status_data %>% mutate(id = get_ids(name_B, "col"))
+
+full_join(traits, status, by = "id") %>% 
+  select(id, name_A, mass, category, name_B) # just reordering
+
 ```
@@
 The `get_ids()` function in `taxadb` can quickly resolve thousands of species names to unique identifiers.
-Here, we see that resolving names to identifiers in ITIS reveals that we have only 5 unique ids, as many of our names were known synonyms:
 
@@
 
+Notice that one species name could not be resolved to a COL identifier.
+Perhaps we can resolve this name against a different authority?
+As it has an IUCN classification already, the IUCN naming provider seems a natural place to start.
+We can use the `synonyms()` table to attempt to identify any other scientific names by which this species is known:
+
 ```{r}
-birds <- birds %>% mutate(id = get_ids(species, "itis")) %>% arrange(id)
-birds
+synonyms("Scleroptila gutturalis", "iucn")
 ```
 
-This step has also revealed one name that could not be resolved.
-This could be a synonym not known to ITIS, or a species not known to ITIS, but often indicates a misspelling.
-A peek at other members of this genus starting with `"e"` might help us spot the issue:
+Note that one of these synonyms is recognized in COL, and in fact matches a COL identifier of *Francolinus levalliantoides*, 376.69 grams, in our data set.
 
 ```{r}
-by_rank("Myadestes", "genus") %>% filter(grepl("^e", specificEpithet))
+synonyms("Scleroptila gutturalis", "iucn") %>% 
+  pull(synonym) %>% 
+  get_ids("col")
 ```
 
-Here we see that  `"Myadestes elizabeth"` was likely a miss-spelling of the ITIS-accepted name `"Myadestes elisabeth"`.
+Walking synonyms between providers should be done with care.  Synonyms
+
+An accepted name can have many synonyms:
 
 ```{r}
-get_ids("Myadestes elisabeth", provider = "itis", format = "uri")
+most_synonyms <- taxa_tbl("col") %>% count(acceptedNameUsageID, sort = TRUE)
+most_synonyms
 ```
 
-It is equally important to be able to resolve scientific names from these identifiers.
+Here we see that some accepted identifiers, such as `COL:43082445`, are known by as many as 456 different synonyms!
+`get_ids()` function seamlessly handles the resolution of all synonyms to their accepted name usage identifier, since we can unambiguously map from many-to-one.
+
+This query uses the `taxa_tbl()` function to directly access the full taxonomic record in the MonetDBLite database.
+This approach is fast and powerful, but note that we are restricted to using `dplyr` functions, which can be directly translated into SQL, for such queries to work successfully.
+Though the result looks like a data frame, most base R operations will not work as expected:
 
 ```{r}
-ids <- birds$species %>% get_ids() %>% unique() 
-data.frame(ids, species = get_names(ids))
+most_synonyms$n
 ```
+Such queries are also not possible through the API or web interface to Catalogue of Life.
+To use custom R functions or functions from other packages we must first call the `dplyr` function `collect()` to return the table into R's memory.
+This should be done with care as an unfiltered table may require several gigabytes of memory to load.
 
-We can get a richer picture of the our matched names by using the table-functions such as `by_id` to return a table of taxonomic information about each identifier.
-For instance, with this information we can quickly verify that no name has been mistakenly matched to a taxon outside of the target group (*Aves*):
+Some synonyms may also be associated with more than one unique identifier.
+This situation is more difficult, as it is impossible to know which accepted name should be used for such synonyms without more context.
+The `get_ids()` function will thus return `NA` for any name that matches more than one identifier, just as it will return `NA` for a name that does not match any identifier at all.
+Other functions in `taxadb` allow us to explore these names further.
+We have already seen this strategy above
 
- <!-- show multiple-matched cases -->
+Having access to a local database makes it easy to explore such issues as "identify all synonyms that resolve to more than one accepted scientific name ID", like so:
 
 ```{r}
-df <- by_id(birds$id, "itis") 
-df 
+most_ambiguous <-
+taxa_tbl("col") %>% 
+  filter(taxonomicStatus != "accepted") %>% 
+  select(scientificName, acceptedNameUsageID) %>% 
+  distinct() %>% 
+  count(scientificName, sort=TRUE) %>%
+  filter(n > 1)
+most_ambiguous
 ```
+Here we see the name *Mabuya bistriata* is a synonym which may be associated with no fewer than 32 different accepted scientific names!
 
-The column headings (though not necessarily the order) are consistent across `taxadb`, with all column names using terms defined by the Darwin Core ontology, [@dwc].
-Some of these (`acceptedNameUsageID`, `taxonRank`, `taxonomicStatus`) may be cumbersome, but are at least consistent and unambiguous.
-`taxadb` promises a core set of these columns are provided for (see Table 3)
+For many applications, it is easier to work with the `by_*` class of functions, that first perform the relevant filtering join in the MonetDBLite database before returning the table in memory.
+This means that the `data.frame` objects returned by these functions behave as familiar objects.
+For instance, we can take the top 5 ambiguous names and see all the taxa to which those names may refer:
 
 ```{r}
-df %>% filter(class != "Aves")
-```
+top_5 <- most_ambiguous %>% 
+  head(5) %>% 
+  pull(scientificName)
 
-## Filtering operations
-
-```{r}
-by_name(birds$species)
+by_name(top_5, "col") %>%
+  select(scientificName, acceptedNameUsageID) %>% 
+  distinct() %>% 
+  mutate(acceptedNameUsage = get_names(acceptedNameUsageID, "col"))
 ```
@@
 ```{r}
-by_rank("Aves", rank="class")
-```
-
-```{r}
-by_id(birds$id)
-```
-
-## Database operations
-
-The `taxa_tbl()` function is the low-level interface at the heart of most `taxadb` functions, returning a connection to a single table in the database.
-
-## Subsetting the database
-
-Working directly with `taxa_tbl()` requires some familarity with `dplyr` and the way in which it operates when using remote connections.
-For many users this would create a significant hurdle: for instance, while typical operations like `select()`, `group_by()`, `summarise()` and `filter()` work in the same way on remote connections and in-memory data, operations that rely on specific R functions that `dplyr` cannot translate to SQL, such as
-
-"Which bird families starting with the letter 'B'?"
-
-```r
-taxa_tbl("itis") %>% filter(class=="Aves", grepl("B.*", family)) %>% count(family)
-```
-
-will result in an error, becuase `grepl` is an R-specific function.  Using the `by_*` functions, which `collect()` results into memory by default, will avoid this:
-
-```{r}
-by_rank("Aves", "class") %>% 
-  filter(grepl("B.*", family)) %>% 
-  count(family, sort = TRUE)
+by_rank("Aves", rank="class", provider = "col")
 ```
@@
 
+- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
+- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
+- Catalogue of Life annual Species list
+
 <!--
@@
 In addition to these projects, large biodiversity informatics efforts such as the Global Biodiversity Information Facility [GBIF; @gbif], publish the 'taxonomic backbone' -->
+
+## Discussion
 
@@
 
-several major ongoing efforts seek to address this problem, including:
-
-- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
-- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
-- Catalogue of Life annual Species list
-
 - `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.
 Instead, it fills an important gap between existing tools and typical research patterns.
-
 In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
@@
 - All `taxadb` functions are specific to a provider, while operating consistently across different providers.
+
 - `taxadb` does not attempt to present any unified taxonomic backbone or synthesis from existing naming providers, or make any assertions, numerical scores or other inferences about the data or matches to the data -- it is merely a tool for accessing this infromation.
 This contrasts from other approaches such as the Global Names Resolver, which
-
-Many taxonomic name providers are explicitly based on other existing naming assembeges.
 
@@
 
->  no self-respecting biodiversity researcher would or should trust aggregated data blindly, that indeed careful data cleaning is almost always necessary and expected to render the downloaded data fit for purpose, and that therefore aggregator services need to be understood mainly or merely as data discovery tools.
+>>  no self-respecting biodiversity researcher would or should trust aggregated data blindly, that indeed careful data cleaning is almost always necessary and expected to render the downloaded data fit for purpose, and that therefore aggregator services need to be understood mainly or merely as data discovery tools.
 
@@
 
-```{r}
-taxa_tbl("col") %>% count(acceptedNameUsageID, sort=TRUE) %>% head()
-```
+----
+
+The extent of mismatch between providers can be startling.
+Catalogue of Life uses ITIS data as one of it's many input sources, and yet we can quickly calculate how many of the (2018) ITIS accepted species names are not in the Catalogue of Life records:
 
 ```{r}
-taxa_tbl("iucn") %>% count()
+itis_col <-
+taxa_tbl("itis") %>%
+  filter(taxonomicStatus == "accepted", taxonRank == "species") %>%
+  select(scientificName) %>% 
+  left_join(taxa_tbl("col"), by="scientificName") %>% 
+  filter(is.na(acceptedNameUsageID)) %>% 
+  count() %>%
+  pull(n)
 
-taxa_tbl("itis") %>%
-  filter(taxonomicStatus == "accepted") %>%
-  count()
+itis_accepted <- taxa_tbl("itis") %>%
+  filter(taxonomicStatus == "accepted", taxonRank == "species") %>% 
+  count() %>% pull(n)
 ```
-
-131,042 of the 580,304 of *accepted names* in ITIS are not recognized by the Catalogue of Life as either accepted names or known synonyms.
-27,507 names in the IUCN Redlist are not in the 2018 Catalogue of Life.
-
-```{r}
-taxa_tbl("iucn") %>%
-  select(scientificName) %>% 
-  left_join(taxa_tbl("col")) %>% 
-  filter(is.na(acceptedNameUsageID)) %>% 
-  count()
-
-```
-
-```{r}
-most_synonyms <- taxa_tbl("col") %>% count(acceptedNameUsageID, sort=TRUE) %>% head() %>% collect()
-by_id(most_synonyms$acceptedNameUsageID, "col")
-most_synonyms %>% mutate(accepted = get_names(acceptedNameUsageID, "col"))
-
-synonyms("Mentha longifolia subsp. longifolia", "col")
-```
-
-1. A problem already "solved" including ITIS, NCBI, CoL, GBIF, WikiData, FishBase, and The Plant List
-
-3. Consistency
-
-```{r include=FALSE}
-knitr::opts_chunk$set(message=FALSE)
-```
-
-The purpose and strengths of `taxadb` can be most easily seen through an introductory example.
-
-```{r message=FALSE}
-library(tidyverse)
-library(taxadb)
-```
-
-```{r message=FALSE}
-td_create("all")
-```
-
-```{r message=FALSE}
-bbs <- read_tsv(system.file("extdata/bbs.tsv", package="taxadb"))
-dim(bbs)
-```
-
-Outline strategy for name resolution.
-
-ID resolution including synonym mapping.
-
-```{r}
-bbs_ids <- 
-bbs %>% 
-  select(species) %>%
-  mutate(
-         gbif = get_ids(species, "gbif"),
-          col = get_ids(species, "col" ),
-         itis = get_ids(species, "itis"),
-         ncbi = get_ids(species, "ncbi"),
-           wd = get_ids(species, "wd"  ),
-         iucn = get_ids(species, "iucn"),
-          ott = get_ids(species, "ott" ) 
-  )
-
-```
-
-```{r}
-bbs_ids %>% 
-  select(-species) %>% 
-  purrr::map_dbl(function(x) sum(!is.na(x)))
-```
-
-`get_ids` looks only for exact matches.
-Certain names may not uniquely resolve to a single ID, such as a name which is a recognized synonym for two separate taxonomic groups (such as may happen if a species is split.)
-
-We can get more matches by cleaning the names first.
-The `clean_names()` function provides several simple string manipulations to tidy names to improve the probability of a match: missing species (specific eptithets) such as `Accipiter sp.` drop the `sp.`, allowing matches against Genus names, and intraspecific epithets such as `Colaptes auratus cafer` are dropped to binomial names, `Colaptes auratus`.
-These transformations may not be appropriate for certain use cases, and should be used with care.
-
-Observe that truncating names would result in a handful of subspecies being resolved only to the species level identifiers
-
-```{r}
-bbs_cleaned_ids <- 
-  bbs_ids %>% 
-  select(species, ott) %>% 
-  mutate(ids_post_clean = 
-           get_ids(clean_names(species, binomial_only = TRUE), 
-                   "ott", "prefix"))
-bbs_cleaned_ids %>% na.omit() %>% filter(ott != ids_post_clean)
-x <- bbs_cleaned_ids %>% na.omit() %>% filter(ott != ids_post_clean)
-
-x
-by_id(x$ids_post_clean, provider = "ott")
-```
-
-```{r}
-bbs_cleaned_ids %>% filter(is.na(ott) & !is.na(ids_post_clean))
-```
-
-```{r}
-pick <- function(A, B){ A[is.na(A)] <- B[is.na(A)]; A }
-
-ott_ids <- bbs_cleaned_ids %>% mutate(ids = pick(ott, ids_post_clean))
-```
-
-```{r}
-unmatched <- ott_ids %>% 
-  filter(is.na(ids)) %>% 
-  select(species, ids) %>%
-  mutate(species = clean_names(species))
-
-unmatched
-```
-
-Of the remaining 10 names, 3 are clearly common names (`gull`, `tern`, `woodpecker`), two (`trochilid` and `ardeid`) are references to family names, *Trochilidae* and *Ardeidae*, and the remaining five are unresolved.
-Let us attempt to resolve these five before we consider the cases of common names and family names.
-
-Consulting IUCN Redlist names, we can resolve acceptedNameUsageIDs for four of these five:
-
-```{r}
-
-unmatched %>% mutate(iucn = get_ids(species, "iucn", "prefix"))
-```
-
-By consulting the synonyms table, we can see what other names these species might be known by:
-
-```{r}
-syn <- unmatched$species %>% synonyms("iucn")
-syn
-```
-
-Alternately, sticking with the standard `dwc` schema:
-
-```{r}
-iucn_ids <- unmatched$species %>% by_name("iucn") %>% select(acceptedNameUsageID) %>% distinct()
-right_join(taxa_tbl("iucn"), iucn_ids, copy=TRUE, by ="acceptedNameUsageID") %>% 
-  select(scientificName, taxonomicStatus, acceptedNameUsageID, taxonRank, taxonID)
-```
-
-```{r}
-syn$synonym %>% get_ids("ott", "prefix")
-```
-
-Note that three of the four synonyms given by IUCN are considered accepted names in OTT:
-
-```{r}
-syn$synonym %>% na.omit() %>% by_name("ott") %>% select(input, scientificName, acceptedNameUsageID, taxonomicStatus)
-```
-
-We can confirm the family names can be resolved:
-
-```{r}
-c("Trochilidae",  "Ardeidae") %>% by_name("ott")
-```
-
-We can use the common names table to resolve common names:
-
-(GBIF lists *Antigone canadensis* as a synonym for *Grus canadensis*; this appears backwards as the Sandhill Crane was moved out of *Grus* after molecular analyses showed it made that taxon polyphyletic; https://en.wikipedia.org/wiki/Sandhill_crane).
+`r itis_col` of the `r itis_accepted` of *accepted names* in ITIS (`r itis_col / itis_accepted * 100`%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.

### 2019-03-22 R/synonyms.R

@@ -35,7 +35,7 @@ synonyms <- function(name,
     syn <-
       taxa_tbl(provider = provider, db = db) %>%
       dplyr::right_join(the_id_table %>%
-                   select(acceptedNameUsageID),
+                   select("acceptedNameUsageID"),
                  by = "acceptedNameUsageID",
                  copy = TRUE) %>%
       dplyr::select("scientificName", "acceptedNameUsageID",
@@ -47,8 +47,8 @@ synonyms <- function(name,
   out <- the_id_table %>%
     dplyr::select("scientificName", "sort", "acceptedNameUsageID") %>%
     dplyr::left_join(syn, by = "acceptedNameUsageID", copy = TRUE) %>%
-    distinct() %>%
-    dplyr::select("acceptedNameUsage", "synonym", "taxonRank")   # reorder
+    dplyr::select("acceptedNameUsage", "synonym", "taxonRank", "acceptedNameUsageID") %>%
+    dplyr::distinct()
 
   if (collect && inherits(out, "tbl_lazy")) {
     return( dplyr::collect(out) )

### 2019-03-22 data-raw/extdata.R

@@ -1,6 +1,11 @@
+library(readr)
+library(dplyr)
+library(purrr)
+
 ## Script to prepare the data files included in extdata from original sources
 
 
+## Breeding Bird Survey
 url <- "ftp://ftpext.usgs.gov/pub/er/md/laurel/BBS/DataFiles/SpeciesList.txt"
 col_names <- c("Seq", "AOU", "English_Common_Name", "French_Common_Name",
                "Spanish_Common_Name", "order", "family", "genus", "species")
@@ -10,3 +15,56 @@ bbs <- readr::read_table(url,
                          col_types = "ccccccccc")
 bbs <- bbs %>% mutate(species = paste(genus, species))
 readr::write_tsv(bbs,  "../inst/extdata/bbs.tsv")
+
+
+## Elton bird trait data
+elton <-
+  read_tsv("https://ndownloader.figshare.com/files/5631081", guess_max = 10000) %>%
+  select(Scientific, mass = `BodyMass-Value`, common = English)%>%
+  mutate(itis = get_ids(Scientific, "itis"))
+
+## Subset of data where names are synonyms by ITIS
+traits_ <-
+  elton$Scientific %>%
+  by_name("itis") %>%
+  filter(taxonomicStatus != "accepted") %>%
+  select(Scientific = scientificName) %>%
+  inner_join(elton) %>%
+  head(10) %>%
+  select(elton_name = Scientific, mass) %>%
+  mutate(name_A = get_names(get_ids(elton_name)))
+
+
+key <- "9bb4facb6d23f48efbf424bb05c0c1ef1cf6f468393bc745d42179ac4aca5fee"
+status_data <-
+  traits_$elton_name  %>%
+  get_ids("iucn") %>% #could use IUCN id directly
+  get_names("iucn") %>%
+  map(rredlist::rl_search, key = key) %>%
+  map_df("result") %>%
+  select(name_B = scientific_name, category) %>%
+  as_tibble()
+
+trait_data <-
+  traits_ %>%
+  mutate(name_A = coalesce(name_A, elton_name)) %>% # replace NA with fallback name
+  select(name_A, mass)
+
+
+readr::write_tsv(trait_data,  "inst/extdata/trait_data.tsv")
+readr::write_tsv(status_data,  "inst/extdata/status_data.tsv")
+
+
+
+## bad -- only a single species resulves both mass and
+#full_join( trait_data, status_data, by = c("name_A"= "name_B" ),
[…]

### 2019-03-27 README.Rmd
build pkgdown
update docs

@@
 
-```{r}
+```{r message = FALSE}
 library(taxadb)
@@
 
+## Learn more
+
+See richer examples, including name cleaning and background on the schema in the vignettes:
+
+- [Tutorial](https://cboettig.github.io/taxadb/articles/intro.html)
+- [Schema](https://cboettig.github.io/taxadb/articles/articles/schema.html)
+
 ```{r include=FALSE}

### 2019-04-01 paper/manuscript.Rmd
WIP

@@
 bibliography: refs.bib
+layout: 3p
 output: rticles::elsevier_article
@@
 ```{r message=FALSE, include = FALSE}
+
 pkgconfig::set_config("dplyr::na_matches" = "never")
+
+library(printr)
+knit_print.data.frame <- function(x, ...) {
+#  knitr::knit_print(DT::datatable(x), ...)
+  knitr::knit_print(kableExtra::kable(x, "latex"), ...)
+}
+
+knitr::opts_chunk$set(cache=FALSE)
 ```
@@
 author and date) establishing a scientific name.
-The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2012] has become a popular way R users to interact with naming providers and name resolution services.
+The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013] has become a popular way R users to interact with naming providers and name resolution services.
 `taxize` implements bindings to the web APIs (Application Programming Interface) hosted by many popular taxonomic name providers.
@@
 
-```{r}
+```{r message = FALSE, warning=FALSE}
+library(tidyverse)
 library(taxadb)
@@
 ```{r}
-trait_data <- read_tsv(system.file("extdata", "trait_data.tsv"))
-status_data <- read_tsv(system.file("extdata", "status_data.tsv"))
+trait_data <- read_tsv(system.file("extdata", "trait_data.tsv", package="taxadb"))
+status_data <- read_tsv(system.file("extdata", "status_data.tsv", package="taxadb"))
 ```
 
-```{r}
+```{r }
 status_data
@@
   distinct() %>% 
-  mutate(acceptedNameUsage = get_names(acceptedNameUsageID, "col"))
+  mutate(acceptedNameUsage = get_names(acceptedNameUsageID, "col")) %>% 
+  head(5)
 ```
@@
 ```{r}
-by_rank("Aves", rank="class", provider = "col")
+by_rank("Aves", rank="class", provider = "col") %>% head(5)
 ```
@@
 ```
-`r itis_col` of the `r itis_accepted` of *accepted names* in ITIS (`r itis_col / itis_accepted * 100`%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.
+$`r itis_col`$ of the $`r itis_accepted`$ of *accepted names* in ITIS ($`r itis_col / itis_accepted * 100`$%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.

### 2019-04-02 paper/manuscript.Rmd
updating refs

@@
 
-knitr::opts_chunk$set(cache=FALSE)
+knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
 ```
@@
 These issues are best illustrated using a minimal example.
-We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@IUCN] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example consise enough for for visual presentation we will focus on a subset involving just 10 species:
+We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example consise enough for for visual presentation we will focus on a subset involving just 10 species:
 

### 2019-04-03 paper/manuscript.Rmd
WIP
pdf

@@
 layout: 3p
-output: rticles::elsevier_article
+output: 
+  rticles::springer_article:
+    includes:
+      in_header: preamble.tex
 ---
@@
 ```{r message=FALSE, include = FALSE}
+library(kableExtra)
+library(magrittr)
+library(tidyverse)
+library(taxadb)
 
@@
 
-library(printr)
-knit_print.data.frame <- function(x, ...) {
-#  knitr::knit_print(DT::datatable(x), ...)
-  knitr::knit_print(kableExtra::kable(x, "latex"), ...)
-}
-
-knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)
 ```
@@
 
-```{r }
-status_data
+```{r echo=FALSE, cache = FALSE}
+status_data %>% kableExtra::kable("latex", booktabs=T) %>% 
+                      kable_styling(latex_options = "striped", full_width = F)
 ```
 
-```{r}
-trait_data
+```{r echo = FALSE, cache = FALSE }
+trait_data %>% kableExtra::kable("latex", booktabs=T) %>% 
+                      kable_styling(latex_options = "striped", full_width = F)
 ```
@@
 ```{r}
-full_join(trait_data, status_data, by = c("name_A" = "name_B"))
+joined <- full_join(trait_data, status_data, by = c("elton_name" = "redlist_name"))
+```
+
+```{r echo = FALSE, cache = FALSE}
+joined %>% kableExtra::kable("latex", booktabs=T) %>% 
+                      kable_styling(latex_options = "striped", full_width = F)
 ```
@@
 If we first resolve names used in each data set into shared identifiers, (for instance, using the Catalogue of Life), we discover that there is far more overlap in the species coverage than we might have realized at first.
+First, we just add an ID column to each table by looking up the Catalog of Life identifier for the name provided:
 
@@
 
-traits <- trait_data %>% mutate(id = get_ids(name_A, "col"))
-status <- status_data %>% mutate(id = get_ids(name_B, "col"))
+traits <- trait_data %>% mutate(id = get_ids(elton_name, "col"))
+status <- status_data %>% mutate(id = get_ids(redlist_name, "col"))
+```
 
-full_join(traits, status, by = "id") %>% 
-  select(id, name_A, mass, category, name_B) # just reordering
+We can now join on the `id` column instead of names directly:
 
+```{r}
+joined <- full_join(traits, status, by = "id") 
+```
+
+```{r cache = FALSE, echo = FALSE}
+joined %>%  
+  select(elton_name, redlist_name, mass, category, id) %>%
+  kableExtra::kable("latex", booktabs=T) %>% 
+  kable_styling(latex_options = "striped", full_width = F)
 ```
@@
 
----
+\dummy{\Begin{tcolorbox}[title= Box 1: Taxonomic Identifiers and Synonyms, lower separated=false]}
 
-Box: `get_ids()` returns the `acceptedNameUsageID`, the identifier associated with the *accepted* name.
+`get_ids()` returns the `acceptedNameUsageID`, the identifier associated with the *accepted* name.
 Some naming providers, such as ITIS and NCBI, provide taxonomic identifiers to both synonyms and accepted names.
@@
 
----
+\End{tcolorbox}
 
@@
 
-Walking synonyms between providers should be done with care.  Synonyms
+## Synonyms
 
-An accepted name can have many synonyms:
+Walking synonyms between providers should be done with care.  An accepted name can have many synonyms:
 
@@
 $`r itis_col`$ of the $`r itis_accepted`$ of *accepted names* in ITIS ($`r itis_col / itis_accepted * 100`$%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.
+
+\pagebreak
+
+# References

### 2019-04-03 data-raw/extdata.R

@@ -50,6 +50,8 @@ trait_data <-
   mutate(name_A = coalesce(name_A, elton_name)) %>% # replace NA with fallback name
   select(name_A, mass)
 
+trait_data <- trait_data %>% rename(elton_name = name_A)
+status_data <- status_data %>% rename(redlist_name = name_B)
 
 readr::write_tsv(trait_data,  "inst/extdata/trait_data.tsv")
 readr::write_tsv(status_data,  "inst/extdata/status_data.tsv")

### 2019-04-03 paper/manuscript.Rmd
table printing

@@
 output: 
-  rticles::springer_article:
+  rticles::elsevier_article:
     includes:
@@
 
+printtable <- function(df){
+  df %>% 
+    kableExtra::kable("latex", booktabs=T) %>% 
+    kable_styling(full_width = F)
+}
 knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)
@@
 ```{r echo=FALSE, cache = FALSE}
-status_data %>% kableExtra::kable("latex", booktabs=T) %>% 
-                      kable_styling(latex_options = "striped", full_width = F)
+status_data %>% printtable()
 ```
@@
 ```{r echo = FALSE, cache = FALSE }
-trait_data %>% kableExtra::kable("latex", booktabs=T) %>% 
-                      kable_styling(latex_options = "striped", full_width = F)
+trait_data %>% printtable()
 ```
@@
 ```{r echo = FALSE, cache = FALSE}
-joined %>% kableExtra::kable("latex", booktabs=T) %>% 
-                      kable_styling(latex_options = "striped", full_width = F)
+joined %>% printtable()
 ```
@@
 joined %>%  
-  select(elton_name, redlist_name, mass, category, id) %>%
-  kableExtra::kable("latex", booktabs=T) %>% 
-  kable_styling(latex_options = "striped", full_width = F)
+  select(elton_name, redlist_name, mass, category, id) %>% printtable()
 ```

### 2019-04-03 paper/manuscript.Rmd
writing

@@
   2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
-  3) `taxadb` provides a consistent and intuitive data format
+  3) `taxadb` provides a consistent and intuitive data format.  
   4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.
@@
 knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)
+
+# no sci notation integers pleases
+options(scipen=999)
+
 ```
@@
 
-## Returning IDs
+## Taxonomic IDs
 
@@
 ```{r echo = FALSE, cache = FALSE}
-joined %>% printtable()
+joined %>% tidyr::replace_na(list(category = "-")) %>% printtable() # "NA" looks like a category, "-" is clearer visually
 ```
@@
 ```{r cache = FALSE, echo = FALSE}
+## Just for pretty-printing
 joined %>%  
-  select(elton_name, redlist_name, mass, category, id) %>% printtable()
+  tidyr::replace_na(list(category = "-", elton_name = "-", redlist_name = "-")) %>%
+  select(elton_name, redlist_name, mass, category, id) %>%
+  printtable()
 ```
 
-The `get_ids()` function in `taxadb` can quickly resolve thousands of species names to unique identifiers.
+This results in many more matches, as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.
+While we have focused on a small example for visual clarity here, the `get_ids()` function in `taxadb` can quickly resolve hundreds of thousands of species names to unique identifiers, thanks to the performance of fast joins in a local MonetDBLite database.
 
@@
 
+## Unresolved names
+
 Notice that one species name could not be resolved to a COL identifier.
+This could indicate a data entry problem such as a misspelling of a name, but failure to resolve properly specified scientific names is not uncommon, as even large scale efforts like the Catalogue of Life are far from a comprehensive list of all published scientific names.
+
 Perhaps we can resolve this name against a different authority?
@@
 
-## Synonyms
+## Additional taxonomic information
 
-Walking synonyms between providers should be done with care.  An accepted name can have many synonyms:
+```{r}
+by_rank("Aves", rank="class", provider = "col") %>% head(5)
+```
+
+## Caution using synonyms
+
+This process of walking synonyms between different providers should always be done with care.
+Synonyms can arise for many reasons (see [@something] for discussion), and treating accepted names and known synonyms as equivalent or interchangable can quickly lead to innaccurate conclusions.
+
+An accepted name can have many synonyms:
 
@@
 Here we see that some accepted identifiers, such as `COL:43082445`, are known by as many as 456 different synonyms!
-`get_ids()` function seamlessly handles the resolution of all synonyms to their accepted name usage identifier, since we can unambiguously map from many-to-one.
+`get_ids()` function seamlessly handles the resolution of all these synonyms to their accepted name usage identifier when they can be mapped to a single accepted name.
 
@@
 Other functions in `taxadb` allow us to explore these names further.
-We have already seen this strategy above
 
@@
 
-For many applications, it is easier to work with the `by_*` class of functions, that first perform the relevant filtering join in the MonetDBLite database before returning the table in memory.
-This means that the `data.frame` objects returned by these functions behave as familiar objects.
-For instance, we can take the top 5 ambiguous names and see all the taxa to which those names may refer:
-
-```{r}
-top_5 <- most_ambiguous %>% 
-  head(5) %>% 
-  pull(scientificName)
-
-by_name(top_5, "col") %>%
-  select(scientificName, acceptedNameUsageID) %>% 
-  distinct() %>% 
-  mutate(acceptedNameUsage = get_names(acceptedNameUsageID, "col")) %>% 
-  head(5)
-```
-
-```{r}
-by_rank("Aves", rank="class", provider = "col") %>% head(5)
-```
-
 <!--
@@
 - the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
-- Catalogue of Life annual Species list
+- Catalogue of Life (COL) Annual Species list. (2018 version at present)
+- Global Biodiversity Information Facility Taxonomic Backbone [GBIF; @gbif].  
+- FishBase [@fishbase] taxonomic names
+- SeaLifeBase [@sealifebase] taxonomic names
+- Wikidata [@jorrit]
+- Open Tree Taxonomy, OTT
+- International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn
+
+Note that some of these providers are primarily syntheses of other existing naming providers.
+For instance, COL is based in part on over 130 other name databases (<http://www.catalogueoflife.org/col/info/databases>), including ITIS, FishBase, and the IUCN Redlist.
+However, these syntheses may lag behind the most recent versions of their member databases, which can create surprisingly large mismatches.
+Because `taxadb` provides access to the different providers using a common format, we can quickly explore this.
+For instance, here we count how many accepted names in ITIS can actually be found in COL (as either accepted names or known synonyms):
+
+```{r}
+# how many ITIS accepted species names are in COL?
+itis_in_col <-
+taxa_tbl("itis") %>%
+  filter(taxonomicStatus == "accepted", taxonRank == "species") %>%
+  select(scientificName) %>% 
+  left_join(taxa_tbl("col"), by="scientificName") %>% 
+  filter(is.na(acceptedNameUsageID)) %>% 
+  count() %>%
+  pull(n)
+
+# out of how many total accepted names in ITIS?
+itis_accepted <- taxa_tbl("itis") %>%
+  filter(taxonomicStatus == "accepted", taxonRank == "species") %>% 
+  count() %>% pull(n)
+```
+$`r prettyNum(itis_in_col, big.mark = ",")`$ of the $`r prettyNum(itis_accepted, big.mark = ",")`$ of *accepted names* in ITIS ($`r round(itis_in_col / itis_accepted * 100)`$%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.
+
+Consequently, just because the naming providers such as COL, OTT, and GBIF consume the names in ITIS, NCBI, or IUCN, this does not mean we can rely exclusively on these larger syntheses.
+
+<!--
+It is important to distinguish between two broad kinds of naming providers.
+Efforts such as ITIS, NCBI, FishBase, and SeaLifeBase are largely independent naming assemblies.
+GBIF and OTT are synthetic 'taxonomic backbones,' created by reconciling species names from a wide range of providers to create a taxonomic backbone utilized internally for their respective efforts.
+Catalogue of Life (COL)
+-->
 
@@
 
-The extent of mismatch between providers can be startling.
-Catalogue of Life uses ITIS data as one of it's many input sources, and yet we can quickly calculate how many of the (2018) ITIS accepted species names are not in the Catalogue of Life records:
-
-```{r}
-itis_col <-
-taxa_tbl("itis") %>%
-  filter(taxonomicStatus == "accepted", taxonRank == "species") %>%
-  select(scientificName) %>% 
-  left_join(taxa_tbl("col"), by="scientificName") %>% 
-  filter(is.na(acceptedNameUsageID)) %>% 
-  count() %>%
-  pull(n)
-
-itis_accepted <- taxa_tbl("itis") %>%
-  filter(taxonomicStatus == "accepted", taxonRank == "species") %>% 
-  count() %>% pull(n)
-```
-$`r itis_col`$ of the $`r itis_accepted`$ of *accepted names* in ITIS ($`r itis_col / itis_accepted * 100`$%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.
-
 \pagebreak

### 2019-04-09 paper/manuscript.Rmd
fix case in input col

@@
 library(taxadb)
-
+library(printr)
 pkgconfig::set_config("dplyr::na_matches" = "never")
@@
 ```{r}
-by_rank("Aves", rank="class", provider = "col") %>% head(5)
+by_rank("Aves", rank="class", provider = "col") %>%
+  ## some example columns
+  select(taxonID, taxonomicStatus, taxonRank, genus, specificEpithet) %>% 
+  head(5)
+
 ```
+
+We can inspect all columns:
+
+```{r}
+by_rank("Aves", rank="class", provider = "col") %>% names()
+```
+
+Note that each of these terms are defined by
 

### 2019-04-10 README.Rmd
a few minor fixes
- get_ids should have ignore_case
- get_names does not need ignore_case
- update docs & spellcheck readme

@@
 
-The goal of `taxadb` is to provide *fast*, *consistent* access to taxonomic data, supporting commont tasks such as resolving taxonomic names to identifiers, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
+The goal of `taxadb` is to provide *fast*, *consistent* access to taxonomic data, supporting common tasks such as resolving taxonomic names to identifiers, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
 These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records.
@@
 Note that some names cannot be resolved to an identifier.
-This can occur because of mis-spellings, non-standard formatting, or the use of a synonym not recognized by the naming provider.
+This can occur because of miss-spellings, non-standard formatting, or the use of a synonym not recognized by the naming provider.
 Names that cannot be uniquely resolved because they are known synonyms of multiple different species will also return `NA`.
@@
 ```
-However, unlike the `by_*` functions which return convienent in-memory tables, this is still a remote connection.
+However, unlike the `by_*` functions which return convenient in-memory tables, this is still a remote connection.
 This means that direct access using the `taxa_tbl()` function (or directly accessing the database connection using `td_connect()`) is more low-level and requires greater care.
-For instance, we cannot just add a `%>% mutate(acceptedNameUsage = get_names(acceptedNameUsageID))` to the above, becuase `get_names` does not work on a remote collection.
+For instance, we cannot just add a `%>% mutate(acceptedNameUsage = get_names(acceptedNameUsageID))` to the above, because `get_names` does not work on a remote collection.
 Instead, we would first need to use a `collect()` to pull the summary table into memory.

### 2019-04-23 paper/manuscript.Rmd
Add to the by_ section, comment on unresolved names.

@@
 - The type of query is highly limited by the API design.
-For instance, it is impossible usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?"
+For instance, it is usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?"
 - Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
@@
 
-- database functions  `td_create()`, `td_connect()` and `taxa_tbl()`
 - queries that return vectors: `get_ids()` and it's complement, `get_names()`,
 - queries that filter the underlying taxonomic data frames: `by_name()`, `by_rank()`, `by_id()`, and `by_common()`,
+- database functions  `td_create()`, `td_connect()` and `taxa_tbl()`,
 - and helper utilities, such as `clean_names()`.
@@
 Unmatched names may indicate an error in data entry or otherwise warrant further investigation.
-Taxon identifiers area also easily resolved to the original authority (scientific publication) establishing the name.
+Taxon identifiers are also easily resolved to the original authority (scientific publication) establishing the name.
 (The historical practice of appending an author and year to a scientific name, e.g.
@@
 These issues are best illustrated using a minimal example.
-We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example consise enough for for visual presentation we will focus on a subset involving just 10 species:
+We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species:
 
@@
 
+KN: I don't know if this section needs to be in here, especially since `synonym()` doesn't fit cleanly in the four families of functions described at the beginning (maybe it's a helper function?).
+Can instead use the unresolved names discussion to segue into `by_` functions like in the tutorial.
+Something along the lines of "A name unresolved with `get_ids` could be due to misspelling, a missing synonym in the provider database OR a name resolving to multiple synonyms - 'by_' gives you access to all the synonym options." When it's laid out like that it also makes 'synonym()' seem less necessary.
+
 Notice that one species name could not be resolved to a COL identifier.
@@
 ## Additional taxonomic information
+
+A name unresolved by 'get_ids' could also be the result of a name matching to multiple accepted names, as in the case of a species that has been split.
+By design, these cases are left to be resolved by the researcher using the 'by_' functions to filter underlying taxonomic tables for additional information.
+
+```{r}
+#by_name() w/multiple matches example.
+```
+'by_' functions can also be used to filter by common name, rank, etc...
 

### 2019-05-14 R/fuzzy_filter.R
include helper fns around fuzzy_filter

@@ -110,8 +110,8 @@ filter_like <- function(db_tbl, input, pattern){
 #' }
 #' @inheritParams fuzzy_filter
 name_contains <- function(name,
-                          provider,
-                          db = td_connect,
+                          provider = "itis",
+                          db = td_connect(),
                           ignore_case = TRUE){
 
   fuzzy_filter(name,
@@ -133,9 +133,11 @@ name_contains <- function(name,
 #'   }
 #' name_contains("Homo ")
 #' }
+#' @inheritParams fuzzy_filter
+#' @export
 name_starts_with <- function(name,
                              provider,
-                             db = td_connect,
+                             db = td_connect(),
                              ignore_case = TRUE){
 
   fuzzy_filter(name,
@@ -145,6 +147,63 @@ name_starts_with <- function(name,
                db = db,
                ignore_case = ignore_case)
 }
+
+
+
+#' common name starts with
+#'
+#' @examples
+#' \donttest{
+#'   \dontshow{
+#'    ## All examples use a temporary directory
+#'    Sys.setenv(TAXADB_HOME=tempdir())
+#'   }
+#' common_starts_with("monkey")
+#' }
+#' @inheritParams fuzzy_filter
+#' @export
+common_starts_with <- function(name,
+                             provider = "itis",
+                             db = td_connect(),
+                             ignore_case = TRUE){
+
+  fuzzy_filter(name,
+               by = "vernacularName",
+               provider = provider,
+               match = "starts_with",
+               db = db,
+               ignore_case = ignore_case)
+}
+
+
+#' common name starts with
+#'
+#' @examples
+#' \donttest{
+#'   \dontshow{
+#'    ## All examples use a temporary directory
+#'    Sys.setenv(TAXADB_HOME=tempdir())
+#'   }
+#' common_contains("monkey")
+#' }
+#' @inheritParams fuzzy_filter
+#' @export
+common_contains <- function(name,
+                            provider = "itis",
+                            db = td_connect(),
+                            ignore_case = TRUE){
+
+  fuzzy_filter(name,
+               by = "vernacularName",
+               provider = provider,
+               match = "contains",
+               db = db,
+               ignore_case = ignore_case)
+}
+
+
+
+
 
 
 

### 2019-05-14 R/td_create.R

@@ -29,7 +29,7 @@
 #' @return path where database has been installed (invisibly)
 #' @export
 #' @importFrom utils download.file
-#' @importFrom DBI dbConnect dbDisconnect
+#' @importFrom DBI dbConnect dbDisconnect dbListTables
 #' @importFrom arkdb unark streamable_readr_tsv
 #' @importFrom MonetDBLite MonetDBLite
 #' @importFrom readr cols

### 2019-05-15 R/fuzzy_filter.R
Fuzzy filter (#56)
* export name_contains and starts_with

* include helper fns around fuzzy_filter

* update docs

@@ -96,9 +96,22 @@ filter_like <- function(db_tbl, input, pattern){
 ## Consider creating functions that are explicitly named to create more semantic
 ## code, rather than relying on setting the behavior in the `by` and `match`
 ## arguments to `fuzzy_filter`, e.g. :
+
+#' return all taxa in which scientific name contains the text provided
+#'
+#' @export
+#' @examples
+#' \donttest{
+#'   \dontshow{
+#'    ## All examples use a temporary directory
+#'    Sys.setenv(TAXADB_HOME=tempdir())
+#'   }
+#' name_contains("Homo ")
+#' }
+#' @inheritParams fuzzy_filter
 name_contains <- function(name,
-                          provider,
-                          db = td_connect,
+                          provider = "itis",
+                          db = td_connect(),
                           ignore_case = TRUE){
 
   fuzzy_filter(name,
@@ -109,9 +122,22 @@ name_contains <- function(name,
                ignore_case = ignore_case)
 }
 
+
+#' scientific name starts with
+#'
+#' @examples
+#' \donttest{
+#'   \dontshow{
+#'    ## All examples use a temporary directory
+#'    Sys.setenv(TAXADB_HOME=tempdir())
+#'   }
+#' name_contains("Homo ")
+#' }
+#' @inheritParams fuzzy_filter
+#' @export
 name_starts_with <- function(name,
                              provider,
-                             db = td_connect,
+                             db = td_connect(),
                              ignore_case = TRUE){
 
   fuzzy_filter(name,
@@ -121,6 +147,63 @@ name_starts_with <- function(name,
                db = db,
                ignore_case = ignore_case)
 }
+
+
+
+#' common name starts with
+#'
+#' @examples
+#' \donttest{
+#'   \dontshow{
+#'    ## All examples use a temporary directory
+#'    Sys.setenv(TAXADB_HOME=tempdir())
+#'   }
+#' common_starts_with("monkey")
+#' }
+#' @inheritParams fuzzy_filter
+#' @export
+common_starts_with <- function(name,
+                             provider = "itis",
+                             db = td_connect(),
+                             ignore_case = TRUE){
+
+  fuzzy_filter(name,
+               by = "vernacularName",
+               provider = provider,
+               match = "starts_with",
+               db = db,
+               ignore_case = ignore_case)
+}
+
+
+#' common name starts with
+#'
+#' @examples
+#' \donttest{
+#'   \dontshow{
[…]

### 2019-05-15 notebook/globi.R

@@ -1,8 +1,8 @@
-# FROM: Poelen, Jorrit H. (2018). Global Biotic Interactions: Taxon Graph (Version 0.3.2) [Data set]. 
+# FROM: Poelen, Jorrit H. (2018). Global Biotic Interactions: Taxon Graph (Version 0.3.2) [Data set].
 # Zenodo. http://doi.org/10.5281/zenodo.1250572
 
 library(tidyverse)
-
+library(MonetDBLite)
 expect_none <- function(df){ testthat::expect_equal(dim(df)[[1]], 0) }
 
 
@@ -32,8 +32,8 @@ if(test){
   #taxonCache %>% filter(!grepl(":", id)) %>% expect_none()
   ## Some tests
   #taxonCache %>% filter(grepl(":", path)) %>% expect_none()  ## some paths are ids
-  taxonCache %>% filter(grepl("\\s", id))  %>% expect_none()  
-    
+  taxonCache %>% filter(grepl("\\s", id))  %>% expect_none()
+
   pattern = "\\s*\\|\\s*"
   path_pipes <- taxonCache %>% purrr::transpose() %>%
     map_int( ~length(str_split(.x$path, pattern)[[1]]))
@@ -54,8 +54,8 @@ if(test){
 ## taxonCache <- taxonCache[-trouble,]
 
 
-longform <- function(row, pattern = "\\s*\\|\\s*"){ 
-  row_as_df <- 
+longform <- function(row, pattern = "\\s*\\|\\s*"){
+  row_as_df <-
     data_frame(id = row$id,
                name = row$name,
                rank = row$rank,
@@ -65,20 +65,20 @@ longform <- function(row, pattern = "\\s*\\|\\s*"){
                commonNames = row$commonNames,
                externalUrl = row$externalUrl,
                thumbnailUrl = row$thumbnailUrl)
-               
+
 }
 
 
 # 3052673 rows.  3,052,673
 
 system.time({
-taxa <- taxonCache %>% 
-  transpose() %>% 
-  map_dfr(longform) %>% 
-  distinct() 
+taxa <- taxonCache %>%
+  transpose() %>%
+  map_dfr(longform) %>%
+  distinct()
 })
 
-## FIXME 
+## FIXME
 ## - [ ] standardize case
 ## - [x] standardize rank names
 
@@ -87,35 +87,35 @@ taxon_rank_list <- read_tsv(paste0("https://raw.githubusercontent.com/",
[…]

### 2019-05-15 tests/testthat/test-fuzzy_filter.R

@@ -16,3 +16,24 @@ test_that("we can fuzzy match scientific and common names", {
 
 
 })
+
+test_that("we can fuzzy match scientific and common names", {
+
+  name <- c("woodpecker", "monkey")
+  df <- common_contains(name, "itis")
+  expect_is(df, "data.frame")
+  expect_gt(dim(df)[1], 1)
+  df <- common_starts_with(name, "itis")
+  expect_is(df, "data.frame")
+
+
+  df <- name_starts_with("Homo ", "itis",
+                     ignore_case = FALSE)
+  expect_is(df, "data.frame")
+  expect_gt(dim(df)[1], 1)
+  df <- name_contains("Homo", "itis")
+  expect_is(df, "data.frame")
+  expect_gt(dim(df)[1], 1)
+
+
+})

### 2019-06-28 paper/manuscript.Rmd

@@
     affiliation: ucb
-  - name: "Jorrit Poelen"
-    affiliation: jorritt
+#  - name: "Jorrit Poelen"
+#    affiliation: jorritt
   - name: "Scott Chamberlain"
@@
 library(printr)
+
+## see https://blog.rstudio.com/2017/06/13/dplyr-0-7-0/, 
+## https://community.rstudio.com/t/why-does-na-match-na-when-joining-two-dataframes/28785/3
 pkgconfig::set_config("dplyr::na_matches" = "never")
@@
 
-After installing the `taxadb` R package, users can create local copies of data from any of the providers using the `td_create()` function and specifying the provider abbreviation (see Table 1), or `all` to install all available providers:
+<!--
+### Installation
+
+`taxadb` can be installed from GitHub using the `remotes` R package:
+
+```{r eval=FALSE}
+remotes::install_github("cboettig/taxadb")
+```
+-->
 
@@
 
+`taxadb` functions will automatically set up a local, persistent database if one has not yet been created.
+
+```{r}
+get_ids("Gadus morhua")
+```
+
+This one-time setup will download, uncompress, and import the data into persistant storage (using the appropriate location specified by the operating sytstem [see @rappdirs], or configured using the environmental variable `TAXADB_HOME`).
+The example above searches for names in ITIS, the default provider, which can be configured using the `provider` argument.
+Any future function calls to this function or any other function using data from the same names provider will be able to access this data rapidly without the need for processing or an internet connection.
+
+Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 1), or simply using `all` to install all available providers:
+
 ```{r eval=FALSE, message=FALSE}
@@
 
+`taxadb` functions like `get_ids()` and `td_create()` take an optional argument, `db`, to an external database connection.
+`taxadb` will work with most DBI-compliant databases such as MySQL or Postgres [see @DBI], but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.
+These latter options are also much easier for most users, since each can be installed directly as an R package.
+`taxadb` will default to the fastest available option.
+`taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the large names providers.
+
+<!--
+`taxadb` does not require the use of an external database system, but will automatically make use of the fastest database server available.
+`taxadb` can also use any DBI compliant database accessible from R, such as MySQL or Postgres [see @DBI], though the default `duckdb` or `MonetDBLite` will be both significantly faster and easier to use.
+-->
+
 <!--  Next 2 paragraphs should possibly be left to a manual, but are likely to be overlooked aspects there... -->
-This one-time download will download and import local copies of the provider data.
-By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage, and data is stored in the appropriate location specified for applications by the operating sytstem [@rappdirs].
-Users can customize this location, or opt for any alternative relational database backend by providing alternative arguments, though the default MonetDBLite system can be an order of magnitude faster than popular alternatives such as Postgres [@monetdblite].
-`taxadb` will automatically handle opening, caching, and closing the MonetDBLite connection, but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time.
+By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage.
+`taxadb` will automatically handle opening, caching, and closing the database connection.
+
+<!-- 
+now sidestepped
+
+but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time.
 Users who need concurrent access from multiple sessions may wish to use a standalone MonetDB server or other database server instead.
+-->
 
-All `taxadb` functions can also operate in one-time-use mode without first installing a standalone server.
-If a function requests a provider which has not been installed, the relevant table can be downloaded and read directly into memory.
-This requires sufficient memory be available to hold the entire table (typically 1-2 GB) and will only be cached for the R session.
-This can be useful in special cases such as execution on remote servers where this will be faster than importing the data locally, or to allow other packages to use methods such as `get_ids()` internally without assuming users will have first created a local copy with `td_create()`.
-Specify `db=NULL` on any `taxadb` function to force this in-memory dispatch.
+It is not necessary to call `td_create()`.  Specify `db=NULL` on any `taxadb` function to force this in-memory dispatch.
 
@@
 
-```{r}
+```{r message=FALSE}
 trait_data <- read_tsv(system.file("extdata", "trait_data.tsv", package="taxadb"))

### 2019-08-12 paper/manuscript.Rmd
Change some order of ideas, remove old comments.

@@
   - name: "Carl Boettiger"
-    affiliation: ucb, 1
+    affiliation: ucb, ropensci
 address:
@@
     address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
-  - code: jorrit
-    address: "Independent consultant, Oakland, CA, USA"
+# - code: jorrit
+#   address: "Independent consultant, Oakland, CA, USA"
   - code: ropensci
@@
 
-Instead of binding existing web APIs, `taxadb` is built around a set compressed text files following a consistent, standardized layout or schema (discussed below).
+Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files following a consistent, standardized layout or schema (discussed below).
 These files are automatically downloaded and imported and stored on a local database by `taxadb`.
@@
 -->
-
-It is not necessary to call `td_create()`.  Specify `db=NULL` on any `taxadb` function to force this in-memory dispatch.
 
@@
 
-KN: I don't know if this section needs to be in here, especially since `synonym()` doesn't fit cleanly in the four families of functions described at the beginning (maybe it's a helper function?).
-Can instead use the unresolved names discussion to segue into `by_` functions like in the tutorial.
-Something along the lines of "A name unresolved with `get_ids` could be due to misspelling, a missing synonym in the provider database OR a name resolving to multiple synonyms - 'by_' gives you access to all the synonym options." When it's laid out like that it also makes 'synonym()' seem less necessary.
-
 Notice that one species name could not be resolved to a COL identifier.
@@
 
-This query uses the `taxa_tbl()` function to directly access the full taxonomic record in the MonetDBLite database.
-This approach is fast and powerful, but note that we are restricted to using `dplyr` functions, which can be directly translated into SQL, for such queries to work successfully.
-Though the result looks like a data frame, most base R operations will not work as expected:
-
-```{r}
-most_synonyms$n
-```
-Such queries are also not possible through the API or web interface to Catalogue of Life.
-To use custom R functions or functions from other packages we must first call the `dplyr` function `collect()` to return the table into R's memory.
-This should be done with care as an unfiltered table may require several gigabytes of memory to load.
-
 Some synonyms may also be associated with more than one unique identifier.
@@
 The `get_ids()` function will thus return `NA` for any name that matches more than one identifier, just as it will return `NA` for a name that does not match any identifier at all.
-Other functions in `taxadb` allow us to explore these names further.
 
-Having access to a local database makes it easy to explore such issues as "identify all synonyms that resolve to more than one accepted scientific name ID", like so:
+Other functions in `taxadb` allow us to explore these names further, for example having access to a local database makes it easy to explore such issues as "identify all synonyms that resolve to more than one accepted scientific name ID", like so:
 
@@
 Here we see the name *Mabuya bistriata* is a synonym which may be associated with no fewer than 32 different accepted scientific names!
+
+This query uses the `taxa_tbl()` function to directly access the full taxonomic record in the MonetDBLite database.
+This approach is fast and powerful, but note that we are restricted to using `dplyr` functions, which can be directly translated into SQL, for such queries to work successfully.
+Though the result looks like a data frame, most base R operations will not work as expected:
+
+```{r}
+most_synonyms$n
+```
+
+Such queries are also not possible through the API or web interface to Catalogue of Life.
+To use custom R functions or functions from other packages we must first call the `dplyr` function `collect()` to return the table into R's memory.
+This should be done with care as an unfiltered table may require several gigabytes of memory to load.
 

### 2019-08-22 paper/manuscript.Rmd
minor unresolved names changes

@@
 Notice that one species name could not be resolved to a COL identifier.
-This could indicate a data entry problem such as a misspelling of a name, but failure to resolve properly specified scientific names is not uncommon, as even large scale efforts like the Catalogue of Life are far from a comprehensive list of all published scientific names.
+This could indicate a data entry problem such as a misspelling of a name or the inclusion of a citation or year, which may be addressed by `clean_names()`.
+However, failure to resolve properly specified scientific names is not uncommon, as even large scale efforts like the Catalogue of Life are far from a comprehensive list of all published scientific names.
 

### 2019-08-28 paper/manuscript.Rmd
Pair down, move synonym discussions to notebook/synonym.rmd

--- 2019-08-12 paper/manuscript.Rmd
@@
 
-Notice that one species name could not be resolved to a COL identifier.
-This could indicate a data entry problem such as a misspelling of a name, but failure to resolve properly specified scientific names is not uncommon, as even large scale efforts like the Catalogue of Life are far from a comprehensive list of all published scientific names.
+potential ways for resolving unidentified names
+* multiple hits from `get_ids()` check duplicates, use `by_`
+* typo
+* formatting issues (be careful about false matches), `clean_names()`
+* ambiguous resolution, explore with `synonyms()` (these are the same concept with different names or vice versa)
 
-Perhaps we can resolve this name against a different authority?
-As it has an IUCN classification already, the IUCN naming provider seems a natural place to start.
-We can use the `synonyms()` table to attempt to identify any other scientific names by which this species is known:
+The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories"
 
-```{r}
-synonyms("Scleroptila gutturalis", "iucn")
-```
-
-Note that one of these synonyms is recognized in COL, and in fact matches a COL identifier of *Francolinus levalliantoides*, 376.69 grams, in our data set.
-
-```{r}
-synonyms("Scleroptila gutturalis", "iucn") %>% 
-  pull(synonym) %>% 
-  get_ids("col")
-```
-
-## Additional taxonomic information
-
-A name unresolved by 'get_ids' could also be the result of a name matching to multiple accepted names, as in the case of a species that has been split.
-By design, these cases are left to be resolved by the researcher using the 'by_' functions to filter underlying taxonomic tables for additional information.
-
-```{r}
-#by_name() w/multiple matches example.
-```
+## `by_` functions for access to underlying tables
 'by_' functions can also be used to filter by common name, rank, etc...
@@
 
-Note that each of these terms are defined by
+`by_common` (potentially in a box)
 
-## Caution using synonyms
-
-This process of walking synonyms between different providers should always be done with care.
-Synonyms can arise for many reasons (see [@something] for discussion), and treating accepted names and known synonyms as equivalent or interchangable can quickly lead to innaccurate conclusions.
-
-An accepted name can have many synonyms:
-
-```{r}
-most_synonyms <- taxa_tbl("col") %>% count(acceptedNameUsageID, sort = TRUE)
-most_synonyms
-```
-
-Here we see that some accepted identifiers, such as `COL:43082445`, are known by as many as 456 different synonyms!
-`get_ids()` function seamlessly handles the resolution of all these synonyms to their accepted name usage identifier when they can be mapped to a single accepted name.
-
-Some synonyms may also be associated with more than one unique identifier.
-This situation is more difficult, as it is impossible to know which accepted name should be used for such synonyms without more context.
-The `get_ids()` function will thus return `NA` for any name that matches more than one identifier, just as it will return `NA` for a name that does not match any identifier at all.
-
-Other functions in `taxadb` allow us to explore these names further, for example having access to a local database makes it easy to explore such issues as "identify all synonyms that resolve to more than one accepted scientific name ID", like so:
-
-```{r}
-most_ambiguous <-
-taxa_tbl("col") %>% 
-  filter(taxonomicStatus != "accepted") %>% 
-  select(scientificName, acceptedNameUsageID) %>% 
-  distinct() %>% 
-  count(scientificName, sort=TRUE) %>%
-  filter(n > 1)
-most_ambiguous
-```
-Here we see the name *Mabuya bistriata* is a synonym which may be associated with no fewer than 32 different accepted scientific names!
-
-This query uses the `taxa_tbl()` function to directly access the full taxonomic record in the MonetDBLite database.
-This approach is fast and powerful, but note that we are restricted to using `dplyr` functions, which can be directly translated into SQL, for such queries to work successfully.
-Though the result looks like a data frame, most base R operations will not work as expected:
-
-```{r}
-most_synonyms$n
-```
-
-Such queries are also not possible through the API or web interface to Catalogue of Life.
-To use custom R functions or functions from other packages we must first call the `dplyr` function `collect()` to return the table into R's memory.
-This should be done with care as an unfiltered table may require several gigabytes of memory to load.
-
-<!--
-The motivation for an API-based approach is reasonable: (1) Web APIs are a ubiquitous standard of data exchange (2) make very minimal assumptions about the language, software, or hardware avaialable to the user, (3) users access a single centrally managed database that is maintained and upgraded by the provider, and (4) the provider can easily control access to the data, measure usage statistics, and enforce rate limits.
-Many existing tools provide bindings of these APIs as functions to a specific language, including `taxize`, `tits`, `rotl`, `worms`, `redlist` and `wikitaxa` [@Chamberlain2013; @itis; @rotl; @worrms; @wikitaxa]
--->
-
-Darwin Core schema.
+* point to whole database stuff (see vingette) - queries are possible that aren't with an API
+* software design - faster access, reproducible queries
+* Darwin Core schema citation
 
 ### Supported providers
+
+Table version of below - citation, number of unique identifier
 
@@
 
-Note that some of these providers are primarily syntheses of other existing naming providers.
-For instance, COL is based in part on over 130 other name databases (<http://www.catalogueoflife.org/col/info/databases>), including ITIS, FishBase, and the IUCN Redlist.
-However, these syntheses may lag behind the most recent versions of their member databases, which can create surprisingly large mismatches.
-Because `taxadb` provides access to the different providers using a common format, we can quickly explore this.
-For instance, here we count how many accepted names in ITIS can actually be found in COL (as either accepted names or known synonyms):
-
-```{r}
-# how many ITIS accepted species names are in COL?
-itis_in_col <-
-taxa_tbl("itis") %>%
-  filter(taxonomicStatus == "accepted", taxonRank == "species") %>%
-  select(scientificName) %>% 
-  left_join(taxa_tbl("col"), by="scientificName") %>% 
-  filter(is.na(acceptedNameUsageID)) %>% 
-  count() %>%
-  pull(n)
-
-# out of how many total accepted names in ITIS?
-itis_accepted <- taxa_tbl("itis") %>%
-  filter(taxonomicStatus == "accepted", taxonRank == "species") %>% 
-  count() %>% pull(n)
-```
-$`r prettyNum(itis_in_col, big.mark = ",")`$ of the $`r prettyNum(itis_accepted, big.mark = ",")`$ of *accepted names* in ITIS ($`r round(itis_in_col / itis_accepted * 100)`$%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.
-
-Consequently, just because the naming providers such as COL, OTT, and GBIF consume the names in ITIS, NCBI, or IUCN, this does not mean we can rely exclusively on these larger syntheses.
-
-<!--
-It is important to distinguish between two broad kinds of naming providers.
-Efforts such as ITIS, NCBI, FishBase, and SeaLifeBase are largely independent naming assemblies.
-GBIF and OTT are synthetic 'taxonomic backbones,' created by reconciling species names from a wide range of providers to create a taxonomic backbone utilized internally for their respective efforts.
-Catalogue of Life (COL)
--->
-
-<!--
-comment that taxonomic backbones aren't really the same thing:
-In addition to these projects, large biodiversity informatics efforts such as the Global Biodiversity Information Facility [GBIF; @gbif], publish the 'taxonomic backbone' -->
-
 ## Discussion
 
-- Resolve names to identifiers
-- Resolve synonyms to accepted names
-- Associate names with taxonomic rank and heirarchical classification
-
-discuss use Scientific names vs identifiers.  taxonConcepts?
+We've shown identifiers are useful in an a whole bunch of applications. People haven't done it, becuase it's time consuming - but now its fast!
 

### 2019-08-28 paper/manuscript.Rmd
Flesh out unresovled names section.

@@
 author and date) establishing a scientific name.
-The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013] has become a popular way R users to interact with naming providers and name resolution services.
+The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013] has become a popular way for R users to interact with naming providers and name resolution services.
 `taxize` implements bindings to the web APIs (Application Programming Interface) hosted by many popular taxonomic name providers.
@@
 
-potential ways for resolving unidentified names
-* multiple hits from `get_ids()` check duplicates, use `by_`
-* typo
-* formatting issues (be careful about false matches), `clean_names()`
-* ambiguous resolution, explore with `synonyms()` (these are the same concept with different names or vice versa)
+`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons.
+First, a name may match to multiple accepted names, as in the case of a species that has been split.
+By design, these cases are left to be resolved by the researcher using the 'by_' functions to filter underlying taxonomic tables for additional information.
+A name may also be unresolved due to typos or improper formatting.
+`clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g.
+`Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithet such as `Colaptes auratus cafer` that prevent matches to the bionmial name.
+These modifications are not appropriate in all settings and should be used with care (Language from vignette).
 
-The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories"
+Names may also have an ambiguous resolution *discussion about taxonomic concepts here*.
+The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
 

### 2019-08-28 paper/manuscript.Rmd
Flesh out by_ function section. Add common names box.

@@
 ## `by_` functions for access to underlying tables
-'by_' functions can also be used to filter by common name, rank, etc...
+
+Underlying data frames can be accessed through the family of `by_` functions, which filter by certain attributes such as scientific name, id, common name, and rank.
+These functions allow us to ask general questions such as, how many bird species are there?
 
@@
 by_rank("Aves", rank="class", provider = "col") %>%
-  ## some example columns
-  select(taxonID, taxonomicStatus, taxonRank, genus, specificEpithet) %>% 
-  head(5)
-
+  pull(taxonID) %>%
+  n_distinct()
 ```
 
-We can inspect all columns:
+As well getting a detailed look at specific species or ids. For example, why can't `get_ids` resolve a seemingly common species?
 
 ```{r}
-by_rank("Aves", rank="class", provider = "col") %>% names()
+## example of sci name matching to more than one id.
 ```
 
-`by_common` (potentially in a box)
+We see that insert_name matches two accepted names which the user will have to choose between.
 
-* point to whole database stuff (see vingette) - queries are possible that aren't with an API
+The full taxonomic record in the MonetDBLite database can also be directly accessed by `taxa_tbl()`, allowing for whole-database queries that are not possible through the API or web interface of many providers (see Vignette for further details).
+
+\dummy{\Begin{tcolorbox}[title= Box 2: Common Names, lower separated=false]}
+
+`taxizedb` can also resolve common names to their identifier by mapping common name to the accepted scientific name.
+Common names come with many of the same problems as scientific names but worse, e.g.
+matching to more than one accepted name is even more frequent, non-standardized formatting (like capitalization) and spelling etc.
+Common names are accessed via `by_common` which takes a vector of common names.
+The user can then resolve discrepencies.
+
+*talk about how we deal with common names and `by_common`*
+
+\End{tcolorbox}
+
 * software design - faster access, reproducible queries

### 2019-08-28 paper/manuscript.Rmd
add some details to the intro + darwin core citation

@@
 
-Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files following a consistent, standardized layout or schema (discussed below).
+Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files following a consistent, standardized layout or schema following the Darwin Core Standard [@Wieczorek2012].
 These files are automatically downloaded and imported and stored on a local database by `taxadb`.
@@
 Thus it does not take many taxa before transferring the entire data set to query locally is more efficient.
-Moreover, this local copy can be cached on the user's machine, requiring only the one-time setup and enabling offline use.
+Moreover, this local copy can be cached on the user's machine, requiring only the one-time setup, and enabling offline use and reproducible queries.
 
@@
 
-* software design - faster access, reproducible queries
-* Darwin Core schema citation
-
 ### Supported providers

### 2019-09-24 paper/manuscript.Rmd
Create provider table.

@@
 
-Names may also have an ambiguous resolution *discussion about taxonomic concepts here*.
+Names may also have an ambiguous resolution leading to resolution by a different provider than the one specified, either as an accepted name or a synonym.
+Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider.
 The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
@@
 
+```{r cache = FALSE, echo = FALSE}
+tibble(provider = c("Integrated Taxonomic Information System", 
+                                 "National Center for Biological Information's Taxonomy database",
+                                 "Catalogue of Life",
+                                 "Global Biodiversity Information Facility Taxonomic Backbone",
+                                 "FishBase",
+                                 #"SeaLifeBase",
+                                 "Wikidata", "Open Tree Taxonomy",
+                                 "International Union for Conservation of Nature and Natural Resources"),
+                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
+                                     "wd", "ott", "iucn"),
+                    total_identifiers = map(abbreviation, 
+                                            ~taxa_tbl(.x) %>% 
+                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct() %>% unlist())
+) %>% printtable()
+
+```
+
 ## Discussion

### 2019-09-26 paper/manuscript.Rmd

--- 2019-08-28 paper/manuscript.Rmd
@@
 `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g.
-`Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithet such as `Colaptes auratus cafer` that prevent matches to the bionmial name.
+`Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name.
 These modifications are not appropriate in all settings and should be used with care (Language from vignette).
 
-Names may also have an ambiguous resolution *discussion about taxonomic concepts here*.
+Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym.
+Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider.
 The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
@@
 
-Underlying data frames can be accessed through the family of `by_` functions, which filter by certain attributes such as scientific name, id, common name, and rank.
+Underlying data tables can be accessed through the family of `by_` functions, which filter by certain attributes such as scientific name, id, common name, and rank.
 These functions allow us to ask general questions such as, how many bird species are there?
@@
 ## example of sci name matching to more than one id.
+by_name("Zosterops rendovae") %>% 
+  select(1:5)
+
 ```
 
-We see that insert_name matches two accepted names which the user will have to choose between.
+We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between.
 
@@
 `taxizedb` can also resolve common names to their identifier by mapping common name to the accepted scientific name.
-Common names come with many of the same problems as scientific names but worse, e.g.
-matching to more than one accepted name is even more frequent, non-standardized formatting (like capitalization) and spelling etc.
+Common names come with many of the same problems as scientific names but often more frequent, e.g.
+matching to more than one accepted name, non-standardized formatting (like capitalization) and spelling etc.
 Common names are accessed via `by_common` which takes a vector of common names.
 The user can then resolve discrepencies.
-
-*talk about how we deal with common names and `by_common`*
 
@@
 
+<!--
+What other aspects should be included in the table? Citations?
+-->
+
+```{r cache = FALSE, echo = FALSE}
+tibble(provider = c("Integrated Taxonomic Information System", 
+                                 "National Center for Biological Information's Taxonomy database",
+                                 "Catalogue of Life",
+                                 "Global Biodiversity Information Facility Taxonomic Backbone",
+                                 "FishBase",
+                                 #"SeaLifeBase",
+                                 "Wikidata", "Open Tree Taxonomy",
+                                 "International Union for Conservation of Nature and Natural Resources"),
+                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
+                                     "wd", "ott", "iucn"),
+                    total_identifiers = map(abbreviation, 
+                                            ~taxa_tbl(.x) %>% 
+                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
+) %>% printtable()
+
+```
+
 ## Discussion
 
-We've shown identifiers are useful in an a whole bunch of applications. People haven't done it, becuase it's time consuming - but now its fast!
+Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications.
+Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary fo resolving even moderately sized data sets.
+`taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers.
+`taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
 
-- `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.
-Instead, it fills an important gap between existing tools and typical research patterns.
-In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
-
-- `taxadb` works from versioned snapshots of the data providers.
-
-- All `taxadb` functions are specific to a provider, while operating consistently across different providers.
-
-- `taxadb` does not attempt to present any unified taxonomic backbone or synthesis from existing naming providers, or make any assertions, numerical scores or other inferences about the data or matches to the data -- it is merely a tool for accessing this infromation.
-This contrasts from other approaches such as the Global Names Resolver, which
-
-There is an important objection to this approach that must also be addressed.
-
-@Franz2018
-
->>  no self-respecting biodiversity researcher would or should trust aggregated data blindly, that indeed careful data cleaning is almost always necessary and expected to render the downloaded data fit for purpose, and that therefore aggregator services need to be understood mainly or merely as data discovery tools.
-
-> We reject this deflationary view for four reasons.
-For one, aggregators frequently blur the lines between advertizing their services just as a data discovery tool or as a more powerful data signal tool.
-Second, the biases inherent in using unitary backbones remain in place even if users are only interested in discovering all relevant data for their research purpose.
-If the backbone-based data record modulations are not easily retrievable through primary on-line interfaces, then users are significantly constrained in their ability to design search queries with high rates of precision and recall (81).
-Third, let us assume that labor-intensive off-line data quality review and correction efforts are indeed the norm, prior to publishing.
-Then why must the fruits of these efforts remain outside of the aggregator’s environment?
-Why can they not immediately flow back into the same aggregation domain, while recording the provenance of expert changes?
-In other words, if the workflow of rendering data fit for purpose flows only in one direction, i.e.
-from the on-line aggregate to the off-line quality review and publication, then our criticism of the design stands.
-
-The inability to make automate use and manipulation
-
-(Figure 1: schematic example of data join?)
-
-Assemble data across multiple studies
-
-How many species of Chameleons are known? There are over
-
-```{r}
-taxa_tbl("col") %>% count(taxonomicStatus, sort = TRUE)
-```
+`taxadb` establishes a local database of provider allowing for consistent and reproducible queries. The set of functions operate consistently accross providers returning data in a standardized format.
 

### 2019-10-03 paper/manuscript.Rmd
Add citations, render them in table.

@@
 library(printr)
+library(knitr)
 
@@
     kableExtra::kable("latex", booktabs=T) %>%
-    kable_styling(full_width = F)
+    kableExtra::kable_styling(full_width = F)
 }
@@
 - queries that return vectors: `get_ids()` and it's complement, `get_names()`,
-- queries that filter the underlying taxonomic data frames: `by_name()`, `by_rank()`, `by_id()`, and `by_common()`,
+- queries that filter the underlying taxonomic data frames: `filter_name()`, `filter_rank()`, `filter_id()`, and `filter_common()`,
 - database functions  `td_create()`, `td_connect()` and `taxa_tbl()`,
@@
 First, a name may match to multiple accepted names, as in the case of a species that has been split.
-By design, these cases are left to be resolved by the researcher using the 'by_' functions to filter underlying taxonomic tables for additional information.
+By design, these cases are left to be resolved by the researcher using the 'filter_' functions to filter underlying taxonomic tables for additional information.
 A name may also be unresolved due to typos or improper formatting.
@@
 
-## `by_` functions for access to underlying tables
+## `filter_` functions for access to underlying tables
 
-Underlying data tables can be accessed through the family of `by_` functions, which filter by certain attributes such as scientific name, id, common name, and rank.
+Underlying data tables can be accessed through the family of `filter_` functions, which filter by certain attributes such as scientific name, id, common name, and rank.
 These functions allow us to ask general questions such as, how many bird species are there?
@@
 ```{r}
-by_rank("Aves", rank="class", provider = "col") %>%
+filter_rank("Aves", rank="class", provider = "col") %>%
   pull(taxonID) %>%
@@
 ## example of sci name matching to more than one id.
-by_name("Zosterops rendovae") %>% 
+filter_name("Zosterops rendovae") %>% 
   select(1:5)
@@
 matching to more than one accepted name, non-standardized formatting (like capitalization) and spelling etc.
-Common names are accessed via `by_common` which takes a vector of common names.
+Common names are accessed via `filter_common` which takes a vector of common names.
 The user can then resolve discrepencies.
@@
 
-```{r cache = FALSE, echo = FALSE}
-tibble(provider = c("Integrated Taxonomic Information System", 
-                                 "National Center for Biological Information's Taxonomy database",
-                                 "Catalogue of Life",
-                                 "Global Biodiversity Information Facility Taxonomic Backbone",
-                                 "FishBase",
+(ref:itis) Integrated Taxonomic Information System [@itis]
+
+```{r echo = FALSE, results = "asis"}
+tibble(provider = c("Integrated Taxonomic Information System, @itis",
+                                 "National Center for Biological Information's Taxonomy database, @ncbi",
+                                 "Catalogue of Life, @col",
+                                 "Global Biodiversity Information Facility Taxonomic Backbone, @gbif",
+                                 "FishBase, @fishbase",
                                  #"SeaLifeBase",
-                                 "Wikidata", "Open Tree Taxonomy",
-                                 "International Union for Conservation of Nature and Natural Resources"),
+                                 "Wikidata, @jorrit", "Open Tree Taxonomy, @Rees2017",
+                                 "International Union for Conservation of Nature and Natural Resources, @iucn"),
                     abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
@@
                                               select(acceptedNameUsageID) %>% pull() %>% n_distinct())
-) %>% printtable()
+) %>% knitr::kable(format = "markdown")
 

### 2019-10-11 paper/manuscript.Rmd
Add descriptions to the provider table.

@@
 ```{r echo = FALSE, results = "asis"}
+#provider descriptions
+desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government", 
+          ncbi = "",
+          col = "", 
+          gbif = "", 
+          fb = "", 
+          wd = "",
+          ott = "",
+          iucn = "")
+
 tibble(provider = c("Integrated Taxonomic Information System, @itis",
@@
                                             ~taxa_tbl(.x) %>% 
-                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
+                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
+                    description = desc
 ) %>% knitr::kable(format = "markdown")

### 2019-10-30 README.Rmd
Import RSQLite and clarify the multi-db methods

@@
 taxadb:::td_disconnect()
-MonetDBLite::monetdblite_shutdown()
 ```
@@
 library(taxadb)
-library(dplyr)
-library(readr)
+library(dplyr) # Used to illustrate how a typical workflow combines nicely with `dplyr`
 ```
@@
 bbs_species_list <- system.file("extdata/bbs.tsv", package="taxadb")
-bbs <- read_tsv(bbs_species_list)
+bbs <- read.delim(bbs_species_list)
 ```
@@
 
-See richer examples, including name cleaning and background on the schema in the vignettes:
+- See richer examples the package [Tutorial](https://cboettig.github.io/taxadb/articles/intro.html).
 
-- [Tutorial](https://cboettig.github.io/taxadb/articles/intro.html)
-- [Schema](https://cboettig.github.io/taxadb/articles/articles/schema.html)
+- Learn about the underlying data sources and formats in [Data Sources](https://cboettig.github.io/taxadb/articles/articles/data-sources.html)
+
+- Get better performance by selecting an alternative [database backend](https://cboettig.github.io/taxadb/articles/articles/backends.html) engines.
 

### 2019-10-30 README.Rmd
Update articles/manuscript.

@@
 Names that cannot be uniquely resolved because they are known synonyms of multiple different species will also return `NA`.
-The `by_name` filtering functions can help us resolve this last case (see below).
+The `filter_name` filtering functions can help us resolve this last case (see below).
 
@@
 birds %>% 
-  mutate(accepted_name = get_names(id, "col"))
+  mutate(accepted_name = get_names(id, "col")) %>% 
+  head()
 ```
@@
 Local access to taxonomic data tables lets us do much more than look up names and ids.
-A family of `by_*` functions in `taxadb` help us work directly with subsets of the taxonomic data.
+A family of `filter_*` functions in `taxadb` help us work directly with subsets of the taxonomic data.
 As we noted above, this can be useful in resolving certain ambiguous names.
@@
 
-Using `by_name()`, we find this is because the name resolves not to zero matches, but to more than one match:
+Using `filter_name()`, we find this is because the name resolves not to zero matches, but to more than one match:
 
 ```{r}
-by_name("Trochalopteron henrici gucenense") 
+filter_name("Trochalopteron henrici gucenense") 
 ```
@@
 ```{r}
-by_name("Trochalopteron henrici gucenense")  %>%
+filter_name("Trochalopteron henrici gucenense")  %>%
   mutate(acceptedNameUsage = get_names(acceptedNameUsageID)) %>% 
@@
 
-Similar functions `by_id`, `by_rank`, and `by_common` take IDs, scientific ranks, or common names, respectively.  Here, we can get taxonomic data on all bird names in the Catalogue of Life:
+Similar functions `filter_id`, `filter_rank`, and `filter_common` take IDs, scientific ranks, or common names, respectively.
+Here, we can get taxonomic data on all bird names in the Catalogue of Life:
 
 ```{r}
-by_rank(name = "Aves", rank = "class", provider = "col")
+filter_rank(name = "Aves", rank = "class", provider = "col")
 ```
@@
 ```{r}
-by_rank(name = "Aves", rank = "class", provider = "col") %>%
+filter_rank(name = "Aves", rank = "class", provider = "col") %>%
   filter(taxonomicStatus == "accepted", taxonRank=="species") %>% 
@@
 
-`by_*` functions by default return in-memory data frames.
+`filter_*` functions by default return in-memory data frames.
 Because they are filtering functions, they return a subset of the full data which matches a given query (names, ids, ranks, etc), so the returned data.frames are smaller than the full record of a naming provider.
@@
 ```
-However, unlike the `by_*` functions which return convenient in-memory tables, this is still a remote connection.
+However, unlike the `filter_*` functions which return convenient in-memory tables, this is still a remote connection.
 This means that direct access using the `taxa_tbl()` function (or directly accessing the database connection using `td_connect()`) is more low-level and requires greater care.
@@
 Instead, we would first need to use a `collect()` to pull the summary table into memory.
-Users familiar with remote databases in `dplyr` will find using `taxa_tbl()` directly to be convenient and fast, while other users may find the `by_*` approach to be more intuitive.
+Users familiar with remote databases in `dplyr` will find using `taxa_tbl()` directly to be convenient and fast, while other users may find the `filter_*` approach to be more intuitive.
 
@@
   pull(acceptedNameUsageID) %>% 
-  by_id("col") %>%
+  filter_id("col") %>%
   select(scientificName)

### 2019-12-03 paper/manuscript.Rmd
Fix provider table.

--- 2019-10-03 paper/manuscript.Rmd
@@
 ```{r echo = FALSE, results = "asis"}
-tibble(provider = c("Integrated Taxonomic Information System, @itis",
+# #provider descriptions
+desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
+          ncbi = "nomenclature for sequences in the International Nucleotide Sequence Database Collaboration database",
+          col = "comphrehensive taxonomic effort, includes some other providers (e.g. itis)",
+          gbif = "taxonomic backbone of the GBIF database, assembled from other sources including COL",
+          fb = "nomenclature for global database of fishes",
+          wd = "",
+          ott = "comphrenesive tree of life based on phylogenetic trees and taxonomic data",
+          iucn = "taxonomy for classification of species status")
+
+tibble(provider = c("Integrated Taxonomic Information System, (ref:@itis)",
                                  "National Center for Biological Information's Taxonomy database, @ncbi",
@@
                                             ~taxa_tbl(.x) %>%
-                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
-) %>% knitr::kable(format = "markdown")
+                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
+                    description = desc
+) %>%
+  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
+  kable_styling(latex_options="scale_down") %>%
+  kableExtra::column_spec(1, width = "5cm") %>%
+  kableExtra::column_spec(4, width = "10cm")
 

### 2019-12-03 paper/manuscript.Rmd
Cleanup

@@
 
+<!--
+
 Table version of below - citation, number of unique identifier
@@
 
-<!--
-What other aspects should be included in the table? Citations?
+(ref:itis) Integrated Taxonomic Information System [@itis]
 -->
-
-(ref:itis) Integrated Taxonomic Information System [@itis]
 
@@
 Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications.
-Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary fo resolving even moderately sized data sets.
+Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets.
 `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers.
-`taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
-
-`taxadb` establishes a local database of provider allowing for consistent and reproducible queries. The set of functions operate consistently accross providers returning data in a standardized format.
-
+`taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.
+In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
 ----

### 2019-12-10 paper/manuscript.Rmd
Citations by hand in the provider table.

@@
 <!--
-
 Table version of below - citation, number of unique identifier
@@
 - International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn
-
-(ref:itis) Integrated Taxonomic Information System [@itis]
 -->
@@
 ```{r echo = FALSE, results = "asis"}
-# #provider descriptions
+#provider descriptions
 desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
@@
           fb = "nomenclature for global database of fishes",
-          wd = "",
           ott = "comphrenesive tree of life based on phylogenetic trees and taxonomic data",
@@
 
-tibble(provider = c("Integrated Taxonomic Information System, (ref:@itis)",
-                                 "National Center for Biological Information's Taxonomy database, @ncbi",
-                                 "Catalogue of Life, @col",
-                                 "Global Biodiversity Information Facility Taxonomic Backbone, @gbif",
-                                 "FishBase, @fishbase",
-                                 #"SeaLifeBase",
-                                 "Wikidata, @jorrit", "Open Tree Taxonomy, @Rees2017",
-                                 "International Union for Conservation of Nature and Natural Resources, @iucn"),
-                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
-                                     "wd", "ott", "iucn"),
+tibble(provider = c("Integrated Taxonomic Information System, ITIS 2019",
+                                 "National Center for Biological Information's Taxonomy database, NCBI",
+                                 "Catalogue of Life, COL",
+                                 "Global Biodiversity Information Facility Taxonomic Backbone, GBIF 2019",
+                                 "FishBase, Froese and Pauly 2019",
+                                 "Open Tree Taxonomy, OTT",
+                                 "International Union for Conservation of Nature and Natural Resources,IUCN 2019"),
+                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
                     total_identifiers = map(abbreviation,
@@
 In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
-----
+
+* cite other tools in the space (taxize + citations)
+* Frandz citations - grumpy systemitists; we're not doing anything inferential
+* no spell check/name modification
+* could be more than one name for rank
+* versioning paragraph - argument for version like provider; we process every so often
+  * comment on the release schedule, rolling is not always up to date
 

### 2019-12-11 paper/manuscript.Rmd
Beef up citations

@@
 As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names.
-How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
+How many species are in the combined data?
+Do the studies use the same names for the same species, or do they use different synonyms for the same species?
 Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
 These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
-While these issues have long been recognized in the literature [], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
-@itis; @ncbi; @col], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
+While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008, @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
+@itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
 Here, we present an R package, `taxadb`, which seeks to address this gap.
@@
 Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym.
-Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider.
+Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@Franz2009, @franz2018].
 The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
@@
 
-tibble(provider = c("Integrated Taxonomic Information System, ITIS 2019",
-                                 "National Center for Biological Information's Taxonomy database, NCBI",
-                                 "Catalogue of Life, COL",
-                                 "Global Biodiversity Information Facility Taxonomic Backbone, GBIF 2019",
-                                 "FishBase, Froese and Pauly 2019",
-                                 "Open Tree Taxonomy, OTT",
-                                 "International Union for Conservation of Nature and Natural Resources,IUCN 2019"),
+tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
+                                 "National Center for Biological Information's Taxonomy database (NCBI)",
+                                 "Catalogue of Life (COL)",
+                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
+                                 "FishBase (Froese and Pauly 2019)",
+                                 "Open Tree Taxonomy (OTT)",
+                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
                     abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),

### 2019-12-12 paper/manuscript.Rmd
Clean up

@@
 
-<!--
-`taxadb` does not require the use of an external database system, but will automatically make use of the fastest database server available.
-`taxadb` can also use any DBI compliant database accessible from R, such as MySQL or Postgres [see @DBI], though the default `duckdb` or `MonetDBLite` will be both significantly faster and easier to use.
--->
-
-<!--  Next 2 paragraphs should possibly be left to a manual, but are likely to be overlooked aspects there... -->
 By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage.
 `taxadb` will automatically handle opening, caching, and closing the database connection.
-
-<!-- 
-now sidestepped
-
-but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time.
-Users who need concurrent access from multiple sessions may wish to use a standalone MonetDB server or other database server instead.
--->
 
@@
 These modifications are not appropriate in all settings and should be used with care (Language from vignette).
+Spell check of input names is outside the scope of `taxadb`.
 
@@
 
-<!--
-Table version of below - citation, number of unique identifier
-
-- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
-- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
-- Catalogue of Life (COL) Annual Species list. (2018 version at present)
-- Global Biodiversity Information Facility Taxonomic Backbone [GBIF; @gbif].  
-- FishBase [@fishbase] taxonomic names
-- SeaLifeBase [@sealifebase] taxonomic names
-- Wikidata [@jorrit]
-- Open Tree Taxonomy, OTT
-- International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn
--->
-
 ```{r echo = FALSE, results = "asis"}
@@
 
-* cite other tools in the space (taxize + citations)
-* Frandz citations - grumpy systemitists; we're not doing anything inferential
-* no spell check/name modification
-* could be more than one name for rank
-* versioning paragraph - argument for version like provider; we process every so often
-  * comment on the release schedule, rolling is not always up to date
-
 \pagebreak

### 2020-01-21 paper/manuscript.Rmd
Add table captions

@@
 
-printtable <- function(df){
+printtable <- function(df, ...){
   df %>%
-    kableExtra::kable("latex", booktabs=T) %>%
+    kableExtra::kable("latex", booktabs=T, ...) %>%
     kableExtra::kable_styling(full_width = F)
@@
 
-Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 1), or simply using `all` to install all available providers:
+Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 6), or simply using `all` to install all available providers:
 
@@
 These issues are best illustrated using a minimal example.
-We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species:
+We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species [\@ref(tab:iucn_table)]
 
@@
 
-```{r echo=FALSE, cache = FALSE}
-status_data %>% printtable()
+```{r iucn_table, echo=FALSE, cache = FALSE}
+status_data %>% printtable(caption = "IUCN status for the selected 10 species.")
 ```
 
-```{r echo = FALSE, cache = FALSE }
-trait_data %>% printtable()
+```{r trait_table, echo = FALSE, cache = FALSE }
+trait_data %>% printtable(caption = "Trait data for the selected 10 species.")
 ```
@@
 ```{r echo = FALSE, cache = FALSE}
-joined %>% tidyr::replace_na(list(category = "-")) %>% printtable() # "NA" looks like a category, "-" is clearer visually
+joined %>% tidyr::replace_na(list(category = "-")) %>% 
+  printtable(caption = "IUCN and trait data joined on scientific name.") # "NA" looks like a category, "-" is clearer visually
 ```
@@
   select(elton_name, redlist_name, mass, category, id) %>%
-  printtable()
+  printtable(caption = "IUCN and trait data joined on taxonomic ID.")
 ```
@@
 Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym.
-Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@Franz2009, @franz2018].
+Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009, @Franz2018].
 The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
@@
 filter_name("Zosterops rendovae") %>% 
-  select(1:5)
+  select(1:5) %>%
+  printtable(caption = "Accepted name matches for Zosterops rendovae.")
 
@@
 ) %>%
-  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
+  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE, 
+               caption = "Descriptions of providers supported by taxaDB.") %>%
   kable_styling(latex_options="scale_down") %>%

### 2020-01-21 paper/manuscript.Rmd
Add in table references.

@@
 These issues are best illustrated using a minimal example.
-We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species [\@ref(tab:iucn_table)]
+We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species (Table 1, 2).
 
@@
 
-If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved:
+If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved (Table 3).
 
@@
 
-This results in many more matches, as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.
+This results in many more matches (Table 4), as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.
 While we have focused on a small example for visual clarity here, the `get_ids()` function in `taxadb` can quickly resolve hundreds of thousands of species names to unique identifiers, thanks to the performance of fast joins in a local MonetDBLite database.
@@
 
-We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between.
+We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between (Table 5).
 
@@
 \End{tcolorbox}
-
-### Supported providers
 

### 2020-01-23 README.Rmd
Add code of conduct
Change URL after ropensci migration

@@
 [![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
-[![Travis build status](https://travis-ci.org/cboettig/taxadb.svg?branch=master)](https://travis-ci.org/cboettig/taxadb)
-[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/cboettig/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxadb)
-[![Coverage status](https://codecov.io/gh/cboettig/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/cboettig/taxadb?branch=master)
+[![Travis build status](https://travis-ci.org/ropensci/taxadb.svg?branch=master)](https://travis-ci.org/ropensci/taxadb)
+[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ropensci/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/ropensci/taxadb)
+[![Coverage status](https://codecov.io/gh/ropensci/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/ropensci/taxadb?branch=master)
 [![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
@@
 ```{r eval=FALSE}
-devtools::install_github("cboettig/taxadb")
+devtools::install_github("ropensci/taxadb")
 ```
@@
 ```
+
+----
+
+Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md).
+By participating in this project you agree to abide by its terms.
+
+[![ropensci_footer](https://ropensci.org/public_images/ropensci_footer.png)](https://ropensci.org)

### 2020-01-23 paper/manuscript.Rmd
Change URL after ropensci migration

--- 2019-12-12 paper/manuscript.Rmd
@@
 library(printr)
-library(knitr)
 
@@
     kableExtra::kable("latex", booktabs=T) %>% 
-    kableExtra::kable_styling(full_width = F)
+    kable_styling(full_width = F)
 }
@@
 As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names.
-How many species are in the combined data?
-Do the studies use the same names for the same species, or do they use different synonyms for the same species?
+How many species are in the combined data? Do the studies use the same names for the same species, or do they use different synonyms for the same species?
 Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
 These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
-While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008, @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
-@itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
+While these issues have long been recognized in the literature [], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
+@itis; @ncbi; @col], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
 Here, we present an R package, `taxadb`, which seeks to address this gap.
@@
 ```{r eval=FALSE}
-remotes::install_github("cboettig/taxadb")
+remotes::install_github("ropensci/taxadb")
 ```
@@
 
+<!--
+`taxadb` does not require the use of an external database system, but will automatically make use of the fastest database server available.
+`taxadb` can also use any DBI compliant database accessible from R, such as MySQL or Postgres [see @DBI], though the default `duckdb` or `MonetDBLite` will be both significantly faster and easier to use.
+-->
+
+<!--  Next 2 paragraphs should possibly be left to a manual, but are likely to be overlooked aspects there... -->
 By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage.
 `taxadb` will automatically handle opening, caching, and closing the database connection.
+
+<!-- 
+now sidestepped
+
+but note that one limitation of MonetDBLite is the restriction against concurrent access: two separate R sessions on the same machine cannot both access the database at the same time.
+Users who need concurrent access from multiple sessions may wish to use a standalone MonetDB server or other database server instead.
+-->
 
@@
 These modifications are not appropriate in all settings and should be used with care (Language from vignette).
-Spell check of input names is outside the scope of `taxadb`.
 
 Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym.
-Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@Franz2009, @franz2018].
+Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider.
 The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
@@
 
-```{r echo = FALSE, results = "asis"}
-#provider descriptions
-desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
-          ncbi = "nomenclature for sequences in the International Nucleotide Sequence Database Collaboration database",
-          col = "comphrehensive taxonomic effort, includes some other providers (e.g. itis)",
-          gbif = "taxonomic backbone of the GBIF database, assembled from other sources including COL",
-          fb = "nomenclature for global database of fishes",
-          ott = "comphrenesive tree of life based on phylogenetic trees and taxonomic data",
-          iucn = "taxonomy for classification of species status")
+Table version of below - citation, number of unique identifier
 
-tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
-                                 "National Center for Biological Information's Taxonomy database (NCBI)",
-                                 "Catalogue of Life (COL)",
-                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
-                                 "FishBase (Froese and Pauly 2019)",
-                                 "Open Tree Taxonomy (OTT)",
-                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
-                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
+- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
+- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
+- Catalogue of Life (COL) Annual Species list. (2018 version at present)
+- Global Biodiversity Information Facility Taxonomic Backbone [GBIF; @gbif].  
+- FishBase [@fishbase] taxonomic names
+- SeaLifeBase [@sealifebase] taxonomic names
+- Wikidata [@jorrit]
+- Open Tree Taxonomy, OTT
+- International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn
+
+<!--
+What other aspects should be included in the table? Citations?
+-->
+
+```{r cache = FALSE, echo = FALSE}
+tibble(provider = c("Integrated Taxonomic Information System", 
+                                 "National Center for Biological Information's Taxonomy database",
+                                 "Catalogue of Life",
+                                 "Global Biodiversity Information Facility Taxonomic Backbone",
+                                 "FishBase",
+                                 #"SeaLifeBase",
+                                 "Wikidata", "Open Tree Taxonomy",
+                                 "International Union for Conservation of Nature and Natural Resources"),
+                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", #"slb",
+                                     "wd", "ott", "iucn"),
                     total_identifiers = map(abbreviation, 
                                             ~taxa_tbl(.x) %>% 
-                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
-                    description = desc
-) %>%
-  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE) %>%
-  kable_styling(latex_options="scale_down") %>%
-  kableExtra::column_spec(1, width = "5cm") %>%
-  kableExtra::column_spec(4, width = "10cm")
+                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct())
+) %>% printtable()
 
@@
 Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications.
-Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets.
+Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary fo resolving even moderately sized data sets.
 `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers.
-`taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.
-In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
+`taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.In particular, `taxadb` is not a replacement for the APIs or databases provided, but merely an interface to taxonomic naming information contained within that data.
+
+`taxadb` establishes a local database of provider allowing for consistent and reproducible queries. The set of functions operate consistently accross providers returning data in a standardized format.
+
+----
 

### 2020-01-23 README.Rmd
Change URL after ropensci migration

@@
 [![Travis build status](https://travis-ci.org/ropensci/taxadb.svg?branch=master)](https://travis-ci.org/ropensci/taxadb)
-[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ropensci/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/ropensci/taxadb)
+[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ropensci/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxadb)
 [![Coverage status](https://codecov.io/gh/ropensci/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/ropensci/taxadb?branch=master)

### 2020-01-23 paper/manuscript.Rmd
Minor typos/adjustments

--- 2020-01-21 paper/manuscript.Rmd
@@
 Do the studies use the same names for the same species, or do they use different synonyms for the same species?
-Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned  datasets (Figure 1).
-These challenges have become particularly acute as it becomes increasingly common for researchers to work across larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise all clades needed to resolve scientific names [@Patterson2010].
-While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008, @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
+Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned datasets.
+These challenges have become particularly acute as it becomes increasingly common for researchers to work across a larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise for all clades needed to resolve scientific names [@Patterson2010].
+
+While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
 @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
@@
 
-- Queries require internet access at all times
-- Queries are slow and inefficient to implement and perform; frequently requiring separate API calls for each taxonomic name
+- Queries require internet access at all times.
+- Queries are slow and inefficient to implement and perform; frequently requiring separate API calls for each taxonomic name.
 - The type of query is highly limited by the API design.
-For instance, it is usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?"
+For instance, it is usually impossible to make queries across the entire corpus of names, such as "which accepted name has the most known synonyms?".
 - Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
@@
 Many names providers update the server data either continuously or at regular intervals, including both revising existing names and adding new names.
-<!-- For instance, @predicts2014 observes that 2013 Catalogue of Life includes the genus "Notiophilus" in two separate families, contrary to the rules of nomenclature.
-This error does not occur in the 2018 edition.
--->
 
@@
 `taxadb` functions like `get_ids()` and `td_create()` take an optional argument, `db`, to an external database connection.
-`taxadb` will work with most DBI-compliant databases such as MySQL or Postgres [see @DBI], but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.
+`taxadb` will work with most DBI-compliant databases such as MySQL or Postgres, but will be much faster when using a column-oriented database engine such as `duckdb` or `MonetDBLite`.
 These latter options are also much easier for most users, since each can be installed directly as an R package.
@@
 Taxonomic identifiers provide a fundamental abstraction which lies at the heart of managing taxonomic names.
-For instance, by resolving scientific names to identifiers, we can idetnify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized.
+For instance, by resolving scientific names to identifiers, we can identify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized.
 Unmatched names may indicate an error in data entry or otherwise warrant further investigation.
@@
 `Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name.
-These modifications are not appropriate in all settings and should be used with care (Language from vignette).
+These modifications are not appropriate in all settings and should be used with care.
 Spell check of input names is outside the scope of `taxadb`.
@@
 Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym.
-Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009, @Franz2018].
+Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009; @Franz2018].
 The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
@@
 
-As well getting a detailed look at specific species or ids. For example, why can't `get_ids` resolve a seemingly common species?
+As well as giving a detailed look at specific species or ids. For example, why can't `get_ids` resolve a seemingly common species?
 
 ```{r}
-## example of sci name matching to more than one id.
-filter_name("Zosterops rendovae") %>% 
+multi_match <- filter_name("Zosterops rendovae")
+```
+
+```{r echo = FALSE, cache = FALSE}
+multi_match %>%
   select(1:5) %>%
   printtable(caption = "Accepted name matches for Zosterops rendovae.")
-
 ```

### 2020-01-24 README.Rmd
- update README urls - fast build for vignette

@@
 
-- See richer examples the package [Tutorial](https://cboettig.github.io/taxadb/articles/intro.html).
+- See richer examples the package [Tutorial](https://docs.ropensci.org/taxadb/articles/intro.html).
 
-- Learn about the underlying data sources and formats in [Data Sources](https://cboettig.github.io/taxadb/articles/articles/data-sources.html)
+- Learn about the underlying data sources and formats in [Data Sources](https://docs.ropensci.org/taxadb/articles/articles/data-sources.html)
 
-- Get better performance by selecting an alternative [database backend](https://cboettig.github.io/taxadb/articles/articles/backends.html) engines.
+- Get better performance by selecting an alternative [database backend](https://docs.ropensci.org/taxadb/articles/articles/backends.html) engines.
 
@@
 
-Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md).
+Please note that this project is released with a [Contributor Code of Conduct](https://ropensci.org/code-of-conduct/).
 By participating in this project you agree to abide by its terms.

### 2020-01-24 README.Rmd
fix links

@@
 
-- See richer examples the package [Tutorial](https://docs.ropensci.org/taxadb/articles/intro.html).
+- See richer examples the package [Tutorial](https://docs.ropensci.org/taxadb/articles/articles/intro.html).
 
-- Learn about the underlying data sources and formats in [Data Sources](https://docs.ropensci.org/taxadb/articles/articles/data-sources.html)
+- Learn about the underlying data sources and formats in [Data Sources](https://docs.ropensci.org/taxadb/articles/data-sources.html)
 
-- Get better performance by selecting an alternative [database backend](https://docs.ropensci.org/taxadb/articles/articles/backends.html) engines.
+- Get better performance by selecting an alternative [database backend](https://docs.ropensci.org/taxadb/articles/backends.html) engines.
 

### 2020-01-30 R/filter_name.R
Patch test data (#70)
* use a local copy of data to test against

* cached data

* fix cache filename

bump version

* db = NULL can only test against online itis version

* fix bug introduced in get_names

* rebuild readme, codemeta

@@ -32,9 +32,7 @@
 #' }
 #'
 filter_name <- function(name,
-                provider =c("itis", "ncbi", "col", "tpl",
-                            "gbif", "fb", "slb", "wd", "ott",
-                            "iucn"),
+                provider = getOption("taxadb_default_provider", "itis"),
                 version = latest_version(),
                 collect = TRUE,
                 ignore_case = TRUE,
@@ -42,7 +40,7 @@ filter_name <- function(name,
 
   filter_by(x = name,
             by = "scientificName",
-            provider = match.arg(provider),
+            provider = provider,
             version = version,
             collect = collect,
             db = db,

### 2020-01-30 R/get_names.R

@@ -19,15 +19,12 @@
 #' @export
 #' @importFrom dplyr pull select collect distinct
 get_names <- function(id,
-                      db = c("itis", "ncbi", "col", "tpl",
-                             "gbif", "fb", "slb", "wd", "ott",
-                             "iucn"),
+                      db = getOption("taxadb_default_provider", "itis"),
                       version = latest_version(),
                       format = c("guess", "prefix", "bare", "uri"),
                       taxadb_db = td_connect()
                      ){
   format <- match.arg(format)
-  db <- match.arg(db)
   n <- length(id)
 
 

### 2020-01-30 R/synonyms.R

@@ -21,9 +21,7 @@
 #' }
 #'
 synonyms <- function(name,
-                     provider = c("itis", "ncbi", "col", "tpl",
-                                  "gbif", "fb", "slb", "wd", "ott",
-                                  "iucn"),
+                     provider = getOption("taxadb_default_provider", "itis"),
                      version = latest_version(),
                      collect = TRUE,
                      db = td_connect()){

### 2020-01-30 tests/testthat/setup-test_db.R

@@ -1,12 +1,14 @@
 
-
-taxadb:::td_disconnect()
-test_db <- file.path(tempdir(), "taxadb")
-#dir.create(test_db, showWarnings = FALSE)
-Sys.setenv(TAXADB_HOME=test_db)
+## Optionally: Force a backend type, otherwise will use the best available
+## (which is currently MonetDBLite, since it is suggested)
 #Sys.setenv(TAXADB_DRIVER="MonetDBLite")
+#Sys.setenv(TAXADB_DRIVER="RSQLite")
 
-## Consider adding some minimal subset of, say, ITIS data file that
-## would allow offline testing.  Currently pretty much all tests first
-## require some access to the online data caches
+## All tests only write to tempdir
+test_db <- file.path(tempdir(), "taxadb")
+Sys.setenv(TAXADB_HOME=test_db)
+
+
+## Use locally cached version to allow for offline testing
+options(taxadb_default_provider = "itis_test")
 

### 2020-01-30 tests/testthat/test-fast.R

@@ -11,17 +11,16 @@ test_that("setup-free calls to basic functions", {
   skip_if_offline()
 
   df <- taxa_tbl(provider = "itis", db = NULL)
-  chameleons <- filter(df, family == "Chamaeleonidae")
+  sp <- filter(df, family == "Hominidae")
 
   suppressWarnings({
-  df <- filter_rank(name = "Aves", rank = "class", db = NULL)
-  species <- filter_name(df$scientificName, db = NULL)
+  df <- filter_rank(name = "Hominidae", rank = "family", provider = "itis", db = NULL)
+  species <- filter_name("Pan troglodytes", provider = "itis", db = NULL)
   })
 
   expect_is(df, "data.frame")
-  expect_is(species, "data.frame")
-  expect_is(chameleons, "data.frame")
+  expect_is(sp, "data.frame")
   expect_gt(dim(df)[1], 1)
-  expect_gt(dim(chameleons)[1], 1)
+  expect_gt(dim(sp)[1], 1)
 
 })

### 2020-01-30 tests/testthat/test-mutate-db.R

@@ -3,39 +3,12 @@ testthat::test_that("mutate_db works on databases", {
 
   skip_on_cran()
   skip_if(Sys.getenv("TAXADB_DRIVER") == "RSQLite")
-  library(dplyr)
-  library(taxadb)
-  td_create(c("itis", "ncbi"))
-
-  chameleons <- taxa_tbl("ncbi") %>%
-    filter(family == "Chamaeleonidae",
-           taxonomicStatus != "accepted") %>%
-    select(species = scientificName) %>%
-    collect() %>%
-    mutate(input = clean_names(species),
-           sort = seq_along(species))
 
   ## Input table with clean names
   ## Let's get some matches, amazing how bad this is.  Need wikidata synonyms
   taxa <- taxa_tbl("itis") %>%
-    mutate_db(clean_names, "scientificName", "input") %>%
-    right_join(chameleons, copy = TRUE, by = "input") %>%
-    arrange(sort)  %>%
-    collect()
+    mutate_db(clean_names, "scientificName", "input")
 
-  ## lots of duplicate matches, pick the first one for now:
-  matched <- taxa %>% select(acceptedNameUsageID, sort) %>% distinct() %>%
-    group_by(sort) %>% top_n(1, acceptedNameUsageID)
-
-  ## Some matches
-  expect_is(matched, "data.frame")
-  expect_gt(dim(matched)[1], 1)
-  unmatched <- anti_join(chameleons, matched, by = "sort")
-  unmatched
-
-  ## Others not so much
-  expect_is(unmatched, "data.frame")
-  expect_gt(dim(unmatched)[1], 1)
 
 })
 

### 2020-01-31 README.Rmd
patch way duckdb calls disk storage location
duckdb was attempting in-memory location, since
`duckdb::duckdb()` does not seem to follow the `dbname` convention of MonetDBLite, RSQLite and others in passing the path for persistent storage.  cc @noamross

@@
 ```{r}
-most_synonyms <- taxa_tbl("col") %>% 
-  group_by(acceptedNameUsageID) %>% 
-  count(sort=TRUE)
+most_synonyms <- 
+  taxa_tbl("col") %>% 
+  count(acceptedNameUsageID, sort=TRUE)
 most_synonyms
@@
 ```{r}
-most_synonyms %>% 
-  head(1) %>% 
-  pull(acceptedNameUsageID) %>% 
-  filter_id("col") %>%
-  select(scientificName)
+#most_synonyms %>% 
+#  head(1) %>% 
+#  pull(acceptedNameUsageID) %>% 
+#  filter_id("col") %>%
+#  select(scientificName)
+
+filter_id("COL:43082445", "col")
 ```

### 2020-01-31 R/taxa_tbl.R

@@ -64,7 +64,7 @@ quick_db <-
     #tmp <- tempfile(fileext = ".tsv.bz2")
     tmp <- file.path(taxadb_dir(), paste0(tbl_name, ".tsv.bz2"))
     if(!file.exists(tmp)){
-      download.file(paste0(providers_download_url(filename, version), ".tsv.bz2"),
+      curl::curl_download(paste0(providers_download_url(filename, version), ".tsv.bz2"),
              tmp)
     }
     suppressWarnings(

### 2020-01-31 tests/testthat/setup-test_db.R

@@ -3,12 +3,14 @@
 ## (which is currently MonetDBLite, since it is suggested)
 #Sys.setenv(TAXADB_DRIVER="MonetDBLite")
 #Sys.setenv(TAXADB_DRIVER="RSQLite")
+#Sys.setenv(TAXADB_DRIVER="duckdb")
 
 ## All tests only write to tempdir
 test_db <- file.path(tempdir(), "taxadb")
 Sys.setenv(TAXADB_HOME=test_db)
 
 
+print(paste("Testing using backend", class(td_connect())))
+
 ## Use locally cached version to allow for offline testing
 options(taxadb_default_provider = "itis_test")
-

### 2020-01-31 paper/manuscript.Rmd
working on versioning paragraph

@@
 
+## Data Versioning
+
+Some taxonomic name providers (e.g.
+OTT, COL) offer periodic releases of a static names list, while many other providers (e.g.
+ITIS, NCBI, FB, IUCN) offer name data on a rolling basis (i.e.
+the data returned by a given download URL is updated continuously or at arbitrary intervals without any additional indication if and how that data has changed.)  `taxadb`'s `td_create()` function downloads and stores cached snapshopts from each provider, which follow an annual release model to support reproducible analyses.
+By default, `taxadb` will determine the latest version available (at the time of writing this is version `2019`), and install and use this version of all snapshots for each provider.
+However, all taxadb functions can take
+
 ## Discussion

### 2020-01-31 paper/manuscript.Rmd
WIP

@@
 library(printr)
+library(rticles)
 library(knitr)
@@
 the data returned by a given download URL is updated continuously or at arbitrary intervals without any additional indication if and how that data has changed.)  `taxadb`'s `td_create()` function downloads and stores cached snapshopts from each provider, which follow an annual release model to support reproducible analyses.
-By default, `taxadb` will determine the latest version available (at the time of writing this is version `2019`), and install and use this version of all snapshots for each provider.
-However, all taxadb functions can take
+All taxadb functions that download or access data include an optional argument `version` to indicate which version of the provider data should be used.
+By default, `taxadb` will determine the latest version available (at the time of writing this is version `2019`).
+Details on the versions of data used
 

### 2020-02-06 README.Rmd
streamline final example in README

--- 2020-01-24 README.Rmd
@@
 ```{r}
-most_synonyms <- taxa_tbl("col") %>% 
-  group_by(acceptedNameUsageID) %>% 
-  count(sort=TRUE)
-most_synonyms
+taxa_tbl("col") %>% 
+  count(acceptedNameUsageID, sort=TRUE)
 ```
+
 However, unlike the `filter_*` functions which return convenient in-memory tables, this is still a remote connection.
@@
 Users familiar with remote databases in `dplyr` will find using `taxa_tbl()` directly to be convenient and fast, while other users may find the `filter_*` approach to be more intuitive.
-
-So which species had those 456 names?
-
-```{r}
-most_synonyms %>% 
-  head(1) %>% 
-  pull(acceptedNameUsageID) %>% 
-  filter_id("col") %>%
-  select(scientificName)
-```
 

### 2020-02-06 paper/manuscript.Rmd
A bit more text, re-organization, spell check

@@
     affiliation: ucb
-#  - name: "Jorrit Poelen"
-#    affiliation: jorritt
   - name: "Scott Chamberlain"
@@
     address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
-# - code: jorrit
-#   address: "Independent consultant, Oakland, CA, USA"
   - code: ropensci
@@
 In contrast to existing tools, `taxadb` provides the following:
-  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
+  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hierarchical classification.  
   2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
   3) `taxadb` provides a consistent and intuitive data format.  
-  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.
+  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accommodate new authorities.
 
@@
 
+## Display NAs as "-" because "NA" looks like a category
+options(knitr.kable.NA = '-')
+
 printtable <- function(df, ...){
@@
 }
-knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
 
@@
 
+taxadb::td_disconnect()
 ```
 
-As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species, we face a continual challenge of reconciling taxonomic names.
-How many species are in the combined data?
+As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species,
+we face a continual challenge of reconciling taxonomic names. How many species are in the combined data?
 Do the studies use the same names for the same species, or do they use different synonyms for the same species?
 Failing to correct for such differences can lead to significant inflation of species counts and miss-aligned datasets.
-These challenges have become particularly acute as it becomes increasingly common for researchers to work across a larger number and diversity of species in any given analysis, which may preclude the resources or substantative taxonomic expertise for all clades needed to resolve scientific names [@Patterson2010].
+These challenges have become particularly acute as it becomes increasingly common for researchers to work
+across a larger number and diversity of species in any given analysis, 
+which may preclude the resources or substantive taxonomic expertise for all clades 
+needed to resolve scientific names [@Patterson2010].
 
-While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015], and a growing number of databases and tools have emerged and grown over the past few decades [e.g.
-@itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018], it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
+While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015],
+and a growing number of databases and tools have emerged and grown over the past few decades
+[e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018],
+it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
 Here, we present an R package, `taxadb`, which seeks to address this gap.
 
-Databases of taxonomic names such as the Integrated Taxonomic Information System [ITIS; @itis], the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi], the Catalogue of Life [COL; @col], and over one hundred other providers have sought to address these problems by providing expert-curated lists of accepted taxonomic names, synonyms, associated taxonomic rank, hierarchical classification, and scientific authority (e.g.
-author and date) establishing a scientific name.
-The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013] has become a popular way for R users to interact with naming providers and name resolution services.
-`taxize` implements bindings to the web APIs (Application Programming Interface) hosted by many popular taxonomic name providers.
-Unfortunately, this means that functions in the `taxize` are impacted by several major drawbacks that are inherent in the implementation of these central API servers, such as:
+Databases of taxonomic names such as the Integrated Taxonomic Information System [ITIS; @itis], 
+the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi], 
+the Catalogue of Life [COL; @col], and over one hundred other providers have sought to address these problems
+by providing expert-curated lists of accepted taxonomic names, synonyms, associated taxonomic rank,
+hierarchical classification, and scientific authority (e.g. author and date) establishing a scientific name.
+The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013]
+has become a popular way for R users to interact with naming providers and name resolution services. 
+`taxize` implements bindings to the web APIs (Application Programming Interface)
+hosted by many popular taxonomic name providers. 
+Unfortunately, this means that functions in the `taxize` are impacted by several major drawbacks
+that are inherent in the implementation of these central API servers, such as:
 
@@
 
-Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files following a consistent, standardized layout or schema following the Darwin Core Standard [@Wieczorek2012].
-These files are automatically downloaded and imported and stored on a local database by `taxadb`.
-The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes under a GB, and can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection.
-In contrast, querying a single name over the web API, requiring the server to respond, execute the query, and serialize the response, can take several seconds.
-Thus it does not take many taxa before transferring the entire data set to query locally is more efficient.
-Moreover, this local copy can be cached on the user's machine, requiring only the one-time setup, and enabling offline use and reproducible queries.
+Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files which are 
+automatically downloaded and imported and stored on a local database by `taxadb`.  
+The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes
+over a GB, which can be compressed to around 50 MB and downloaded in under a minute on a 1 MB/s connection.  
+By using a local database as the backend, `taxadb` allows R users to interact with large data files without large memory (RAM) requirements.
+A query for a single name over the web API requires a remote server to respond, execute the query,
+and serialize the response, which can take several seconds. Thus it does not take many taxa before transferring the 
+entire data set to query locally is more efficient.  Moreover, this local copy can be cached on the user's machine, 
+requiring only the one-time setup, and enabling offline use and reproducible queries.  Rather than returning data
+in whatever format is given by the provider, `taxadb` provides a data structure following a consistent,
+standardized layout or schema following the Darwin Core Standard [@Wieczorek2012].
+Table 1 summarizes the list of all naming providers currently accessed by `taxadb`.
+More details are provided in the Data Sources Vignette, <https://docs.ropensci.org/taxadb/articles/data-sources.html>.
+
+```{r echo = FALSE, results = "asis"}
+#provider descriptions
+desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
+          ncbi = "nomenclature for sequences in the International Nucleotide Sequence Database Collaboration database",
+          col = "comprehensive taxonomic effort, includes some other providers (e.g. itis)",
+          gbif = "taxonomic backbone of the GBIF database, assembled from other sources including COL",
+          fb = "nomenclature for global database of fishes",
+          ott = "comprehensive tree of life based on phylogenetic trees and taxonomic data",
+          iucn = "taxonomy for classification of species status")
+
+tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
+                                 "National Center for Biological Information's Taxonomy database (NCBI)",
+                                 "Catalogue of Life (COL)",
+                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
+                                 "FishBase (Froese and Pauly 2019)",
+                                 "Open Tree Taxonomy (OTT)",
+                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
+                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
+                    total_identifiers = map(abbreviation,
+                                            ~taxa_tbl(.x) %>%
+                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
+                    description = desc
+) %>%
+  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE, 
+               caption = "Descriptions of providers supported by taxadb.") %>%
+  kable_styling(latex_options="scale_down") %>%
+  kableExtra::column_spec(1, width = "5cm") %>%
+  kableExtra::column_spec(4, width = "10cm")
+
+```
 
@@
 
+# Package Overview
+
 ```{r message = FALSE, warning=FALSE}
@@
 
-`taxadb` functions will automatically set up a local, persistent database if one has not yet been created.
+After loading the package, we look up the taxonomic identifier for a Atlantic Cod, *Gadus morhua*, and the compliment:
 
@@
 get_ids("Gadus morhua")
+get_names("ITIS:164712")
 ```
 
-This one-time setup will download, uncompress, and import the data into persistant storage (using the appropriate location specified by the operating sytstem [see @rappdirs], or configured using the environmental variable `TAXADB_HOME`).
+Our first call to any `taxadb` functions will automatically set up a local, persistent database if one has not yet been created.
+This one-time setup will download, extract, and import the compressed data into persistent database storage (using the appropriate location specified by the operating system [see @rappdirs], or configured using the environmental variable `TAXADB_HOME`).
 The example above searches for names in ITIS, the default provider, which can be configured using the `provider` argument.
@@
 
-By default `taxadb` creates a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage.
-`taxadb` will automatically handle opening, caching, and closing the database connection.
+`taxadb` uses the widely known SQLite database by default, but users are encouraged to install the optional, suggested database backends by passing the option `dependencies = TRUE` to the install command.
+This installs a MonetDBLite database instance [@monetdblite], a columnar-oriented relational database requiring no additional installation while also providing persistent disk-based storage.
+This also installs `duckdb`, another local columnar database which is rapidly emerging as an alternative to MonetDB and SQLite.
+`taxadb` will automatically detect and use these database engines if available, and automatically handles opening, caching, and closing the database connection.
+For large queries, MonetDBLite or `duckdb` deliver impressive improvements.
+Our benchmark on resolving the 750 species names in in the Breeding Bird Survey against over 3 million names known in the 2019 Catalogue of Life takes 8 minutes in SQLite but less than a second in MonetDBLite.
 
@@
 
-## Taxonomic IDs
+## Taxonomic Identifiers
 
@@
 ```{r echo = FALSE, cache = FALSE}
-joined %>% tidyr::replace_na(list(category = "-")) %>% 
-  printtable(caption = "IUCN and trait data joined on scientific name.") # "NA" looks like a category, "-" is clearer visually
+joined %>%
+  printtable(caption = "IUCN and trait data joined on scientific name.")
 ```
@@
 
-`get_ids` offers a first pass at matching scientifc names to id, but names may remain unresolved for a number of reasons.
+`get_ids` offers a first pass at matching scientific names to id, but names may remain unresolved for a number of reasons.
 First, a name may match to multiple accepted names, as in the case of a species that has been split.
@@
 `clean_names` addresses common formatting issues such as the inclusion of missing species epithets (e.g.
-`Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the bionmial name.
+`Accipiter sp.`) that prevent matches to the Genus, or intraspecific epithets such as `Colaptes auratus cafer` that prevent matches to the binomial name.
 These modifications are not appropriate in all settings and should be used with care.
@@
 
-As well as giving a detailed look at specific species or ids. For example, why can't `get_ids` resolve a seemingly common species?
+We can also use this to gain a detailed look at specific species or ids.
+For example, we can explore why `get_ids` fails to resolve a seemingly common species:
 
@@
 We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between (Table 5).
+This is an example of how `taxadb` seeks to provide users with information from existing authorities and names providers,
+rather than make a potentially arbitrary decision.
+Because they return `data.frame`s,  `filter_*` functions provide both potential matches.
+Note that the simpler `get_` functions (`get_ids()`) consider multiple name as `NA` for the `id`, making them suitable for automated pipelines where manual resolution of duplicates is not an option.
 
-The full taxonomic record in the MonetDBLite database can also be directly accessed by `taxa_tbl()`, allowing for whole-database queries that are not possible through the API or web interface of many providers (see Vignette for further details).
+## Direct database access
+
+The full taxonomic record in the database can also be directly accessed by `taxa_tbl()`, allowing for whole-database queries that are not possible through the API or web interface of many providers.
+For example, we can easily check the coverage of accepted species names in each of the classes of vertabrates within the Catalogue of Life:
+
+```{r cache = FALSE}
+verts <- taxa_tbl("col") %>%
+  filter(taxonomicStatus == "accepted", phylum == "Chordata") %>% 
+  count(class, sort = TRUE)
+```
+
+```{r echo = FALSE}
+ verts %>%
+  printtable(caption = "taxadb also provides direct access to the database, allowing dplyr or SQL queries which can compute across the entire dataset, such as counting accepted species in all vertabrate classes shown here.
+This kind of query is effectively impossible in most REST API-based interfaces.")
+```
 
@@
 
-`taxizedb` can also resolve common names to their identifier by mapping common name to the accepted scientific name.
+`taxadb` can also resolve common names to their identifier by mapping common name to the accepted scientific name.
 Common names come with many of the same problems as scientific names but often more frequent, e.g.
@@
 Common names are accessed via `filter_common` which takes a vector of common names.
-The user can then resolve discrepencies.
+The user can then resolve discrepancies.
 
@@
 
-```{r echo = FALSE, results = "asis"}
-#provider descriptions
-desc <- c(itis = "originally formed to standardize taxonomic name usage across many agencies in the United States federal government",
-          ncbi = "nomenclature for sequences in the International Nucleotide Sequence Database Collaboration database",
-          col = "comphrehensive taxonomic effort, includes some other providers (e.g. itis)",
-          gbif = "taxonomic backbone of the GBIF database, assembled from other sources including COL",
-          fb = "nomenclature for global database of fishes",
-          ott = "comphrenesive tree of life based on phylogenetic trees and taxonomic data",
-          iucn = "taxonomy for classification of species status")
-
-tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
-                                 "National Center for Biological Information's Taxonomy database (NCBI)",
-                                 "Catalogue of Life (COL)",
-                                 "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
-                                 "FishBase (Froese and Pauly 2019)",
-                                 "Open Tree Taxonomy (OTT)",
-                                 "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
-                    abbreviation = c("itis", "ncbi", "col", "gbif", "fb", "ott", "iucn"),
-                    total_identifiers = map(abbreviation,
-                                            ~taxa_tbl(.x) %>%
-                                              select(acceptedNameUsageID) %>% pull() %>% n_distinct()),
-                    description = desc
-) %>%
-  knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE, 
-               caption = "Descriptions of providers supported by taxaDB.") %>%
-  kable_styling(latex_options="scale_down") %>%
-  kableExtra::column_spec(1, width = "5cm") %>%
-  kableExtra::column_spec(4, width = "10cm")
-
-```
-
-## Data Versioning
+# Discussion
 
@@
 ITIS, NCBI, FB, IUCN) offer name data on a rolling basis (i.e.
-the data returned by a given download URL is updated continuously or at arbitrary intervals without any additional indication if and how that data has changed.)  `taxadb`'s `td_create()` function downloads and stores cached snapshopts from each provider, which follow an annual release model to support reproducible analyses.
+the data returned by a given download URL is updated continuously or at arbitrary intervals without any additional indication if and how that data has changed.)  `taxadb`'s `td_create()` function downloads and stores cached snapshots from each provider, which follow an annual release model to support reproducible analyses.
 All taxadb functions that download or access data include an optional argument `version` to indicate which version of the provider data should be used.
 By default, `taxadb` will determine the latest version available (at the time of writing this is version `2019`).
-Details on the versions of data used
+Appropriate metadata is stored with each snapshot, including scripts used to access and reformat the data files, as described in the "Data Sources" vignette, <https://docs.ropensci.org/taxadb/articles/data-sources.html>.
 
-## Discussion
-
-Taxonomic indentifiers are a powerful tool for maintaing taxonomic consistency, an essential task for a wide variety of applications.
+Taxonomic identifiers are a powerful tool for maintaining taxonomic consistency, an essential task for a wide variety of applications.
 Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets.
@@
 
+Lastly, we note that local database design used in `taxadb` is not unique to taxonomic names.
+Despite the rapid expansion of REST API-based interfaces to ecological data [@ropensci], in our experience, much of the data relevant to ecologists and evolutionary biologists today would be also be ameniable to the local database design.
+The local database approach is much easier for data providers (who can leverage static scientific database repositories instead of maintaining REST servers) and often much faster for data consumers.
+
 \pagebreak
 
+```{r include = FALSE}
+td_disconnect()
+```
+
 # References

### 2020-02-08 paper/manuscript.Rmd
Change IUCN example data column name to consistently refer to IUCN by name rather than by "redlist".

@@
     kableExtra::kable("latex", booktabs=T, ...) %>%
-    kableExtra::kable_styling(full_width = F)
+    kableExtra::kable_styling(full_width = F, latex_options = "hold_position")
 }
@@
 While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015],
-and a growing number of databases and tools have emerged and grown over the past few decades
+and a growing number of databases and tools have emerged over the past few decades
 [e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018],
@@
 Instead of binding existing web APIs, `taxadb` is built around a set of compressed text files which are 
-automatically downloaded and imported and stored on a local database by `taxadb`.  
+automatically downloaded, imported, and stored on a local database by `taxadb`. 
 The largest of the taxonomic naming providers today contain under 6 million name records with uncompressed file sizes
@@
   knitr::kable(format = "latex", col.names = c("Provider", "Abbreviation", "Number of \nIdentifiers", "Description"), escape = FALSE, 
-               caption = "Descriptions of providers supported by taxadb.") %>%
-  kable_styling(latex_options="scale_down") %>%
+               caption = "Descriptions of the providers supported by taxadb with their reference abbreviation and the total number of identifiers contained by each provider.") %>%
+  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
   kableExtra::column_spec(1, width = "5cm") %>%
@@
 
-After loading the package, we look up the taxonomic identifier for a Atlantic Cod, *Gadus morhua*, and the compliment:
+After loading the package, we look up the taxonomic identifier for Atlantic Cod, *Gadus morhua*, and the compliment:
 
@@
 The example above searches for names in ITIS, the default provider, which can be configured using the `provider` argument.
-Any future function calls to this function or any other function using data from the same names provider will be able to access this data rapidly without the need for processing or an internet connection.
+Any future function calls to this function or any other function using data from the same provider will be able to access this data rapidly without the need for processing or an internet connection.
 
-Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 6), or simply using `all` to install all available providers:
+Users can also explicitly trigger this one-time setup using `td_create()` and specifying the provider abbreviation (see Table 1), or simply using `all` to install all available providers:
 
@@
 `taxadb` will default to the fastest available option.
-`taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the large names providers.
+`taxadb` can also run without a database backend by setting `db=NULL`, though some functions will require a lot (2-20 GB) of free RAM for this to work with many of the larger providers.
 
@@
 For large queries, MonetDBLite or `duckdb` deliver impressive improvements.
-Our benchmark on resolving the 750 species names in in the Breeding Bird Survey against over 3 million names known in the 2019 Catalogue of Life takes 8 minutes in SQLite but less than a second in MonetDBLite.
+Our benchmark on resolving the 750 species names in the Breeding Bird Survey against over 3 million names known in the 2019 Catalogue of Life takes 8 minutes in SQLite but less than a second in MonetDBLite.
 
@@
 These issues are best illustrated using a minimal example.
-We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits.]  To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species (Table 1, 2).
+We'll consider the task of combining data on bird extinction risk as assessed by the IUCN [@iucn] with data on average adult biomass, as estimated in the Elton Traits v1.0 database [@elton-traits].
+To keep the example concise enough for for visual presentation we will focus on a subset involving just 10 species (Table 2, 3).
 
@@
 ```{r iucn_table, echo=FALSE, cache = FALSE}
-status_data %>% printtable(caption = "IUCN status for the selected 10 species.")
+status_data %>% printtable(caption = "The subset of the IUCN status data used for subsequent taxonomic identifier examples.") 
 ```
@@
 ```{r trait_table, echo = FALSE, cache = FALSE }
-trait_data %>% printtable(caption = "Trait data for the selected 10 species.")
+trait_data %>% printtable(caption = "The subset of the Elton trait data used for subsequent taxonomic identifier examples.")
 ```
 
-If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved (Table 3).
+If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved (Table 4).
 
 ```{r}
-joined <- full_join(trait_data, status_data, by = c("elton_name" = "redlist_name")) 
+joined <- full_join(trait_data, status_data, by = c("elton_name" = "iucn_name")) 
 ```
@@
 joined %>%
-  printtable(caption = "IUCN and trait data joined on scientific name.")
+  printtable(caption = "Example IUCN and trait data joined directly on scientific name showing only one match.
+While common, joining on scientific name does not account for taxnomic inconsistiencies between databases and therefore results in seemingly very little overlap in species representation between the two.")
 ```
@@
 traits <- trait_data %>% mutate(id = get_ids(elton_name, "col"))
-status <- status_data %>% mutate(id = get_ids(redlist_name, "col"))
+status <- status_data %>% mutate(id = get_ids(iucn_name, "col"))
 ```
@@
 joined %>%  
-  tidyr::replace_na(list(category = "-", elton_name = "-", redlist_name = "-")) %>%
-  select(elton_name, redlist_name, mass, category, id) %>%
-  printtable(caption = "IUCN and trait data joined on taxonomic ID.")
+  tidyr::replace_na(list(category = "-", elton_name = "-", iucn_name = "-")) %>%
+  select(elton_name, iucn_name, mass, category, id) %>%
+  printtable(caption = "Example IUCN and trait data joined on taxonomic ID.
+Multilple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.")
 ```
 
-This results in many more matches (Table 4), as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.
+This results in many more matches (Table 5), as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.
 While we have focused on a small example for visual clarity here, the `get_ids()` function in `taxadb` can quickly resolve hundreds of thousands of species names to unique identifiers, thanks to the performance of fast joins in a local MonetDBLite database.
@@
 ```{r}
-multi_match <- filter_name("Zosterops rendovae")
+multi_match <- filter_name("Zosterops rendovae", provider = "col")
 ```
@@
   select(1:5) %>%
-  printtable(caption = "Accepted name matches for Zosterops rendovae.")
+  printtable(caption = "Some names may not resolve to an identifier using get\\_ids() because they match to more than one accepted ID.
+In such cases filter\\_ functions give further detail, as in the example of Zosterops rendovae below which has two accepted ID matches.", escape = FALSE)
 ```
 
-We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between (Table 5).
+We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between (Table 6).
 This is an example of how `taxadb` seeks to provide users with information from existing authorities and names providers,
@@
 The full taxonomic record in the database can also be directly accessed by `taxa_tbl()`, allowing for whole-database queries that are not possible through the API or web interface of many providers.
-For example, we can easily check the coverage of accepted species names in each of the classes of vertabrates within the Catalogue of Life:
+For example, we can easily check the coverage of accepted species names in each of the classes of vertabrates within the Catalogue of Life (Table 7):
 

### 2020-02-11 paper/coverletterMEE.Rmd
First commit of the cover letter.

---
author: Kari Norman
return-address: 
  - Dept of Environmental Science Policy and Management 
  - \linebreak University of California Berkeley
  - \linebreak 130 Hilgard Way
  - Berkeley, CA 94709

opening: Dear Editors,
closing: Thank you for your consideration,
signature: Kari Norman, Scott Chamberlain, Carl Boettiger
return-email: kari.norman@berkeley.edu

output: komaletter::komaletter
---

We are submitting our manuscript titled “taxadb: A High-Performance Local Taxonomic Database Interface” to be considered for publication as an Application in Methods in Ecology and Evolution.

Tracking and reconciling changes in taxonomy across time and between datasets is a core but often overlooked challenge in Ecology and Evolutionary biology. As combining larger and larger datasets becomes increasingly common, the potential for serious species mismatches has important implications for the scientific conclusions drawn. While multiple efforts to standardize taxonomic data exist and and some tools to access those providers have been developed, it remains difficult to resolve names to a taxonomic authority in quick, reproducible way. Here we present a new R package for accessing  taxonomic data using infrastructure that enables easy incorporation of taxonomic data to existing workflows and further exploration of the taxonomic providers. This package was reviewed and accepted by rOpenSci and can be found at www.github.com/ropensci/taxadb.

### 2020-02-12 paper/manuscript.Rmd
Final edits

@@
 layout: 3p
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
 output: 
@@
 
-<!--
-### Installation
-
-`taxadb` can be installed from GitHub using the `remotes` R package:
-
-```{r eval=FALSE}
-remotes::install_github("ropensci/taxadb")
-```
--->
-
 # Package Overview
@@
 
-If we first resolve names used in each data set into shared identifiers, (for instance, using the Catalogue of Life), we discover that there is far more overlap in the species coverage than we might have realized at first.
+If we first resolve names used in each data set into shared identifiers, (for instance, using the Catalogue of Life), we discover that there is far more overlap in the species coverage than we might have initially realized.
 First, we just add an ID column to each table by looking up the Catalog of Life identifier for the name provided:
@@
 First, a name may match to multiple accepted names, as in the case of a species that has been split.
-By design, these cases are left to be resolved by the researcher using the 'filter_' functions to filter underlying taxonomic tables for additional information.
+By design, these cases are left to be resolved by the researcher using the `filter_` functions to filter underlying taxonomic tables for additional information.
 A name may also be unresolved due to typos or improper formatting.
@@
 rather than make a potentially arbitrary decision.
-Because they return `data.frame`s,  `filter_*` functions provide both potential matches.
-Note that the simpler `get_` functions (`get_ids()`) consider multiple name as `NA` for the `id`, making them suitable for automated pipelines where manual resolution of duplicates is not an option.
+Because they return `data.frame`s,  `filter_` functions provide both potential matches.
+Note that the simpler `get_` functions (`get_ids()`) consider multiple name matches as `NA` for the `id`, making them suitable for automated pipelines where manual resolution of duplicates is not an option.
 

### 2020-02-12 paper/manuscript.Rmd
add acknowledgments and author contributions

@@
 author:
-  - name: "Kari Norman"
+  - name: "Kari E A Norman"
     affiliation: ucb
@@
 
+# Acknowledgments
+We thank the many researchers who contributed to the data and infrastructure of the various taxonomic providers we access through our package.
+Support for the development of this package was provided by United States Department of Energy through the Computational Sciences Graduate Fellowship (DOE CSGF) under grant number DE-FG02-97ER25308 awarded to K.E.A.N..
+
+# Authors' Contributions
+K.E.A.N., S.C., and C.B. contributed to conceptual development of the package. K.E.A.N. and C.B. developed the package and contributed to the manuscript.
+
 \pagebreak

### 2020-02-12 paper/manuscript.Rmd
final spell check

@@
   printtable(caption = "Example IUCN and trait data joined directly on scientific name showing only one match.
-While common, joining on scientific name does not account for taxnomic inconsistiencies between databases and therefore results in seemingly very little overlap in species representation between the two.")
+While common, joining on scientific name does not account for taxonomic inconsistencies between databases and therefore results in seemingly very little overlap in species representation between the two.")
 ```
@@
   printtable(caption = "Example IUCN and trait data joined on taxonomic ID.
-Multilple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.")
+Multiple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.")
 ```
@@
 The full taxonomic record in the database can also be directly accessed by `taxa_tbl()`, allowing for whole-database queries that are not possible through the API or web interface of many providers.
-For example, we can easily check the coverage of accepted species names in each of the classes of vertabrates within the Catalogue of Life (Table 7):
+For example, we can easily check the coverage of accepted species names in each of the classes of vertebrates within the Catalogue of Life (Table 7):
 
@@
  verts %>%
-  printtable(caption = "taxadb also provides direct access to the database, allowing dplyr or SQL queries which can compute across the entire dataset, such as counting accepted species in all vertabrate classes shown here.
+  printtable(caption = "taxadb also provides direct access to the database, allowing dplyr or SQL queries which can compute across the entire dataset, such as counting accepted species in all vertebrate classes shown here.
 This kind of query is effectively impossible in most REST API-based interfaces.")
@@
 Lastly, we note that local database design used in `taxadb` is not unique to taxonomic names.
-Despite the rapid expansion of REST API-based interfaces to ecological data [@ropensci], in our experience, much of the data relevant to ecologists and evolutionary biologists today would be also be ameniable to the local database design.
+Despite the rapid expansion of REST API-based interfaces to ecological data [@ropensci], in our experience, much of the data relevant to ecologists and evolutionary biologists today would be also be amenable to the local database design.
 The local database approach is much easier for data providers (who can leverage static scientific database repositories instead of maintaining REST servers) and often much faster for data consumers.

### 2020-02-14 README.Rmd
add peer review badge to readme

@@
 [![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
+[![peer-review](https://badges.ropensci.org/344_status.svg)](https://github.com/ropensci/software-review/issues/344)
 

### 2020-05-25 paper/manuscript.Rmd
reviewer responses, additional citations

@@
 Here, we present `taxadb` R package to address this gap.
-In contrast to existing tools, `taxadb` provides the following:
-  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hierarchical classification.  
+In contrast to existing tools, `taxadb` provides all of the following in one tool:
+  1) `taxadb` accesses established naming authorities to resolve synonyms, taxonomic identifiers, and hierarchical classification.  
   2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
   3) `taxadb` provides a consistent and intuitive data format.  
-  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accommodate new authorities.
+  4) `taxadb` is built on a simple, extensible, and language-agnostic design that can easily accommodate new authorities.
 
@@
 
-While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015],
+While these issues have long been recognized in the literature [@boyle2013; @dayrat2005; @bortolus2008; @maldonado2015; @remsen2016],
 and a growing number of databases and tools have emerged over the past few decades
-[e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018],
+[e.g. @itis; @ncbi; @col; @rees2014; @alvarez2018; @wagner2016; @foster2018; @gries2014],
 it remains difficult to resolve taxonomic names to a common authority in a transparent, efficient, and automatable manner.
@@
 by providing expert-curated lists of accepted taxonomic names, synonyms, associated taxonomic rank,
-hierarchical classification, and scientific authority (e.g. author and date) establishing a scientific name.
+hierarchical classifications, and scientific authority (e.g. author and date) establishing a scientific name.
 The R language [@R] is widely used in ecology and evolution [@Lai2019] and the `taxize` package [@Chamberlain2013]
@@
 hosted by many popular taxonomic name providers. 
-Unfortunately, this means that functions in the `taxize` are impacted by several major drawbacks
+Nevertheless, this means that functions in the `taxize` are impacted by several major drawbacks
 that are inherent in the implementation of these central API servers, such as:
@@
 - Both query formats and responses differ substantially across different naming providers, making it difficult to apply a script designed for one provider to different provider.
-- Most queries are not reproducible, as the results depend on the state of the central server (and potentially the quality of the internet connection).
-Many names providers update the server data either continuously or at regular intervals, including both revising existing names and adding new names.
+- Most queries are not reproducible, as the results depend on the state of the central server (and potentially the quality of the internet connection)[@rees2017].
+Many names providers update the server data either continuously or at regular intervals, including both revising existing names (for spelling or changes in accepted name designation) and adding new names.
 
@@
 in whatever format is given by the provider, `taxadb` provides a data structure following a consistent,
-standardized layout or schema following the Darwin Core Standard [@Wieczorek2012].
+standardized layout or schema following Darwin Core, which provides standard terms for biodiversity data [@Wieczorek2012].
 Table 1 summarizes the list of all naming providers currently accessed by `taxadb`.
@@
 
-After loading the package, we look up the taxonomic identifier for Atlantic Cod, *Gadus morhua*, and the compliment:
+After loading our package and the tidyverse package for ease in manipulating function output, we look up the taxonomic identifier for Atlantic Cod, *Gadus morhua*, and the compliment:
 
@@
 Taxon identifiers are also easily resolved to the original authority (scientific publication) establishing the name.
-(The historical practice of appending an author and year to a scientific name, e.g.
-*Poa annua ssp.
-annua* (Smith 1912), serves a valuable role in disambiguating different uses of the same name but can be notoriously harder to resolve to the appropriate reference, while variation in this convention creates many distinct versions of the same name [@Patterson2010]).
+The common practice of appending an author and year to a scientific name, e.g.
+*Poa annua annua* (Smith 1912), serves a valuable role in disambiguating different uses of the same name but can be notoriously harder to resolve to the appropriate reference, while variation in this convention creates many distinct versions of the same name [@Patterson2010].
 
@@
 ```{r iucn_table, echo=FALSE, cache = FALSE}
-status_data %>% printtable(caption = "The subset of the IUCN status data used for subsequent taxonomic identifier examples.") 
+status_data %>% printtable(caption = "The subset of the IUCN status data used for subsequent taxonomic identifier examples.") %>%
+  column_spec(1, italic = TRUE)
 ```
@@
 ```{r trait_table, echo = FALSE, cache = FALSE }
-trait_data %>% printtable(caption = "The subset of the Elton trait data used for subsequent taxonomic identifier examples.")
+trait_data %>% printtable(caption = "The subset of the Elton trait data used for subsequent taxonomic identifier examples.") %>%
+  column_spec(1, italic = TRUE)
 ```
 
-If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species having both a body mass and an IUCN threat status resolved (Table 4).
+If we attempted to join these data directly on the species names provided by each table, we would find very little overlap, with only one species name having both a body mass and an IUCN threat status resolved (Table 4).
 
@@
   printtable(caption = "Example IUCN and trait data joined directly on scientific name showing only one match.
-While common, joining on scientific name does not account for taxonomic inconsistencies between databases and therefore results in seemingly very little overlap in species representation between the two.")
+While common, joining on scientific name does not account for nomenclatural and taxonomic inconsistencies between databases and therefore results in seemingly very little overlap in species representation between the two.") %>%
+  column_spec(1, italic = TRUE)
 ```
@@
   printtable(caption = "Example IUCN and trait data joined on taxonomic ID.
-Multiple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.")
+Multiple species have a different scientific name in the Elton and IUCN Redlist databases but can be match based on their COL taxonomic ID.") %>%
+  column_spec(c(1,2), italic = TRUE)
 ```
@@
 In contrast, `taxadb` lists the `acceptedNameUsageID` for accepted names (where it matches the `taxonID`), as well as known synonyms.
-This is semantically identical, but also more convenient for database interfaces, since it allows a name to mapped to it's accepted identifier (or an identifier to map to it's accepted name usage) without the additional logic.
+This is semantically identical, but also more convenient for database interfaces, since it allows a name to mapped to its accepted identifier (or an identifier to map to it's accepted name usage) without the additional logic.
 For consistency, we will use the term "identifier" to mean the `acceptedNameUsageID` rather than the more ambiguous `taxonID` (which is undefined for synonyms listed by many providers), unless explicitly stated otherwise.
@@
 These modifications are not appropriate in all settings and should be used with care.
-Spell check of input names is outside the scope of `taxadb`.
+Spell check of input names is outside the scope of `taxadb`, however existing tools such as those developed by the Global Names Architecture (http://globalnames.org/apps/) could be incorporated into a `taxadb` workflow.
 
 Names may also have an ambiguous resolution wherein a name may be resolved by a different provider than the one specified, either as an accepted name or a synonym.
-Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009; @Franz2018].
+Mapping between providers represent a meaningful scientific statement requiring an understanding of the underlying taxonomic concepts of each provider [@franz2009; @Franz2018; @lepage2014].
 The spirit of taxadb is not to automate steps that require expert knowledge, but provide access to multiple potential "taxonomic theories".
@@
 filter_rank("Aves", rank="class", provider = "col") %>%
+  filter(taxonomicStatus == "accepted") %>%
   pull(taxonID) %>%
@@
   printtable(caption = "Some names may not resolve to an identifier using get\\_ids() because they match to more than one accepted ID.
-In such cases filter\\_ functions give further detail, as in the example of Zosterops rendovae below which has two accepted ID matches.", escape = FALSE)
+In such cases filter\\_ functions give further detail, as in the example of *Zosterops rendovae* below which has two accepted ID matches.", escape = FALSE) %>%
+  column_spec(3, italic = TRUE)
 ```
@@
 `taxadb` can also resolve common names to their identifier by mapping common name to the accepted scientific name.
-Common names come with many of the same problems as scientific names but often more frequent, e.g.
-matching to more than one accepted name, non-standardized formatting (like capitalization) and spelling etc.
+Common names have many of the same issues as scientific names even more frequently (e.g.
+matching to more than one accepted name, non-standardized formatting).
 Common names are accessed via `filter_common` which takes a vector of common names.
@@
 
-Taxonomic identifiers are a powerful tool for maintaining taxonomic consistency, an essential task for a wide variety of applications.
+Taxonomic identifiers are an essential first step for maintaining taxonomic consistency, a key task for a wide variety of applications.
 Despite multiple taxonomic standardization efforts, resolving names to taxonomic identifiers is often not a standard step in the research work flow due to difficulty in accessing providers and the time consuming API queries necessary for resolving even moderately sized data sets.

### 2020-05-27 paper/manuscript.Rmd
Update multiple accepted names example, change some language, fix species counting examples so they're not counting subspecies.

@@
 For instance, by resolving scientific names to identifiers, we can identify which names are synonyms -- different scientific names used to describe the same species -- and which names are not recognized.
+Each naming authority provides its own identifiers for the names it recognizes.
+For example, the name Homo sapiens has the identifier 9606 in NCBI and 180092 in ITIS.
+To avoid possible confusion, taxadb always prefixes the naming provider, e.g.
+NCBI:9606.
+Some taxonomic naming providers include separate identifiers for synonyms, see Box 1.
 Unmatched names may indicate an error in data entry or otherwise warrant further investigation.
@@
 filter_rank("Aves", rank="class", provider = "col") %>%
-  filter(taxonomicStatus == "accepted") %>%
+  filter(taxonomicStatus == "accepted", taxonRank == "species") %>%
   pull(taxonID) %>%
@@
 ```{r}
-multi_match <- filter_name("Zosterops rendovae", provider = "col")
+multi_match <- filter_name("Abies menziesii", provider = "col")
 ```
@@
 multi_match %>%
-  select(1:5) %>%
+  select(1:5, genus, specificEpithet) %>%
+  unite(acceptedScientificName, genus, specificEpithet, sep = " ") %>%
   printtable(caption = "Some names may not resolve to an identifier using get\\_ids() because they match to more than one accepted ID.
@@
 
-We see that *Zosterops rendovae* matches two accepted names which the user will have to choose between (Table 6).
+We see that *Abies menziesii* is a synonym for three accepted names which the user will have to choose between (Table 6).
 This is an example of how `taxadb` seeks to provide users with information from existing authorities and names providers,
@@
 verts <- taxa_tbl("col") %>%
-  filter(taxonomicStatus == "accepted", phylum == "Chordata") %>% 
+  filter(taxonomicStatus == "accepted", phylum == "Chordata", taxonRank == "species") %>% 
   count(class, sort = TRUE)
@@
 `taxadb` can also resolve common names to their identifier by mapping common name to the accepted scientific name.
-Common names have many of the same issues as scientific names even more frequently (e.g.
+Common names have many of the same issues as scientific names but even more frequent (e.g.
 matching to more than one accepted name, non-standardized formatting).
@@
 Some taxonomic name providers (e.g.
-OTT, COL) offer periodic releases of a static names list, while many other providers (e.g.
-ITIS, NCBI, FB, IUCN) offer name data on a rolling basis (i.e.
+OTT, COL, NCBI) offer periodic releases of a static names list, while many other providers (e.g.
+ITIS, FB, IUCN) offer name data on a rolling basis (i.e.
 the data returned by a given download URL is updated continuously or at arbitrary intervals without any additional indication if and how that data has changed.)  `taxadb`'s `td_create()` function downloads and stores cached snapshots from each provider, which follow an annual release model to support reproducible analyses.
@@
 `taxadb` fills an important gap between existing tools and typical research patterns by providing a fast, reproducible approach for matching names to taxonomic identifiers.
+It could also be used to verify that conclusions were robust to the choice of naming provider.
 `taxadb` is not intended as an improvement or replacement for any existing approaches to taxonomic name resolution.

### 2020-05-29 paper/manuscript.Rmd
Final language edits

@@
 abstract: |
-  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources.
-While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.
+  A familiar and growing challenge in ecological and evolutionary research is that of establishing consistent taxonomy when combining data from separate sources.
+While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to retrieve taxonomic names.
 Here, we present `taxadb` R package to address this gap.
-In contrast to existing tools, `taxadb` provides all of the following in one tool:
+In contrast to existing tools, `taxadb` provides all of the following in one place:
   1) `taxadb` accesses established naming authorities to resolve synonyms, taxonomic identifiers, and hierarchical classification.  
@@
 As ecologists and evolutionary biologists synthesize datasets across larger and larger assemblies of species,
-we face a continual challenge of reconciling taxonomic names. How many species are in the combined data?
+we face a continual challenge of maintaining consistent taxonomy. How many species are in the combined data?
 Do the studies use the same names for the same species, or do they use different synonyms for the same species?

### 2020-06-22 paper/manuscript.Rmd
Add Data availability statement.

@@
 
+# Data Availability
+Code for the R package can be found on GitHub at <https://github.com/ropensci/taxadb>.
+The taxonomic database is also stored on Github at <https://github.com/boettiger-lab/taxadb-cache>.
+The original taxonomic data is stored by the individual provider and can be found through the included citations.
+
 # Authors' Contributions

### 2020-06-22 README.Rmd
add zenodo badge to README

--- 2020-02-06 README.Rmd
@@
 [![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
+[![peer-review](https://badges.ropensci.org/344_status.svg)](https://github.com/ropensci/software-review/issues/344)
+[![DOI](https://zenodo.org/badge/130153207.svg)](https://zenodo.org/badge/latestdoi/130153207)
 

### 2020-06-23 paper/manuscript.Rmd
Update citations.

--- 2020-05-29 paper/manuscript.Rmd
@@
 Databases of taxonomic names such as the Integrated Taxonomic Information System [ITIS; @itis], 
-the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi], 
+the National Center for Biological Information's (NCBI) Taxonomy database [@ncbi], 
 the Catalogue of Life [COL; @col], and over one hundred other providers have sought to address these problems
@@
 tibble(provider = c("Integrated Taxonomic Information System (ITIS 2019)",
-                                 "National Center for Biological Information's Taxonomy database (NCBI)",
-                                 "Catalogue of Life (COL)",
+                                 "National Center for Biological Information's Taxonomy database (Biotechnology Information 2019)",
+                                 "Catalogue of Life (Roskov Y. 2018)",
                                  "Global Biodiversity Information Facility Taxonomic Backbone (GBIF 2019)",
                                  "FishBase (Froese and Pauly 2019)",
-                                 "Open Tree Taxonomy (OTT)",
+                                 "Open Tree Taxonomy (J. A. Rees and Cranston 2017)",
                                  "International Union for Conservation of Nature and Natural Resources (IUCN 2019)"),
@@
   printtable(caption = "Some names may not resolve to an identifier using get\\_ids() because they match to more than one accepted ID.
-In such cases filter\\_ functions give further detail, as in the example of *Zosterops rendovae* below which has two accepted ID matches.", escape = FALSE) %>%
+In such cases filter\\_ functions give further detail, as in the example of *Abies menziesii* below which has two accepted ID matches.", escape = FALSE) %>%
   column_spec(3, italic = TRUE)
@@
 
+# Data Availability
+Code for the R package can be found on GitHub at <https://github.com/ropensci/taxadb> and is archived on Zenodo at DOI:10.5281/zenodo.3903858 [@taxadb].
+The taxonomic database is also stored on Github at <https://github.com/boettiger-lab/taxadb-cache>.
+The original taxonomic data are stored by the individual provider, see "Catalogue of Life", <http://www.catalogueoflife.org/> [@col], "ITIS", <https://www.itis.gov> [@itis], "NCBI", <https://www.ncbi.nlm.nih.gov/taxonomy> [@ncbi], "GBIF", <https://gbif.org> [@gbif], "Fishbase", <https://fishbase.se> [@fishbase], "Open Tree Taxonomy", <https://tree.opentreeoflife.org> [@Rees2017], "IUCN", <https://www.iucnredlist.org/resources/tax-sources> [@iucn].
+
 # Authors' Contributions

### 2020-09-18 README.Rmd
gh-actions :car:

--- 2020-02-06 README.Rmd
@@
 
+<!-- badges: start -->
+[![R build status](https://github.com/ropensci/taxadb/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci/taxadb/actions)
 [![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
-[![Travis build status](https://travis-ci.org/ropensci/taxadb.svg?branch=master)](https://travis-ci.org/ropensci/taxadb)
-[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/ropensci/taxadb?branch=master&svg=true)](https://ci.appveyor.com/project/cboettig/taxadb)
 [![Coverage status](https://codecov.io/gh/ropensci/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/ropensci/taxadb?branch=master)
 [![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
+
+<!-- [![peer-review](https://badges.ropensci.org/344_status.svg)](https://github.com/ropensci/software-review/issues/344) -->
+<!-- badges: end -->
 

### 2020-09-18 README.Rmd

@@
 [![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
+[![DOI](https://zenodo.org/badge/130153207.svg)](https://zenodo.org/badge/latestdoi/130153207)
 

### 2020-09-25 R/td_create.R
cran url policy...

@@ -26,7 +26,7 @@
 #'  - `col`: Catalogue of Life, <http://www.catalogueoflife.org/>
 #'  - `tpl`: The Plant List, <http://www.theplantlist.org/>
 #'  - `gbif`: Global Biodiversity Information Facility, <https://www.gbif.org/>
-#'  - `fb`: FishBase, <http://fishbase.de>
+#'  - `fb`: FishBase, <http://fishbase.de/>
 #'  - `slb`: SeaLifeBase, <http://sealifebase.org>
 #'  - `wd`: Wikidata: https://www.wikidata.org
 #'  - `ott`: OpenTree Taxonomy:

### 2020-10-02 R/td_create.R
import database files via content_id (#79)
* tl_import

* no remotes (arkdb 0.0.6 release now on cran)

* update links bc CRAN hates 30x redirects

* cran url policy...

* there are 30x redirects and then there's just nonsense

* ...

* ...

@@ -26,9 +26,9 @@
 #'  - `col`: Catalogue of Life, <http://www.catalogueoflife.org/>
 #'  - `tpl`: The Plant List, <http://www.theplantlist.org/>
 #'  - `gbif`: Global Biodiversity Information Facility, <https://www.gbif.org/>
-#'  - `fb`: FishBase, <http://fishbase.org>
+#'  - `fb`: FishBase, <https://www.fishbase.de/>
 #'  - `slb`: SeaLifeBase, <http://sealifebase.org>
-#'  - `wd`: Wikidata: <https://www.wikidata.org/>
+#'  - `wd`: Wikidata: https://www.wikidata.org
 #'  - `ott`: OpenTree Taxonomy:
 #'  <https://github.com/OpenTreeOfLife/reference-taxonomy>
 #'  - `iucn`: IUCN Red List, https://iucnredlist.org
@@ -65,48 +65,10 @@ td_create <- function(provider = getOption("taxadb_default_provider", "itis"),
                       db = td_connect(dbdir)
                       ){
 
-  if(!dir.exists(dbdir))
-    dir.create(dbdir, FALSE, TRUE)
 
-  recognized_provider <- c(known_providers, "itis_test")
-  if (any(provider == "all")) {
-    provider <- known_providers
-  }
-  stopifnot(all(provider %in% recognized_provider))
 
-  ## supports vectorized schema and provider lists.
-  files <- unlist(lapply(schema, function(s)
-    paste0(s, "_", provider, ".tsv.bz2")))
-  #remove common name tables for providers without common names
-  files <- files[!files %in% paste0(NO_COMMON, ".tsv.bz2")]
-  dest <- file.path(dbdir, paste0(version, "_", files))
-
-  new_dest <- dest
-  new_files <- files
-
-  test_db <- grepl("itis_test", new_files)
-  if(any(test_db)){
-
-    file.copy(from = system.file(file.path("extdata", new_files), package = "taxadb"),
-              to = new_dest[test_db])
-    new_dest <- dest[!test_db]
-    new_files <- files[!test_db]
-  }
-
-  ## can this be removed? should be handled by arkdb
-  if (!overwrite) {
-    drop <- vapply(dest, file.exists, logical(1))
-    new_dest <- dest[!drop]
-    new_files <- files[!drop]
-  }
-
-  if (length(new_files) >= 1L) {
-    ## FIXME eventually these should be Zenodo URLs
[…]

### 2020-10-02 data-raw/col.R

@@ -1,131 +1,0 @@
-#' @importFrom utils globalVariables
-globalVariables(c("taxonID", "scientificName", "kingdom", "phylum", "class", "order", "family", "genus", "species",
-                  "taxonConceptID", "isExtinct", "specificEpithet", "infraspecificEpithet", "nameAccordingTo",
-                  "namePublishedIn", "scientificNameAuthorship", "namePublishedIn", "vernacularName", "language",
-                  "Species", "SpecCode", "Class", "SuperClass", "Family", "Order", "Genus", "Species", "type", "TaxonLevel",
-                  "synonym", "name", "Language", "commonNames", "accepted_id", "synonym_id", "SynCode", "taxon",
-                  "taxonomicStatus", "acceptedNameUsageID"))
-
-#' preprocess_col
-#'
-#' @param url Source of Catalogue of Life
-#' @param output_paths paths where output will be written: must be two paths, one named dwc, one named common
-#' @param dir working directory for downloads, will be tempdir() by default
-#'
-#' @export
-#' @import stringi forcats readr dplyr
-#' @importFrom methods className
-#' @importFrom stats family setNames
-#' @importFrom utils download.file untar unzip
-#' @details NOTE: A list of  all snapshots available from: http://www.catalogueoflife.org/DCA_Export/archive.php
-preprocess_col <- function(url = paste0("http://www.catalogueoflife.org/DCA_Export/zip-fixed/",
-                                        2019,
-                                        "-annual.zip"),
-                           output_paths = c(dwc = "2019/dwc_col.tsv.bz2",
-                                            common = "2019/common_col.tsv.bz2"),
-                           dir = file.path(tempdir(), "col")){
-
-
-  dir.create(dir, FALSE, FALSE)
-  curl::curl_download(url,
-                file.path(dir, "col-annual.zip"))
-  unzip(file.path(dir, "col-annual.zip"), exdir=dir)
-
-  ## a better read_tsv
-  read_tsv <- function(...) readr::read_tsv(..., quote = "",
-                                            col_types = readr::cols(.default = "c"))
-
-
-  taxon <- read_tsv(file.path(dir, "taxa.txt"))
[…]

### 2020-10-02 data-raw/gbif.R

@@ -1,106 +1,0 @@
-
-
-preprocess_gbif <- function(url = "http://rs.gbif.org/datasets/backbone/backbone-current.zip",
-                            output_paths = c(dwc = "2019/dwc_gbif.tsv.bz2",
-                                             common = "2019/common_gbif.tsv.bz2")
-                              ){
-
-  dir <- file.path(tempdir(), "gbif")
-  dir.create(dir, FALSE, FALSE)
-  archive <- file.path(dir, "backbone.zip")
-  download.file("http://rs.gbif.org/datasets/backbone/backbone-current.zip",
-                archive)
-
-
-  message(paste(file_hash(archive)))
-
-  unzip(archive, exdir=dir)
-
-  ## a better read_tsv
-  read_tsv <- function(...) readr::read_tsv(..., quote = "", col_types = readr::cols(.default = "c"))
-
-  ## And here we go!
-  taxon <- read_tsv(file.path(dir, "Taxon.tsv"))
-  ## canonicalName appears to be: Genus + specificEpithet + infraspecificEpithet
-  ## i.e. SpecificEpithet ~ a name at the "species" rank
-  ## Scientific name is ~ canonical name + citation parenthetical
-  ## Darwin Core defines ScientificName as: 	The full scientific name,
-  ## with authorship and date information if known. When forming part of an
-  ## Identification, this should be the name in lowest level taxonomic rank
-  ## that can be determined. This term should not contain identification
-  ## qualifications, which should instead be supplied in the IdentificationQualifier term.
-
-  ## This is unfortunate as author & date formatting introduces a huge
-  ## amount of variation that is difficult to resolve against, and best treated as a separate field.
-
-  ## genericName and canonicalName are not Darwin Core Taxon properties
-  ## Note that even within GBIF, formatting of scientificNameAuthorship
-  ## is highly non-standard (parens, abbrv, initials, etc)
-  gbif <- taxon %>%
-    select(taxonID,
-           scientificName = canonicalName,
-           taxonRank,
-           taxonomicStatus,
-           acceptedNameUsageID,
-           kingdom, phylum, class, order, family, genus, specificEpithet, infraspecificEpithet,
-           parentNameUsageID,
-           originalNameUsageID,
[…]

### 2020-10-02 data-raw/helper-routines.R

@@ -1,41 +1,0 @@
-#' @importFrom rlang quo := !!
-de_duplicate <- function(species){
-
-  ## Note for further testing: commented code below
-  ## adds a column indicating how many times species name appears,
-  ## & sorts with duplicates at top for inspection.
-  # duplicates <- species %>% count(name) %>%
-  #   inner_join(species, by ="name") %>% arrange(desc(n))
-
-  ## Note: based on old convention:
-  ## - name : scientificName
-  ## - name_type : taxonomicStatus
-
-  if("name_type" %in% colnames(species)){
-    ## A common reason for duplicates is that the same name matches
-    name_type <- quo(name_type)
-    name <- quo(name)
-    levels <- unique(species$name_type)
-    levels <- c(levels[!(levels %in% "accepted")], "accepted") # make accepted the "highest"
-    species <- species %>%
-      ## Uses explicit factor to enforce order -- no! drops all non-matched types
-      mutate(!!name_type := factor(!!name_type,  levels)) %>%
-      group_by(!!name) %>% top_n(1, !!name_type)
-
-  }
-  species
-}
-
-n_in_group <- function(data, group_var, ...){
-  data %>%
-    group_by_(group_var) %>%
-    top_n(...) %>%
-    distinct_(group_var, .keep_all = TRUE)
-}
-
-
-file_hash <- function(x, method = openssl::sha256, ...){
-  con <- lapply(x, file, raw=TRUE, ...)
-  hash <- lapply(con, method)
-  unlist(lapply(hash, as.character))
-}

### 2020-10-02 data-raw/metadata-generator.R

@@ -1,90 +1,0 @@
-library(openssl)
-library(fs)
-library(EML)
-library(tidyverse)
-library(uuid)
-library(here)
-library(jsonlite)
-
-file_hash <- function(x, method = openssl::sha256, ...){
-  con <- lapply(x, file, ...)
-  hash <- lapply(con, method)
-  unlist(lapply(hash, as.character))
-}
-
-release <- here::here("2019")
-meta <- fs::dir_info(release) %>%
-  mutate(sha256 = map_chr(path, file_hash, raw = TRUE),
-         name = fs::path_file(path),
-         contentType = "text/tab-separated-values",
-         contentEncoding = "bz2") %>%
-  select(name, size, sha256,
-         dateCreated = modification_time,
-         contentType, contentEncoding, path)
-
-
-## Lightweight file metadata
-#write_csv(meta, here::here("data-raw/meta.csv"))
-meta %>%
-  select(-path) %>%
-  mutate(size = as.character(size)) %>%
-  write_json(here::here("data-raw/meta.json"),
-             pretty = TRUE, auto_unbox=TRUE)
-
-
-dwc_terms <-
-  read_csv("https://github.com/tdwg/dwc/raw/master/vocabulary/term_versions.csv") %>%
-  select(attributeName = label, attributeDefinition = definition, definition = term_iri) %>%
-  distinct() %>%
-  bind_rows(tibble(  #
-    attributeName = "isExtinct",
-    definition = "logical indicating whether this taxon is now extinct",
-    attributeDefinition = definition))
-
-
-path <- meta$path[[1]]
-
-meta_dataset <- function(path){
-
-  sha256 <- file_hash(path, openssl::sha256, raw = TRUE)
-  info <- fs::file_info(path)
-
-  attributeName = path %>% read_tsv(n_max = 1) %>% colnames()
-  attrs <- tibble(attributeName = attributeName) %>% left_join(dwc_terms)
-
-  set_attributes(attrs, )
-  physical <- set_physical(path)
-
-
-  dataTable <- eml$dataTable(
-    entityName = basename(path),
-    entityDescription = "List of recognized taxonomic names in the Darwin Core format",
-    physical = physical,
-    attributeList = attrList)
-
-}
-
-
-me <- list(individualName = list(givenName = "Carl",
-                                 surName = "Boettiger"),
-           electronicMailAddress = "cboettig@berkeley.edu",
-           id = "http://orcid.org/0000-0002-1642-628X")
[…]

### 2020-10-02 data-raw/ncbi.R

@@ -1,213 +1,0 @@
-source("data-raw/helper-routines.R")
-preprocess_ncbi <- function(url,
-                            output_paths =
-                              c(dwc = "2020/dwc_ncbi.tsv.gz",
-                                common = "2020/common_ncbi.tsv.gz")){
-
-
-  dir <- file.path(tempdir(), "ncbi")
-  archive <- file.path(dir, "taxdmp.zip")
-  dir.create(dir, FALSE, FALSE)
-  download.file(url, archive)
-  message(paste(file_hash(archive)))
-
-
-  unzip(archive, exdir=dir)
-
-  node_cols <- c(
-  "tax_id",
-  "parent_tax_id",
-  "rank",
-  "embl_code",
-  "division_id",
-  "inherited_div_flag",
-  "genetic_code_id",
-  "inherited_GC_flag",
-  "mitochondrial_genetic_code_id",
-  "inherited_MGC_flag",
-  "GenBank_hidden_flag",
-  "hidden_subtree_root",
-  "comments"
-  )
-node_types <- c("iiclilililllc")
-
-  nodes <- vroom::vroom(file.path(dir, "nodes.dmp"), quote = "", delim = "\t|\t",
-                             col_names = node_cols, col_types = node_types)
-
-  name_cols <- c("tax_id", "name_txt", "unique_name", "name_class", "blank")
-  name_type <- c("icccc")
-  names <-  vroom::vroom(file.path(dir, "names.dmp"),
-                         delim = "\t|",
-                         col_names = name_cols,
-                         col_types = name_type, quote ="")
-  ncbi_taxa <-
-    inner_join(nodes,names) %>%
-    select(tax_id, parent_tax_id, rank, name_txt, unique_name, name_class)
-
-
-  ncbi_ids <- ncbi_taxa %>%
-    select(tsn = tax_id, parent_tsn = parent_tax_id) %>%
-    distinct() %>%
-    mutate(tsn = paste0("NCBI:", tsn),
-           parent_tsn = paste0("NCBI:", parent_tsn))
-
-  ## note: NCBI doesn't have rank ids
-  ncbi <- ncbi_taxa %>%
-    select(id = tax_id, name = name_txt, rank, name_type = name_class) %>%
-    mutate(id = paste0("NCBI:", id))
-
[…]

### 2020-10-02 data-raw/slb.R

@@ -1,122 +1,0 @@
-#' @importFrom tidyr drop_na gather
-
-###### sealifebase #######################
-#' @export
-preprocess_slb <- function(output_paths = c(dwc = "2019/dwc_slb.tsv.bz2",
-                                            common = "2019/common_slb.tsv.bz2")){
-
-  slb <- as_tibble(rfishbase::load_taxa(server = "sealifebase"))
-  slb_wide <- slb %>%
-    select( id = SpecCode,
-            genus = Genus,
-            species = Species,
-            subfamily = Subfamily,
-            family = Family,
-            order = Order,
-            class = Class,
-            phylum = Phylum,
-            kingdom = Kingdom) %>%
-    mutate(id = stri_paste("SLB:", id))
-
-  slb_accepted <- slb_wide %>% select(id, species) %>% gather(rank, name, -id)
-
-
-
-  species <- rfishbase:::fb_species(server = "sealifebase")
-  syn <- rfishbase::synonyms(NULL, server = "sealifebase")
-
-  slb_synonyms <- syn %>%
-    left_join(species) %>%
-    rename(id = SpecCode)  %>%
-    mutate(id = stri_paste("SLB:", id),
-           SynCode = stri_paste("SLB:", SynCode),
-           TaxonLevel = tolower(TaxonLevel)) %>%
-    select(id,
-           accepted_name = Species,
-           name = synonym,
-           type = Status,
-           synonym_id = SynCode,
-           rank = TaxonLevel)
-
-  slb_taxonid <- slb_synonyms %>%
-    rename(accepted_id = id, id = synonym_id) %>%
-    select(id, name, rank, accepted_id, name_type = type) %>%
-    bind_rows(mutate(slb_accepted, accepted_id = id, name_type = "accepted")) %>%
-    distinct() %>% de_duplicate()
-
-  ## Get common names
-  slb_common <- rfishbase::common_names(server = "sealifebase") %>%
-    drop_na(ComName)
-
-  #first english names
-  comm_eng <- slb_common %>%
-    filter(Language == "English") %>%
-    n_in_group(group_var = "SpecCode", n = 1, wt = ComName)
-
-  #get the rest
-  comm_names <- slb_common %>%
-    filter(!SpecCode %in% comm_eng$SpecCode) %>%
-    n_in_group(group_var = "SpecCode", n = 1, wt = ComName) %>%
-    bind_rows(comm_eng) %>%
-    mutate(taxonID = stri_paste("SLB:", SpecCode)) %>%
[…]

### 2020-10-02 data-raw/tpl.R

@@ -1,36 +1,0 @@
-#' @export
-preprocess_tpl <- function(output_files = c(dwc = "2019/dwc_tpl.tsv.bz2")){
-  archive <- file.path(tempdir(), "plantlist.tsv.bz2")
-  download.file("https://github.com/cboettig/taxadb/releases/download/data/taxizedb.2ftpl.2fplantlist.tsv.bz2",
-                archive)
-
-  read_tsv <- function(...) readr::read_tsv(..., quote = "", col_types = readr::cols(.default = "c"))
-  tpl_taxa <- read_tsv(archive)
-
-  ## note: only has accepted names
-  # tpl_taxa %>% count(taxonomic_status_in_tpl)
-  tpl_dwc <- tpl_taxa %>%
-    mutate(id = stri_paste("TPL:", id),
-           scientificName = stri_paste(genus, species),
-           taxonRank = "species",
-           taxonomicStatus = "accepted",
-           acceptedNameUsageID = id,
-           kingdom = "plantae", phylum = NA, class = NA, order = NA)  %>%
-    select(taxonID = id,
-           scientificName,
-           taxonRank,
-           acceptedNameUsageID,
-           kingdom, phylum, class, order, family, genus,
-           specificEpithet = species,
-           infraspecificEpithet = infraspecific_epithet,
-           scientificNameAuthorship = authorship,
-           namePublishedInYear = date,
-           nomenclaturalStatus = nomenclatural_status_from_original_data_source)
-
-  write_tsv(tpl_dwc, output_files["dwc"])
-  file_hash(output_files)
-}
-
-
-
-

### 2020-10-02 inst/examples/taxadb_github_releases.R

@@ -1,0 +1,46 @@
+
+providers_download_url <- function(files, version = latest_version()){
+  paste0("https://github.com/boettiger-lab/taxadb-cache/",
+         "releases/download/", version, "/", files)
+}
+
+
+taxadb_cache <- new.env()
+
+#' List available releases
+#'
+#' taxadb uses pre-computed cache files that are released on an annual
+#' version schedule.
+#'
+#' @export
+#' @examples
+#' available_versions()
+available_versions <- function(){
+
+  ## check for cached version first
+  avail_releases <- mget("avail_releases",
+                         envir = taxadb_cache,
+                         ifnotfound = NA)[[1]]
+  if(!all(is.na(avail_releases)))
+    return(avail_releases)
+
+  ## FIXME consider using direct access of a metadata record file instead
+  ## of relying on GitHub release tags to provide information about
+  ## available versions
+
+  ## Okay, check GH for a release
+  req <- curl::curl_fetch_memory(paste0(
+    "https://api.github.com/repos/",
+    "boettiger-lab/taxadb-cache/releases"))
+  json <- jsonlite::fromJSON(rawToChar(req$content))
+  avail_releases <- json[["tag_name"]]
+
+  ## Cache this so we don't hit GH every single time!
+  assign("avail_releases", avail_releases, envir = taxadb_cache)
+
+  avail_releases
+}
+
+latest_version <- function() {
+  available_versions()[[1]]
+}

### 2020-10-02 tests/testthat/test-taxadb.R

@@ -6,11 +6,8 @@ library(dplyr)
 
 
 test_that("we can set up a db and call basic functions", {
-
-  td_create(dbdir = test_db)
-  db <- td_connect(test_db)
-
-  df <- taxa_tbl(db = db)
+  db <- td_connect()
+  df <- taxa_tbl(db = db, provider = "itis_test")
 
   sp <- df %>%
     filter(family == "Cebidae") %>%
@@ -18,16 +15,18 @@ test_that("we can set up a db and call basic functions", {
 
   df <- filter_rank(name = "Cebidae",
                     rank = "family",
-                    db = db)  %>%
+                    db = db,
+                    provider = "itis_test")  %>%
     filter(taxonomicStatus == "accepted")
 
 
   species <- filter_name(df$scientificName,
-                 db = db) %>%
+                         db = db,
+                         provider = "itis_test") %>%
     filter(taxonomicStatus == "accepted")
 
   ## confirm order did not change
-  expect_identical(df$scientificName, species$scientificName)
+  expect_true(all(df$scientificName %in% species$scientificName))
 
 
   expect_is(df, "data.frame")

### 2021-02-25 README.Rmd
release prep

--- 2020-09-18 README.Rmd
@@
 [![R build status](https://github.com/ropensci/taxadb/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci/taxadb/actions)
-[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
+[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html)
 [![Coverage status](https://codecov.io/gh/ropensci/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/ropensci/taxadb?branch=master)
 [![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
+[![DOI](https://zenodo.org/badge/130153207.svg)](https://zenodo.org/badge/latestdoi/130153207)
 
@@
 taxadb:::td_disconnect()
-MonetDBLite::monetdblite_shutdown()
 

### 2021-02-25 R/td_create.R

@@ -20,7 +20,7 @@
 #' fast database connection.
 #' @details
 #'  Authorities currently recognized by taxadb are:
-#'  - `itis`: Integrated Taxonomic Information System, <https://www.itis.gov/>
+#'  - `itis`: Integrated Taxonomic Information System, https://www.itis.gov
 #'  - `ncbi`:  National Center for Biotechnology Information,
 #'  <https://www.ncbi.nlm.nih.gov/taxonomy>
 #'  - `col`: Catalogue of Life, <http://www.catalogueoflife.org/>

### 2021-03-04 paper/manuscript.Rmd

@@
 abstract: |
-  A familiar and growing challenge in ecological and evolutionary research is that of establishing consistent taxonomy when combining data from separate sources.
+  1)  A familiar and growing challenge in ecological and evolutionary research is that of establishing consistent taxonomy when combining data from separate sources.
 While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to retrieve taxonomic names.
-Here, we present `taxadb` R package to address this gap.
-In contrast to existing tools, `taxadb` provides all of the following in one place:
-  1) `taxadb` accesses established naming authorities to resolve synonyms, taxonomic identifiers, and hierarchical classification.  
-  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
-  3) `taxadb` provides a consistent and intuitive data format.  
-  4) `taxadb` is built on a simple, extensible, and language-agnostic design that can easily accommodate new authorities.
+  2) We present `taxadb` R package which creates a local database, managed automatically from within R, to provide fast operations on millions of taxonomic names. 
+  3) `taxadb` provides access to established naming authorities to resolve synonyms, taxonomic identifiers, and hierarchical classification in a consistent and intuitive data format.
+  4) `taxadb` makes operation on millions of taxonomic names fast and manageable.
 

### 2021-03-12 README.Rmd
get_ids bugfix

@@
   collapse = TRUE,
+  warning = FALSE,
   comment = "#>"
@@
 
-To get started, install the development version directly from GitHub:
+To get started, install from CRAN
 
-```{r eval=FALSE}
+``` r
+install.pacakges("taxadb")
+```
+
+or install the development version directly from GitHub:
+
+``` r
 devtools::install_github("ropensci/taxadb")
@@
 ```{r }
-td_create("col")
+td_create("col", overwrite=FALSE)
 ```
@@
   select(species) %>% 
-  mutate(id = get_ids(species, "col"))
+  mutate(id = taxalight::get_ids(species, "col"))
 

### 2021-03-12 R/get_ids.R

@@ -19,6 +19,8 @@
 #' @param taxadb_db Connection to from `[td_connect()]`.
 #' @param ignore_case should we ignore case (capitalization) in matching names?
 #' default is `TRUE`.
+#' @param warn should we display warnings on NAs resulting from multiply-resolved matches?
+#' (Unlike unmatched names, these NAs can usually be resolved manually via [filter_id()])
 #' @param ... additional arguments (currently ignored)
 #' @return a vector of IDs, of the same length as the input names Any
 #' unmatched names or multiply-matched names will return as [NA]s.
@@ -58,6 +60,7 @@ get_ids <- function(names,
                     version = latest_version(),
                     taxadb_db = td_connect(),
                     ignore_case = TRUE,
+                    warn = TRUE,
                     ...){
   format <- match.arg(format)
   n <- length(names)
@@ -67,7 +70,7 @@ get_ids <- function(names,
   # be compatible with common space delimiters
   names <- gsub("[_|-|\\.]", " ", names)
 
-  df <- filter_name(name = names,
+  taxa <- filter_name(name = names,
                 provider = provider,
                 version = version,
                 collect = TRUE,
@@ -76,7 +79,8 @@ get_ids <- function(names,
     arrange(sort)
 
   out <- vapply(names, function(x){
-    df <- df[df$scientificName == x, ]
+    df <- taxa[x == taxa$scientificName, ]
+    df <- df[!is.na(df$scientificName),]
 
     if(nrow(df) < 1) return(NA_character_)
 
@@ -91,12 +95,14 @@ get_ids <- function(names,
     if(length(ids)==1){
       return(ids)
     } else {
+      if(warn){
       warning(paste0("  Found ", bb(length(ids)), " possible identifiers for ",
                      ibr(x),
                      ".\n  Returning ", bb('NA'), ". Try ",
-                     bb(paste0("tl('", x, "', '", provider,"')")),
+                     bb(paste0("filter_id('", x, "', '", provider,"')")),
                      " to resolve manually.\n"),
               call. = FALSE)
+      }
       return(NA_character_)
     }
   },

### 2021-03-12 README.Rmd
update readme

@@
   collapse = TRUE,
-  warning = FALSE,
+  warning = TRUE,
   comment = "#>"
@@
   select(species) %>% 
-  mutate(id = taxalight::get_ids(species, "col"))
+  mutate(id = get_ids(species, "col"))
 

### 2021-03-12 R/get_ids.R

@@ -99,7 +99,7 @@ get_ids <- function(names,
       warning(paste0("  Found ", bb(length(ids)), " possible identifiers for ",
                      ibr(x),
                      ".\n  Returning ", bb('NA'), ". Try ",
-                     bb(paste0("filter_id('", x, "', '", provider,"')")),
+                     bb(paste0("filter_id('", x, "', '", db,"')")),
                      " to resolve manually.\n"),
               call. = FALSE)
       }

### 2021-12-15 README.Rmd

@@
 The goal of `taxadb` is to provide *fast*, *consistent* access to taxonomic data, supporting common tasks such as resolving taxonomic names to identifiers, looking up higher classification ranks of given species, or returning a list of all species below a given rank.
-These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records.
+These tasks are particularly common when synthesizing data across large species assemblies, such as combining occurrence records with trait records (see `taxadb` companion paper [Norman et al.
+2020](https://doi.org/10.1111/2041-210X.13440)).
 
@@
 
+## How to cite `taxadb`?
+
+You can access the citation information through `citation("taxadb")` within R, which refers to `taxadb` companion paper:
+
+> Kari E. A. Norman, Scott Chamberlain, and Carl Boettiger (2020). taxadb: A high-performance local taxonomic
+>  database interface. Methods in Ecology and Evolution, 11(9), 1153-1159.
+>  https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.13440.
+
 ## Learn more

### 2022-09-10 R/td_create.R
0.1.5 release

@@ -26,7 +26,7 @@
 #'  - `col`: Catalogue of Life, <http://www.catalogueoflife.org/>
 #'  - `tpl`: The Plant List, <http://www.theplantlist.org/>
 #'  - `gbif`: Global Biodiversity Information Facility, <https://www.gbif.org/>
-#'  - `fb`: FishBase, <https://www.fishbase.de/>
+#'  - `fb`: FishBase, https://www.fishbase.org
 #'  - `slb`: SeaLifeBase, <http://sealifebase.org>
 #'  - `wd`: Wikidata: https://www.wikidata.org
 #'  - `ott`: OpenTree Taxonomy:

### 2023-02-13 README.Rmd
Update README.Rmd
modern check

--- 2021-03-12 README.Rmd
@@
 <!-- badges: start -->
-[![R build status](https://github.com/ropensci/taxadb/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci/taxadb/actions)
+[![R-CMD-check](https://github.com/ropensci/taxadb/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ropensci/taxadb/actions/workflows/R-CMD-check.yaml)
 [![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html)
@@
 [![DOI](https://zenodo.org/badge/130153207.svg)](https://zenodo.org/badge/latestdoi/130153207)
-
-<!-- [![peer-review](https://badges.ropensci.org/344_status.svg)](https://github.com/ropensci/software-review/issues/344) -->
 <!-- badges: end -->
@@
 ``` r
-install.pacakges("taxadb")
+install.packages("taxadb")
 ```
@@
 
-Create a local copy of the Catalogue of Life (2018) database:
+Create a local copy of the (current) Catalogue of Life database:
 
 ```{r }
-td_create("col", overwrite=FALSE)
+td_create("col")
 ```
@@
 
-For instance, *Trochalopteron henrici gucenense* does not resolve to an identifier in ITIS:
+For instance, *Agrostis caespitosa* does not resolve to an identifier in ITIS:
 
 ```{r}
-get_ids("Trochalopteron henrici gucenense") 
+get_ids("Agrostis caespitosa", "itis") 
 ```
 
-Using `filter_name()`, we find this is because the name resolves not to zero matches, but to more than one match:
+Using `filter_name()`, we find this is because the name resolves not to zero matches, but is a known synonym to more than one accepted name (as indicated by the accepted name usage id)
 
 ```{r}
-filter_name("Trochalopteron henrici gucenense") 
+filter_name('Agrostis caespitosa', 'itis')
 ```
 
+We can resolve the scientific name to the acceptedNameUsage using `get_names()` on the _accepted_ IDs: (These also correspond to the genus and specificEpithet column, as the classification is always given only based on acceptedNameUsageID).
+
 ```{r}
-filter_name("Trochalopteron henrici gucenense")  %>%
+filter_name("Agrostis caespitosa")  %>%
   mutate(acceptedNameUsage = get_names(acceptedNameUsageID)) %>% 
@@
 ```{r}
-taxa_tbl("col") %>% 
+taxa_tbl("itis") %>% 
   count(acceptedNameUsageID, sort=TRUE)
@@
 
-if(require(codemetar)) codemetar::write_codemeta()
+if(require(codemeta)) codemeta::write_codemeta()
 ```

### 2023-02-13 R/td_connect.R

@@ -36,7 +36,7 @@ td_connect <- function(dbdir = NULL,
   db <- mget(db_name, envir = taxadb_cache, ifnotfound = NA)[[1]]
 
   if(!inherits(db, "duckdb_connection")){
-    db <- DBI::dbConnect(drv = duckdb::duckdb())
+    db <- DBI::dbConnect(duckdb::duckdb())
     assign(db_name, db, envir = taxadb_cache)
   }
   db

### 2023-03-08 README.Rmd
drop codecov badge due to possible html 500 error
CRAN hates when a link doesn't respond

@@
 [![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html)
-[![Coverage status](https://codecov.io/gh/ropensci/taxadb/branch/master/graph/badge.svg)](https://codecov.io/github/ropensci/taxadb?branch=master)
 [![CRAN status](https://www.r-pkg.org/badges/version/taxadb)](https://cran.r-project.org/package=taxadb)
